### 2020-04-10 doc/manuscript/ms.Rmd
Started to write ms

---
title: "Testing the thermal buffer of mire groundwater"
author: "Eduardo Fernández-Pascual, University of Oviedo (fernandezpeduardo@uniovi.es)"
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output:
  bookdown::html_document2
bibliography: refs.bib
biblio-style: apalike
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(here); library(kableExtra)
load(here("results", "logs.RData"))
source(here("src", "analysis.R"))
nsites <- length(unique(temperatures$Logger))/2
```

# Materials and methods

This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W). Local geography and climate are dominated by the Cantabrian Mountains (> 1500 m above sea level), which run parallel to the coast and trap the prevailing NW Atlantic winds. The resulting humid climate harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246]. Rain-fed raised bogs and acid valley mires can be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes. Glacial lakes undergoing silting develop transition mires and quaking bogs communities in the water-to-land transition. Spring fens appear in the mountains above 1000 m; they range from soft-water poor fens on acid bedrocks, to alkaline calcareous fens on limestone.

A selection of `r nsites` mire sites was made, representing the regional elevation gradient of mire vegetation and the different mire types (Table \@ref(tab:table1))). In each site, two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany) were buried at a depth of 5 cm below the upper layer of the soil. They were programmed to record temperature on an hourly basis. One datalogger was placed in a flat waterlogged spot, within the mire. The other dataloggers was placed in the close vicinity, but in a flat and dry area outside the mire. The vegetation was always either mire or pasture, with no shrubs, trees or any other landscape features shading the measuring points. The dataloggers were left on site for approximately five years, after which they were retrieved and their records downloaded. At the moment of retrieval, the internal clock of all the dataloggers had not deviated for more than four hours.

Some steps were taken to clean the logs. First, the records from the first week after installation were removed, to account for the own installation process and the settling of the soils. Second, and because some of the dataloggers had failed at different points in time, a selection was made for each site to keep only time series with records both for the dry and waterlogged points. Afterwards four bioclimatic variables were calculated for each datalogger:

* The mean diurnal range; i.e. the average for the whole period of the daily differences between the maximum and the minimum temperatures recorded in the day.
* The maximum temperature of the warmest month; i.e. the average of the daily maximum records, for the warmst month.
* The minimum temperature of the codlest month; i.e. the average of the daily minimum records, for the coldest month.
* The annual range; i.e. the difference between the maximum temperature of the warmest month and the minimum temperature of the coldest month.

To test if the differences between the dry and the waterlogged points of each site was significant, paired t-tests were used. Tests were one-tailed, according to the following hypotheses: the dry site would have a higher diurnal range, a higher maximum temperature, a lower minimum temperature, and a higher annual range.

# Results

The dataloggers recorded temperatures for five years in five of the sites, four years in two, and two years in one.

```{r }
print(ttests)
```

# Tables

```{r table1, echo = FALSE}
table1 %>% kable(caption = "Mire sites included in this study, indicating the type of fen, the elevation, coordinates, and length of the temperature recording period.", booktabs = TRUE) %>% kable_styling(latex_options = c("striped", "hold_position"))
```

```{r, echo = FALSE}
table2 %>% kable(caption = "Bioclimatic variables calculated for the mire sites, using the hourly temperature records.", booktabs = TRUE) %>% kable_styling(latex_options = c("striped", "hold_position"))
```

# Figures

```{r echo = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites. The blue series was recorded within the mire, in a waterloged area. The red series was recorded in a neighbouring dry area. Dataloggers were buried at 5 cm depth.", fig.align = "center", fig.width=3, fig.height=8}
plot1
```

# References

### 2020-04-10 src/analysis.R

@@ -1,0 +1,150 @@
+library(tidyverse); library(ggthemes); library(lubridate)
+
+# Calculations
+
+## Bioclimatic variables US Geological Survey, WorlClim
+
+### Bio 2 Annual Mean Diurnal Range
+
+temperatures %>%
+  group_by(Logger, Day = floor_date(Time, "day")) %>%
+  summarise(Tmax = max(Temperature), 
+            Tmin = min(Temperature),
+            Tdif = Tmax - Tmin) %>%
+  group_by(Logger) %>%
+  summarise(Mean.Diurnal.Range = mean(Tdif)) -> bio2
+
+### Bio 5 Max Temperature of Warmest Month
+
+temperatures %>%
+  group_by(Logger, Day = floor_date(Time, "day")) %>%
+  summarise(Tmax = max(Temperature)) %>%
+  mutate(Month = month(Day)) %>%
+  group_by(Logger, Month) %>%
+  summarise(Max.Temperature.of.Warmest.Month = mean(Tmax)) %>%
+  group_by(Logger) %>%
+  filter(Max.Temperature.of.Warmest.Month == max(Max.Temperature.of.Warmest.Month)) %>%
+  select(-Month) -> bio5
+
+### Bio 6 Min Temperature of the Coldest Month
+
+temperatures %>%
+  group_by(Logger, Day = floor_date(Time, "day")) %>%
+  summarise(Tmin = min(Temperature)) %>%
+  mutate(Month = month(Day)) %>%
+  group_by(Logger, Month) %>%
+  summarise(Min.Temperature.of.Coldest.Month = mean(Tmin)) %>%
+  group_by(Logger) %>%
+  filter(Min.Temperature.of.Coldest.Month == min(Min.Temperature.of.Coldest.Month)) %>%
+  select(-Month) -> bio6
+
+### Merge and calculate Bio 7 Annual Temperature Range
+
+bio2 %>%
+  merge(bio5) %>%
+  merge(bio6) %>%
+  mutate(Temperature.Annual.Range = 
+           Max.Temperature.of.Warmest.Month -
+           Min.Temperature.of.Coldest.Month) -> bioclimatics
+
+## Paired T-tests
+
+t.testEFP <- function(df) 
+{
+  t <- t.test(df$Waterlogged, df$Dry, alternative = "less", paired = TRUE)
+  data.frame(t$statistic,
+             t$p.value,
+             t$estimate)
+}
+
+t.testEFPgreater <- function(df) 
+{
[…]

### 2020-04-10 src/cleaning.R

@@ -1,0 +1,154 @@
+library(tidyverse); library(sp); library(lubridate)
+
+# Function reads and formats GP5WSHELL csv export files
+
+read_gp5 <- function(x)
+{
+  require(tidyverse)
+  read.csv(x, skip = 2, header = TRUE) %>% # Rows to skip depend on GP5WShell version!!
+    dplyr::rename(Temperature = X.1.oC) %>%
+    mutate(Time = as.POSIXct(Time, 
+                             format =" %d.%m.%Y %H:%M:%S", 
+                             tz = "UTC"),
+           Logger = gsub("\\_.*", "" , x)) %>%
+    dplyr::select(Logger, Time, Temperature)
+}
+
+# Functions convert coordinates to decimal degrees
+
+decdeg30T <- function(x) { x -> o1 
+  o1 %>% dplyr::select(ED50X, ED50Y) -> o2
+  SpatialPoints(o2, proj4string = CRS("+proj=utm +zone=30T +datum=WGS84")) -> o3
+  spTransform(o3, CRS("+proj=longlat +datum=WGS84")) %>% data.frame -> o4
+  colnames(o4) <- c("Longitude", "Latitude")
+  o1 %>% dplyr::select(-c(ED50X, ED50Y)) -> o5
+  cbind(o5, o4)
+}
+
+decdeg29T <- function(x) { x -> o1 
+  o1 %>% dplyr::select(ED50X, ED50Y) -> o2
+  SpatialPoints(o2, proj4string = CRS("+proj=utm +zone=29T +datum=WGS84")) -> o3
+  spTransform(o3, CRS("+proj=longlat +datum=WGS84")) %>% data.frame -> o4
+  colnames(o4) <- c("Longitude", "Latitude")
+  o1 %>% dplyr::select(-c(ED50X, ED50Y)) -> o5
+  cbind(o5, o4)
+}
+
+# Read logger header data
+
+read.table("data/Loggers.txt", sep = "\t", header = TRUE) %>%
+  mutate(Installed = as.POSIXct(Installed, 
+                                format ="%d/%m/%Y", 
+                                tz = "UTC")) -> df1
+
+df1 %>% 
+  filter(ED50Zone == "29T") %>%
+  decdeg29T -> df29T
+
+df1 %>% 
+  filter(ED50Zone == "30T") %>%
+  decdeg30T -> df30T
+
+rbind(df29T, df30T) %>%
[…]

### 2020-04-10 src/exploratory analysis.R

@@ -1,254 +1,0 @@
-library(tidyverse); library(lubridate)
-load("results/clean_data/temperatures.RData")
-
-# Days of recordings
-
-temperatures %>%
-  group_by(Logger) %>%
-  summarise(Start = min(Time),
-            End = max(Time),
-            Length = End - Start) -> recording.period
-
-# Bioclimatic variables US Geological Survey, WorlClim
-
-# Bio1 Annual Mean Temperature
-
-temperatures %>%
-  group_by(Logger) %>%
-  summarise(Annual.Mean.Temperature = mean(Temperature)) -> bio1
-
-# Bio 2 Annual Mean Diurnal Range
-
-temperatures %>%
-  group_by(Logger, Day = floor_date(Time, "day")) %>%
-  summarise(Tmax = max(Temperature), 
-            Tmin = min(Temperature),
-            Tdif = Tmax - Tmin) %>%
-  group_by(Logger) %>%
-  summarise(Mean.Diurnal.Range = mean(Tdif)) -> bio2
-
-# Bio 4 Temperature Seasonality (Standard Deviation)
-
-temperatures %>%
-  group_by(Logger) %>%
-  summarise(Temperature.Seasonality = sd(Temperature)) -> bio4
-
-# Bio 5 Max Temperature of Warmest Month
-
-temperatures %>%
-  group_by(Logger, Day = floor_date(Time, "day")) %>%
-  summarise(Tmax = max(Temperature)) %>%
-  mutate(Month = month(Day)) %>%
-  group_by(Logger, Month) %>%
-  summarise(Max.Temperature.of.Warmest.Month = mean(Tmax)) %>%
-  group_by(Logger) %>%
-  filter(Max.Temperature.of.Warmest.Month == max(Max.Temperature.of.Warmest.Month)) %>%
-  select(-Month) -> bio5
-
-# Bio 6 Min Temperature of the Coldest Month
-
-temperatures %>%
-  group_by(Logger, Day = floor_date(Time, "day")) %>%
-  summarise(Tmin = min(Temperature)) %>%
-  mutate(Month = month(Day)) %>%
-  group_by(Logger, Month) %>%
-  summarise(Min.Temperature.of.Coldest.Month = mean(Tmin)) %>%
-  group_by(Logger) %>%
-  filter(Min.Temperature.of.Coldest.Month == min(Min.Temperature.of.Coldest.Month)) %>%
[…]

### 2020-04-10 src/final analysis.R

@@ -1,152 +1,0 @@
-library(tidyverse); library(lubridate)
-load("results/clean_data/temperatures.RData")
-
-# Calculations
-
-## Bioclimatic variables US Geological Survey, WorlClim
-
-### Bio 2 Annual Mean Diurnal Range
-
-temperatures %>%
-  group_by(Logger, Day = floor_date(Time, "day")) %>%
-  summarise(Tmax = max(Temperature), 
-            Tmin = min(Temperature),
-            Tdif = Tmax - Tmin) %>%
-  group_by(Logger) %>%
-  summarise(Mean.Diurnal.Range = mean(Tdif)) -> bio2
-
-### Bio 5 Max Temperature of Warmest Month
-
-temperatures %>%
-  group_by(Logger, Day = floor_date(Time, "day")) %>%
-  summarise(Tmax = max(Temperature)) %>%
-  mutate(Month = month(Day)) %>%
-  group_by(Logger, Month) %>%
-  summarise(Max.Temperature.of.Warmest.Month = mean(Tmax)) %>%
-  group_by(Logger) %>%
-  filter(Max.Temperature.of.Warmest.Month == max(Max.Temperature.of.Warmest.Month)) %>%
-  select(-Month) -> bio5
-
-### Bio 6 Min Temperature of the Coldest Month
-
-temperatures %>%
-  group_by(Logger, Day = floor_date(Time, "day")) %>%
-  summarise(Tmin = min(Temperature)) %>%
-  mutate(Month = month(Day)) %>%
-  group_by(Logger, Month) %>%
-  summarise(Min.Temperature.of.Coldest.Month = mean(Tmin)) %>%
-  group_by(Logger) %>%
-  filter(Min.Temperature.of.Coldest.Month == min(Min.Temperature.of.Coldest.Month)) %>%
-  select(-Month) -> bio6
-
-### Merge and calculate Bio 7 Annual Temperature Range
-
-bio2 %>%
-  merge(bio5) %>%
-  merge(bio6) %>%
-  mutate(Temperature.Annual.Range = 
-           Max.Temperature.of.Warmest.Month -
-           Min.Temperature.of.Coldest.Month) -> bioclimatics
-
-## Paired T-tests
-
-t.testEFP <- function(df) 
-{
-  t <- t.test(df$Waterlogged, df$Dry, alternative = "less", paired = TRUE)
-  data.frame(t$statistic,
-             t$p.value,
-             t$estimate)
-}
-
-t.testEFPgreater <- function(df) 
-{
[…]

### 2020-04-10 src/read loggers.R

@@ -1,150 +1,0 @@
-library(tidyverse); library(sp); library(lubridate)
-
-# Function reads and formats GP5WSHELL csv export files
-
-read_gp5 <- function(x)
-{
-  require(tidyverse)
-  read.csv(x, skip = 2, header = TRUE) %>% # Rows to skip depend on GP5WShell version!!
-    dplyr::rename(Temperature = X.1.oC) %>%
-    mutate(Time = as.POSIXct(Time, 
-                             format =" %d.%m.%Y %H:%M:%S", 
-                             tz = "UTC"),
-           Logger = gsub("\\_.*", "" , x)) %>%
-    dplyr::select(Logger, Time, Temperature)
-}
-
-# Functions convert coordinates to decimal degrees
-
-decdeg30T <- function(x) { x -> o1 
-  o1 %>% dplyr::select(ED50X, ED50Y) -> o2
-  SpatialPoints(o2, proj4string = CRS("+proj=utm +zone=30T +datum=WGS84")) -> o3
-  spTransform(o3, CRS("+proj=longlat +datum=WGS84")) %>% data.frame -> o4
-  colnames(o4) <- c("Longitude", "Latitude")
-  o1 %>% dplyr::select(-c(ED50X, ED50Y)) -> o5
-  cbind(o5, o4)
-}
-
-decdeg29T <- function(x) { x -> o1 
-  o1 %>% dplyr::select(ED50X, ED50Y) -> o2
-  SpatialPoints(o2, proj4string = CRS("+proj=utm +zone=29T +datum=WGS84")) -> o3
-  spTransform(o3, CRS("+proj=longlat +datum=WGS84")) %>% data.frame -> o4
-  colnames(o4) <- c("Longitude", "Latitude")
-  o1 %>% dplyr::select(-c(ED50X, ED50Y)) -> o5
-  cbind(o5, o4)
-}
-
-# Read logger header data
-
-read.table("data/dataloggers/Loggers.txt", sep = "\t", header = TRUE) %>%
-  mutate(Installed = as.POSIXct(Installed, 
-                                format ="%d/%m/%Y", 
-                                tz = "UTC")) -> df1
-
-df1 %>% 
-  filter(ED50Zone == "29T") %>%
-  decdeg29T -> df29T
-
-df1 %>% 
-  filter(ED50Zone == "30T") %>%
-  decdeg30T -> df30T
-
-rbind(df29T, df30T) %>%
[…]

### 2020-04-26 doc/manuscript/ms.Rmd
changes to ms

@@
 source(here("src", "analysis.R"))
-nsites <- length(unique(temperatures$Logger))/2
 ```
+
+# Introduction
 
@@
 
-A selection of `r nsites` mire sites was made, representing the regional elevation gradient of mire vegetation and the different mire types (Table \@ref(tab:table1))).
+A selection of `r length(unique(temperatures$Logger))/2` mire sites was made, representing the regional elevation gradient of mire vegetation and the different mire types (Table \@ref(tab:table1))).
 In each site, two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany) were buried at a depth of 5 cm below the upper layer of the soil.
@@
 
-The dataloggers recorded temperatures for five years in five of the sites, four years in two, and two years in one.
+The dataloggers recorded temperatures for five years in five of the sites, four years in two, and two years in one (Fig.
+\@ref(fig:fig1)).
+A visual inspection of the time series shows more homogeneous temperatures in the waterlogged time series, which are almost always encompassed within the dry time series.
+The bioclimatic variables (Table \@ref(tab:table2)) support this impression.
+The maximum temperatures of the warmest months are usually higher in the dry measuring points, by more than 6 ºC in some sites, both at low (El Molinucu, La Malva) and high (La Recoleta) elevations.
+The opposite is true for the minimums of the coldest months, in which case the temperature is generally colder in the dry points, although the difference is less pronounced than for the maximums.
+The mean diurnal range of temperatures is wider in the dry sites, as is annual temperature range; and again the effect is larger in the low sites and in La Recoleta.
+The t-tests supported the original hypotheses; namely waterlogged measuring points had (a) smaller diurnal fluctuactions (t = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.estimate) %>% round(2)` ºC); (b) lower maximums (t = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.estimate) %>% round(2)` ºC); (c) higher minimums (t = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.estimate) %>% round(2)` ºC), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.estimate) %>% round(2)` ºC).
 
-```{r }
-print(ttests)
-```
+# Discussion
 
@@
 
-```{r, echo = FALSE}
+```{r table2, echo = FALSE}
 table2 %>% kable(caption = "Bioclimatic variables calculated for the mire sites, using the hourly temperature records.", booktabs = TRUE) %>% kable_styling(latex_options = c("striped", "hold_position"))
@@
 
-```{r echo = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites.
+```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites.
 The blue series was recorded within the mire, in a waterloged area.

### 2020-05-07 doc/manuscript/ms.Rmd
intro drafted

@@
 # Introduction
+
+Climate change is already affecting global biodiversity patterns (IPCC 2014).
+Among other trends, there has been an expansion of drylands (Huang, J., Yu, H., Guan, X.
+et al.
+Accelerated dryland expansion under climate change), an increase of disturbations in forests (Seidl, R., Thom, D., Kautz, M.
+et al.
+Forest disturbances under climate change.
+Nature Clim Change 7, 395–402 (2017).
+https://doi.org/10.1038/nclimate3303) and ocean acidification (Coral Reef Ecosystems under Climate Change and Ocean Acidification).
+As (M.
+Horsák et al.
+/ Science of the Total Environment 634 (2018) 355–365) point out, some characteristics of ecosystems that make them especially vulnerable to climate change are (i) a preponderance of species that evolved under a cold climate, (ii) low productivity caused by nutrient limitation, which makes them sensitive to increased nutrient
+cycling caused by warming (Cornelissen et al., 2007), and (iii) a scatterred distribution that limits species dispersal and migration (Pearson and Dawson, 2005).
+These characteristics are found in temperate peatlands such as mires, permanent semi-terrestrial wetlands where soils remain waterlogged but not inundated during most of the year (Wheeler and Proctor, 2000).
+Mires are priority habitats for conservation that harbour a high number of endangered species (Bergamini et al.
+2009) and support highly-adapted flora in spatially reduced areas (Grootjans et al.
+2006, Jiménez-Alfaro et al.
+2013).
+Habitat distribution models have predicted a loss of mire surface in face of climate change (Essl et al.
+2012).
+
+Recent research has focused on the importance of groundwater as a regulator of mire microclimate. As azonal habitats whose occurence is driven by local edaphic conditions, locally obtained
+thermal data are hence crucial for any prediction about the impact of climatically-induced warming (Breckle 2002).
+In Europe, fens are distributed from the Iberian Peninsula to Fennoscandia and from low valleys to the alpine belt, but nonetheless the constituent and dominant species remain relatively homogeneous (Jiménez-Alfaro et al.
+2014).
+An apparent climatic resilience of fen species is further suggested by the persistence of relicts in their distribution limits (Amon et al.
+2002, Topić and Stančić 2006, Hájek et al.
+2009, Jiménez-Alfaro et al.
+2012).
+A possible cause for this climatic resilience may be the buffer effect of groundwater on soil temperature, reducing the expected differences of the realized niche in diverse climatic regions (Ellenberg 1988).
+Indeed, the cool waters from underground aquifers that saturate fens are expected to keep temperatures warmer during cold periods and vice versa (Ellenberg 1988, Geiger et al.
+2009).
+As root-zone temperature is known to be a major determinant of plant ecophysiology (Körner and Paulsen 2004), this buffer effect might allow fen species to regenerate and live along a wider range of air temperatures than they could otherwise.
+Soil temperature measurements have become available for mires of North America (Raney 2014), Western Europe (Fernández-Pascual et al.
+2015), Central Europe (Horsak 2018).
+The comparison of said temperatures with air temperatures derived from models has shown that mire soil temperature is warmer in winter and cooler in summer, thus indicating the existence of a thermal buffer by groundwater (FerPas, Horsak).
+This effect has been linked to the composition of flora and fauna (Horsak 2018, Shecnkova 2020), growth of mire trees /Raney 2016) and the role of mires as glacial refugia (JimAlf 2016, Dite 2017).
+It is evident that the thermal buffer of groundwater will play a profound role in the response of these habitats to climate change, and needs to be studied in more depth.
+
+To my knowledge, the soil thermal buffer of mires has not yet been compared to adjacent dry soild with temperatures measured in situ.
+This article provides the first record of the thermal buffer against surrounding non-mire areas, based on soil temperatures recorded during a period of five years.
+Specifically, this article test the hypothesis that temperatures in the mire soil will be warmer in winter, colder in summer and less variable than temperatures in adjacent dry soils.
 

### 2020-05-07 doc/manuscript/ms.Rmd
intro drafted 2

@@
 2009) and support highly-adapted flora in spatially reduced areas (Grootjans et al.
-2006, Jiménez-Alfaro et al.
-2013).
+2006).
 Habitat distribution models have predicted a loss of mire surface in face of climate change (Essl et al.
@@
 thermal data are hence crucial for any prediction about the impact of climatically-induced warming (Breckle 2002).
-In Europe, fens are distributed from the Iberian Peninsula to Fennoscandia and from low valleys to the alpine belt, but nonetheless the constituent and dominant species remain relatively homogeneous (Jiménez-Alfaro et al.
-2014).
-An apparent climatic resilience of fen species is further suggested by the persistence of relicts in their distribution limits (Amon et al.
-2002, Topić and Stančić 2006, Hájek et al.
-2009, Jiménez-Alfaro et al.
-2012).
+In Europe, fens are distributed from the Iberian Peninsula to Fennoscandia and from low valleys to the alpine belt, but nonetheless the constituent and dominant species remain relatively homogeneous (Peterka).
 A possible cause for this climatic resilience may be the buffer effect of groundwater on soil temperature, reducing the expected differences of the realized niche in diverse climatic regions (Ellenberg 1988).

### 2020-05-09 doc/manuscript/ms.Rmd
polish intro

--- 2020-04-26 doc/manuscript/ms.Rmd
@@
 # Introduction
+
+Climate change [@RN3586] affects global biodiversity from drylands [@RN4672] to forests [@RN4673] and oceans [@RN4674].
+Ecosystems can be especially vulnerable when they meet certain criteria [@RN4675]: (a) preponderance of species that evolved under a cold climate; (b) low productivity due to nutrient limitation, making them sensitive to increased nutrient cycling caused by warming [@RN4676]; and (c) scatterred distribution pattern, which limits species dispersal and migration [@RN4677].
+These characteristics are found in temperate peatlands such as mires, i.e.
+permanent semi-terrestrial wetlands where soils remain waterlogged but not inundated during most of the year [@RN3161].
+Mires are priority habitats for conservation that harbour high numbers of endangered species [@RN3122] and support highly-adapted floras in spatially-reduced areas [@RN2960].
+Habitat distribution models have predicted a loss of mire surface as a consequence of ongoing climate change [@RN2937].
+
+Mires are however azonal habitats whose existence depends on local soil properties [@RN3328]: by definition, they are areas with waterlogged soils.
+Grounwater can produce a buffer effect on soil temperature, keeping soils warmer than the air during cold periods, and vice versa [@RN3344; @RN3201].
+Root-zone temperature is a major determinant of plant ecophysiology [@RN3024], so the groundwater buffer effect is expected to allow mire plants to live along a wider range of air temperatures than they could otherwise.
+Indeed, mires have a relatively homogeneous flora despite being distributed from Mediterranean to Boreal biomes, and from low valleys to the alpine belt [@RN4678].
+Recently, soil temperature measurements have become available for mires of North America [@RN3204], Western Europe [@RN2356] and Central Europe [@RN4675].
+Their comparison with air temperatures derived from models has shown that mire soils are indeed warmer in winter and cooler in summer, thus giving support to the existence of the groundwater thermal buffer [@RN2356; @RN4675].
+Furthermore, the effect has been linked to the composition of mire flora and fauna [@RN4675; @RN4679], the growth rings of mire trees [@RN3060] and the role of mires as glacial refugia [@RN2513; @RN4680].
+
+It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change.
+As is the case for all azonal habitats, locally measured temperatures are essential to understand this response.
+Available data recorded temperatures only on the waterlogged soils, using model-derived air temperatures for comparison [@RN2356; @RN4675; @RN4679]; or recorded temperatures in both wet and dry spots but for less than a year [@RN3204; @RN3060].
+This article provides the first measurement of the thermal buffer against surrounding non-mire areas, based on soil temperatures recorded during a period of five years.
+These measures are used to test the hypotheses that, when compared to adjacent dry soils, waterlogged mire soil are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales.
 
@@
 The mean diurnal range of temperatures is wider in the dry sites, as is annual temperature range; and again the effect is larger in the low sites and in La Recoleta.
-The t-tests supported the original hypotheses; namely waterlogged measuring points had (a) smaller diurnal fluctuactions (t = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.estimate) %>% round(2)` ºC); (b) lower maximums (t = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.estimate) %>% round(2)` ºC); (c) higher minimums (t = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.estimate) %>% round(2)` ºC), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.estimate) %>% round(2)` ºC).
+The t-tests supported the original hypotheses (Fig.
+\@ref(fig:fig2)); namely waterlogged measuring points had (a) smaller diurnal fluctuactions (t = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.estimate) %>% round(2)` ºC); (b) lower maximums (t = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.estimate) %>% round(2)` ºC); (c) higher minimums (t = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.estimate) %>% round(2)` ºC), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.estimate) %>% round(2)` ºC).
 
 # Discussion
+
+Less effect in winter could be because this is soil (snow, etc)
 
@@
 The red series was recorded in a neighbouring dry area.
-Dataloggers were buried at 5 cm depth.", fig.align = "center", fig.width=3, fig.height=8}
+Dataloggers were buried at 5 cm depth.", fig.align = "center", fig.width=7, fig.height=6}
 plot1
@@
 
+```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "Average bioclimatic variables in the dry and waterlogged sites.
+The bars represent the mean value, and the bars the standard error of 8 records", fig.align = "center", fig.width=7, fig.height=3}
+plot2
+```
+
 # References

### 2020-05-09 src/analysis.R

@@ -1,4 +1,6 @@
-library(tidyverse); library(ggthemes); library(lubridate)
+library(tidyverse); library(ggthemes); library(lubridate); library(here)
+
+#Iload(here("results", "logs.RData"))
 
 # Calculations
 
@@ -139,7 +141,7 @@ recordsdf %>%
   ggplot(aes(x = Time, y = Temperature, color = Groundwater)) +
   geom_hline(yintercept = 0, linetype = "dashed") +
   geom_line(alpha = 0.85) +
-  facet_wrap(~ Facet, ncol = 1) +
+  facet_wrap(~ Facet, ncol = 4) +
   xlab("Time (hourly records)") + ylab("Temperature (ºC)") +
   #scale_color_manual(values = c("darkorchid", "gold")) +
   theme_tufte() +
@@ -147,4 +149,28 @@ recordsdf %>%
         legend.position = "top", 
         legend.justification = "center",
         legend.title = element_blank()) +
-  guides(colour = guide_legend(override.aes = list(alpha=1))) -> plot1
+  guides(colour = guide_legend(override.aes = list(alpha = 1))) -> plot1
+
+# Fig 2
+
+merge(header, bioclimatics) %>%
+  select(Site, Groundwater, Mean.Diurnal.Range:Temperature.Annual.Range) %>%
+  gather(Trait, Value, Mean.Diurnal.Range:Temperature.Annual.Range) %>%
+  mutate(Trait = gsub("\\.", " ", Trait), 
+         Value = round(Value, 2)) %>%
+  group_by(Trait, Groundwater) %>%
+  summarise(y = mean(Value), SD = sd(Value), n = length(Value),
+            SE = SD/sqrt(n)) %>%
+  ggplot(aes(x = Groundwater, y = y, ymin = y - SE, ymax = y + SE,
+             fill = Groundwater)) +
+  geom_bar(position = "dodge", stat = "identity", alpha = 0.7) +
+  facet_wrap(~ Trait, scales = "free", strip.position = "left", ncol = 4) +
+  geom_errorbar(width = .2, position = position_dodge(.9),
+                aes(color = Groundwater), show.legend = FALSE) +
+  theme_tufte() +
+  theme(text = element_text(size = 10),
+        legend.position = "top", 
+        legend.justification = "center",
+        legend.title = element_blank(),
+        strip.placement = "outside",
+        axis.title = element_blank()) -> plot2

### 2020-05-12 doc/manuscript/ms.Rmd
word format

@@
 ---
-title: "Testing the thermal buffer of mire groundwater"
-author: "Eduardo Fernández-Pascual, University of Oviedo (fernandezpeduardo@uniovi.es)"
+title: "Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils"
+subtitle: "Short title: Testing the thermal buffer of mire groundwater"
+author: "Eduardo Fernández-Pascual.
+Correspondence: Departamento de Biología de Organismos y Sistemas, Universidad de Oviedo; C/ Catedrático Rodrigo Uría, 33006 Oviedo/Uviéu, Spain; Tel.: +34985104787; email: efernandezpascual@gmail.com"
 date: "`r format(Sys.time(), '%d/%m/%Y')`"
 output: 
-  bookdown::html_document2
+  bookdown::word_document2:
+    reference_docx: ijctemplate.docx
 bibliography: refs.bib
@@
 link-citations: yes
+always_allow_html: true
 ---
@@
 ```{r message = FALSE, echo = FALSE, warning = FALSE}
-library(here); library(kableExtra)
+library(here); library(kableExtra); library(flextable)
 load(here("results", "logs.RData"))
@@
 ```
+
+# Graphical abstract
+
+![Mires are semi-terrestrial wetlands that remain waterlogged for most of the year.
+Waterlogged soils experience a thermal buffer effect that isolates them from surrounding temperature variations.
+This article tests the buffer effect in waterlogged mire soils and adjacent sry soils.](Laguna de Reconcos.jpg)
+
+# Abstract
+
+# Keywords
+
+Microniche, fen, bog, dataloggers, azonal habitats, wetlands, Cantabrian Mountains
 
@@
 A visual inspection of the time series shows more homogeneous temperatures in the waterlogged time series, which are almost always encompassed within the dry time series.
-The bioclimatic variables (Table \@ref(tab:table2)) support this impression.
+The bioclimatic variables (Fig.
+\@ref(fig:fig2)) support this impression.
 The maximum temperatures of the warmest months are usually higher in the dry measuring points, by more than 6 ºC in some sites, both at low (El Molinucu, La Malva) and high (La Recoleta) elevations.
@@
 The mean diurnal range of temperatures is wider in the dry sites, as is annual temperature range; and again the effect is larger in the low sites and in La Recoleta.
-The t-tests supported the original hypotheses (Fig.
-\@ref(fig:fig2)); namely waterlogged measuring points had (a) smaller diurnal fluctuactions (t = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.estimate) %>% round(2)` ºC); (b) lower maximums (t = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.estimate) %>% round(2)` ºC); (c) higher minimums (t = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.estimate) %>% round(2)` ºC), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.estimate) %>% round(2)` ºC).
+The t-tests supported the original hypotheses; namely waterlogged measuring points had (a) smaller diurnal fluctuactions (t = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.estimate) %>% round(2)` ºC); (b) lower maximums (t = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.estimate) %>% round(2)` ºC); (c) higher minimums (t = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.estimate) %>% round(2)` ºC), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.estimate) %>% round(2)` ºC).
 
@@
 
-Less effect in winter could be because this is soil (snow, etc)
+The results presented here prove the existence of a thermal buffer effect in waterlogged mire soils, when compared with adjacent dry soils.
+The mire thermal buffer had been compared previously with air temperatures derived from models, with generally similar results [@RN2356; @RN4675].
+The mire buffer had also been compared to dry soils but only during the growing season [@RN3204; @RN3060]; our results confirm those findings and extend them to the full year.
+Therefore, there exists a thermal buffer effect in mire soils that makes their temperatures less extreme than the surrounding landscape.
+This highlights the importance of using fine-scale microclimatic data to assess vegetation responses to climate change [@RN4683].
+
+One important difference with previous studies is the importance of the buffer during winter.
+The articles that had used model air temperatures as a control concluded that the buffer effect was stronger at the cold end of the thermal gradient, i.e.
+in winter and at night [@RN2356; @RN4675].
+In the case of this investigation, the situation was reverse: the effect was weaker when considering the minimum temperatures of the cold period.
+This indicates the importance of identifying root temperatures when working with plant communities.
+At high elevations of the study region, the soil can remain covered by snow for periods of winter, and this has its own insulating effect on soil temperatures, an effect which is not captured when using model derived air temperatures [@RN2392].
+Nonetheless, these records were made in South-Western Euope, and they should be repeated at more northern latitudes, where the winter effect may be more relevant [@RN2356; @RN4675].
+
+The buffering effect was much stronger during the summer.
+This was most noticeable in the two lowest elevations.
+Specially in La Malva, very high temperatures (> 40ºC) were recorded during summer.
+This site is in a south-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681].
+This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344].
+
+In summary, this article shows that waterlogged mire soils have a thermal buffer when compared to adjacent soils, contributing to their behaviour as mild island habitats in a landscape that can be more thermally variable [@RN4675].
+The effect occurs during cold and warm periods, but it is stronger during the summer at least in the study area.
+Future recording schemes are needed to obtain local soil temperatures from other latitudes, and from more microtopographies within the same mire.
+
+# Data availability
+
+Upon publication, the original temperature records along with R analysis scripts will be uploaded to a public repository in GitHub.
+
+# Acknowledgements
+
+E.F.P. received financial support from the Government of Asturias and the FP7 – Marie Curie - COFUND programme of the European Commission (Grant ‘Clarín’ ACB17-19).
 
@@
 ```{r table1, echo = FALSE}
-table1 %>% kable(caption = "Mire sites included in this study, indicating the type of fen, the elevation, coordinates, and length of the temperature recording period.", booktabs = TRUE) %>% kable_styling(latex_options = c("striped", "hold_position"))
-```
-
-```{r table2, echo = FALSE}
-table2 %>% kable(caption = "Bioclimatic variables calculated for the mire sites, using the hourly temperature records.", booktabs = TRUE) %>% kable_styling(latex_options = c("striped", "hold_position"))
+regulartable(table1) %>% width(width = 1) %>% set_caption("Mire sites included in this study, indicating the type of fen, the elevation, coordinates, and length of the temperature recording period.")
 ```

### 2020-08-20 doc/manuscript/v2_IJOC/ms2.Rmd
map of the sites

---
title: "Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils"
subtitle: "Short title: Testing the thermal buffer of mire groundwater"
output: 
  bookdown::word_document2:
    reference_docx: ijctemplate.docx
bibliography: refs.bib
biblio-style: international-journal-of-climatology.csl
link-citations: yes
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(here); library(tidyverse); library(flextable)
load(here("results", "logs.RData"))
load(here("results", "MSoutput.RData"))
```

Eduardo Fernández-Pascual^1^ ^+^, Eva Correia-Álvarez^2^

^1^ Universidad de Oviedo, ^2^ Independent researcher

^+^ Correspondence: Departamento de Biología de Organismos y Sistemas, Universidad de Oviedo, C/ Catedrático Rodrigo Uría, 33006 Oviedo/Uviéu, Spain. Email: efernandezpascual@gmail.com. Telephone: +34985104787.

# Financial support

E.F.P. received financial support from the Government of Asturias and the FP7 – Marie Curie - COFUND programme of the European Commission (Grant ‘Clarín’ ACB17-19).

# Graphical abstract

![Mires are semi-terrestrial wetlands that remain waterlogged for most of the year. Mire groundwater produces a thermal buffer effect that insulates these habitats from the surrounding landscape, especially at the warm end of the climatic spectrum, i.e. during summer and at lower elevations. This highlights the potential refugial character of mires from global warming, and the need to integrate in situ microclimate measurements into climate change models.](Laguna de Reconcos.jpg)

# Abstract

Ecosystems adapt differently to global warming through microclimate factors. Mires are sensitive habitats that strongly rely on local soil properties, which makes them a good model to understand how local climate parameters counteract the effects of climate change. We tested the hypothesis that temperature in waterlogged mire soils are less variable and extreme than in adjacent dry soils.

We buried dataloggers at 5 cm depth in waterlogged and dry points in 8 mires of the Cantabrian Mountains (Spain, Southwestern Europe) and recorded soil temperatures for c. 5 years. We also compared our local measures with air temperatures predicted by the CHELSA model.

Waterlogged soils had less diurnal thermal amplitude (-2.3 ºC), less annual thermal amplitude (-5.1 ºC), cooler summer maximums (-4.3 ºC) and warmer winter minimums (+0.8 ºC). CHELSA failed to predict soil temperatures except for the summer maximums in dry soils and the winter minimums in both dry and waterlogged soils.

We conclude that mire soils show a thermal buffer effect that insulates them from the surrounding landscape. This effect is stronger at the warm end of the climatic spectrum, i.e. during summer and at lower elevations. These results highlight the potential refugial character of mires under global warming, and the need to integrate microclimate measurements into climate change models.

# Keywords

bog, CHELSA, climatic model, datalogger, fen, peatland, soil temperature, wetland

# Introduction

Climate change [@RN3586] affects global biodiversity, from drylands [@RN4672] to forests [@RN4673] and oceans [@RN4674]. Traditional models of species responses to global warming are based on macroclimatic data from weather stations. Recently, several authors have pointed out the need of complementing these models with *in situ* microclimatic measures [@RN4758; @RN4759; @RN4760; @RN4761]. As has being shown for European forests [@RN4763], ecosystems usually respond to broad climatic changes through local processes. An accurate prediction of species responses requires to focus on physiologically-relevant variables related to critical plant growth periods [@RN4765].

Mires are permanent semi-terrestrial peatlands whose soils remain waterlogged but not inundated during most of the year [@RN3161]. They are azonal habitats whose existence depends on local soil properties rather than macroclimatic zonation [@RN3328]. Groundwater can produce a buffer effect on soil temperature, keeping soils warmer than air during cold periods, and vice versa [@RN3344; @RN3201]. Root-zone temperature is a major determinant of plant ecophysiology [@RN3024], so the groundwater buffer effect is expected to allow mire plants to live in a wider range of air temperatures than they could otherwise. Indeed, mires have a relatively homogeneous flora despite being distributed from Mediterranean to Boreal biomes, and from low valleys to the alpine belt [@RN4678]. Recently, soil temperature measurements have become available for mires of North America [@RN3204], Western Europe [@RN2356] and Central Europe [@RN4675]. Their comparison with air temperatures derived from models has shown that mire soils are indeed warmer in winter and cooler in summer, thus giving support to the existence of the groundwater thermal buffer [@RN2356; @RN4675]. Furthermore, the effect has been linked to the composition of mire flora and fauna [@RN4675; @RN4679], the growth rings of mire trees [@RN3060] and the role of mires as glacial refugia [@RN2513; @RN4680].

Mires meet certain criteria that make them especially vulnerable to climate change [@RN4675]: (a) preponderance of species that evolved under a cold climate; (b) low productivity due to nutrient limitation, making them sensitive to increased nutrient cycling caused by warming [@RN4676]; and (c) scattered distribution pattern, which limits species dispersal and migration [@RN4677]. Therefore, mires are priority habitats for biodiversity conservation, harbouring high numbers of endangered species [@RN3122], and supporting highly-adapted floras in spatially-reduced areas [@RN2960]. Worryingly, mires retain high levels of methane which can be released due to global warming [@RN4764]. Habitat distribution models have predicted a loss of mire surface as a consequence of ongoing climate change [@RN2937].

It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change. As is the case for all azonal habitats, locally measured temperatures are essential to understand this response. Available references recorded temperatures only on waterlogged soils, using model-derived air temperatures for comparison [@RN2356; @RN4675; @RN4679]; or recorded temperatures in both wet and dry spots but for less than a year, lacking representativeness throughout the growth cycle of mire vegetation [@RN3204; @RN3060]. This article provides the first measurement of the thermal buffer against surrounding non-mire areas, based on soil temperatures recorded during a period of five years. These measures are used to test the hypotheses that, when compared to adjacent dry soils, waterlogged mire soil are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales. In addition, we compare *in situ* measurements with data derived from the CHELSA climatic models [@RN4766].

# Materials and methods

This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W) (Fig. \@ref(fig:fig1)a). Local geography and climate are dominated by the Cantabrian Mountains (> 1500 m above sea level), which run parallel to the coast and trap the prevailing NW Atlantic winds. The resulting humid climate harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246]. Rain-fed raised bogs and acid valley mires can be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes. Glacial lakes undergoing silting develop transition mires and quaking bogs communities in the water-to-land transition. Spring fens appear in the mountains above 1000 m; they range from soft-water poor fens on acid bedrocks, to alkaline calcareous fens on limestone.

We selected `r length(unique(temperatures$Logger))/2` mire sites representing the regional elevation gradient of mire vegetation and the different mire types (Table 1). In each site, we buried two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany) at a depth of 5 cm below the upper layer of the soil: one datalogger in a flat waterlogged spot within the mire; the other one in the close vicinity, but in a flat and dry upland area outside the mire (Fig. \@ref(fig:fig1)b). The vegetation was always either mire or pasture, with no shrubs, trees or any other landscape features shading the measuring points. Dataloggers recorded temperature on an hourly basis and stayed on site for approximately five years, after which we retrieved them and downloaded their records. At the moment of retrieval, the internal clock of all dataloggers had not deviated for more than four hours.

To clean the logs we took the following steps: (i) removing records from the first week after installation, to account for the installation process and the settling of the soils; and (ii) keeping only time series with records for both the dry and waterlogged points, because some of the dataloggers had failed at different points in time. Afterwards, we calculated four bioclimatic variables for each datalogger: (1) the mean diurnal range; i.e. the average for the whole period of the daily differences between the maximum and the minimum temperatures recorded in the day; (2) the maximum temperature of the warmest month; i.e. the average of the daily maximum records, for the warmest month; (3) the minimum temperature of the coldest month; i.e. the average of the daily minimum records, for the coldest month; and (4) the annual range; i.e. the difference between the maximum temperature of the warmest month and the minimum temperature of the coldest month. To compare our measurements with model-based predictions of air temperature, we downloaded from CHELSA the same bioclimatic variables for our measuring coordinates.

To test if the differences between the dry and the waterlogged points of each site was significant, we used paired t-tests. Tests were one-tailed, according to the following hypotheses: the dry point would have a higher diurnal range, a higher maximum temperature, a lower minimum temperature, and a higher annual range. To test whether the CHELSA values predicted our measurements, we fitted linear models.

# Results

Dataloggers recorded temperatures for five years in five of the sites, four years in two, and two years in one. Time series showed considerably less variation at waterlogged points, for all eight site comparisons (Fig. \@ref(fig:fig2)).

Bioclimatic variables (Table 2 and Fig. \@ref(fig:fig3)) supported this notion. The mean annual range was wider at dry points, as was the mean diurnal range. Maximum temperatures of the warmest months were usually higher at dry measuring points. The opposite was true for the minimums of the coldest months, in which case temperature was generally colder at dry points, although the difference was less pronounced than for the maximums (Fig. \@ref(fig:fig3)).

Patterns regarding the maximum temperatures and annual range were especially noticeable  at both low (El Molinucu, La Malva) and high (La Recoleta) elevations. The pattern for the diurnal range, however, was more prominent just at the low sites (El Molinucu, La Malva), whereas the minimums of the coldest months showed no specific pattern (Table 2).

T-tests supported the original hypotheses; namely waterlogged measuring points had (a) smaller diurnal fluctuations (t = `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.estimate) %>% round(2)` ºC); (b) lower maximums (t = `r ttests %>% filter(Trait == "Summer max") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Summer max") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Summer max") %>% pull(t.estimate) %>% round(2)` ºC); (c) higher minimums (t = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.estimate) %>% round(2)` ºC), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Annual range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Annual range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Annual range") %>% pull(t.estimate) %>% round(2)` ºC).

The CHELSA climatic model (Table 3 and Fig. \@ref(fig:fig4)) predicted well the minimums of the coldest month, especially at dry points (R^2^ = 0.78) but also at waterlogged points (R^2^ = 0.66). CHELSA also predicted to some extent the maximums of the warmest month at dry points (R^2^ = 0.42), but failed to predict waterlogged points (R^2^ = 0.16). CHELSA did not predict the values for the annual and diurnal range, at both dry and waterlogged points.

# Discussion

The results presented here prove the existence of a thermal buffer effect in waterlogged mire soils, when compared with adjacent dry soils. The mire thermal buffer had been compared previously with air temperatures derived from models, with generally similar results [@RN2356; @RN4675]. The mire buffer had also been compared to dry soils but only during the growing season [@RN3204; @RN3060]; our results confirm those findings and extend them to the full year. Therefore, there exists a thermal buffer effect in mire soils that makes their temperatures less extreme than the surrounding landscape. This highlights the importance of using fine-scale microclimatic data to assess vegetation responses to climate change [@RN4683; @RN4758; @RN4763].

One important difference with previous studies is the importance of the buffer during winter. The articles that had used model air temperatures as a control concluded that the buffer effect was stronger at the cold end of the thermal gradient, i.e. in winter and at night [@RN2356; @RN4675]. In the case of this investigation, the situation was reverse: the effect was weaker when considering the minimum temperatures of the cold period. This indicates the importance of identifying root temperatures when working with plant communities. At high elevations of the study region, the soil can remain covered by snow for periods of winter, and this has its own insulating effect on soil temperatures [@RN2392]. Indeed, snow cover has being described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760].

In concordance with our results, the air temperature measured at 0.5 m from the surface of boreal bogs is lower at the wetter zones, at least during the warmest months [@RN4761]. Thus, the water buffering effect seems to take place both at southern and northern latitudes and may affect not only the temperature in the soil but also the air temperature within certain highness from the soil. Nonetheless, Philippov *et al.* did not record winter months, so more monitoring is needed at more northern latitudes, where the winter effect may be more relevant [@RN4675].

The buffering effect was much stronger during the summer. This was most noticeable in the two lowest elevations, El Molinucu and La Malva. La Malva showed specially high summer temperatures within the dry soils (> 40 ºC). This site is in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681]. This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344]. The buffering was also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog.

Most models currently employed to predict vegetation and species distribution use macroclimatic parameters, like the ones obtained from CHELSA [@RN4766]. This study revealed that CHELSA can predict quite well the minimum temperatures at the coldest month in the mire habitats. However, it fails to predict temperatures at the warm extremes, especially at the waterlogged points. This brings out the importance of studying local factors as drivers of microclimatic changes [@RN4762]. These factors contribute to landscape heterogeneity, producing safe sites that can act as micro-refugees and buffer species from regional climatic warming.

In summary, this article shows that waterlogged mire soils have a thermal buffer when compared to adjacent soils, contributing to their behaviour as mild island habitats in a landscape that can be more thermally variable [@RN4675]. The effect occurs during cold and warm periods, but it is stronger during the summer, at least in the study area. Future recording schemes are needed to obtain local soil temperatures from other latitudes, and from more microtopographies within the same mire. Our study provides useful microclimate parameters to improve the current models that predict the impact of global warming on specific ecosystems.

# Data availability

Upon publication, the original temperature records along with R analysis scripts will be uploaded to a public repository in GitHub.

# References

::: {#refs}
:::

# Tables

```{r table1, echo = FALSE}
regulartable(table1) %>% width(width = 1) %>% set_caption("Table 1: Mire sites included in this study, indicating the type of fen, the elevation, coordinates, and length of the temperature recording period. Coordinates are in decimal degrees WGS84.")
```

```{r table2, echo = FALSE}
regulartable(table2) %>% width(width = 1) %>% set_caption("Table 2: Groundwater buffer effect per mire and bioclimatic variable. The buffer effect was calculated as the difference between the value in the waterlogged and the dry points.")
```

```{r table3, echo = FALSE}
regulartable(table3) %>% width(width = 1) %>% set_caption("Table 3: Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation.")
```

# Figures

```{r fig1, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "(a) Location of the study sites in northern Spain. (b) Representation of the sampling design at El Riotuertu site. The position of the dry and waterlogged temperature logggers is indicated, the blue area represents the extent of the waterlogged mire habitat.", fig.align = "center", fig.width=7, fig.height=6}
knitr::include_graphics("../Fig1.png")
```

```{r fig2, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites. The blue series was recorded within the mire, in a waterlogged area. The red series was recorded in a neighbouring dry area. Dataloggers were buried at 5 cm depth.", fig.align = "center", fig.width=7, fig.height=6}
plot2
```

```{r fig3, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Average bioclimatic variables in the dry and waterlogged points. The bars represent the mean value, and the brackets the standard error of 8 records.", fig.align = "center", fig.width=7, fig.height=3}
plot3
```

```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Scatter plots of the bioclimatic variables predicted by CHELSA air temperatures vs. soil temperatures measured in situ, in dry and waterlogged points.", fig.align = "center", fig.width=7, fig.height=7}
plot4
```

### 2020-08-26 doc/manuscript.Rmd
readme, license, cleaning

--- 2020-08-20 doc/manuscript/v2_IJOC/ms2.Rmd
@@
 
-# Graphical abstract
-
-![Mires are semi-terrestrial wetlands that remain waterlogged for most of the year.
-Mire groundwater produces a thermal buffer effect that insulates these habitats from the surrounding landscape, especially at the warm end of the climatic spectrum, i.e.
-during summer and at lower elevations.
-This highlights the potential refugial character of mires from global warming, and the need to integrate in situ microclimate measurements into climate change models.](Laguna de Reconcos.jpg)
-
 # Abstract
@@
 Mires are sensitive habitats that strongly rely on local soil properties, which makes them a good model to understand how local climate parameters counteract the effects of climate change.
-We tested the hypothesis that temperature in waterlogged mire soils are less variable and extreme than in adjacent dry soils.
+We quantified the temperature buffering effect in waterlogged mire soils as compared with adjacent dry soils.
 
@@
 Waterlogged soils had less diurnal thermal amplitude (-2.3 ºC), less annual thermal amplitude (-5.1 ºC), cooler summer maximums (-4.3 ºC) and warmer winter minimums (+0.8 ºC).
-CHELSA failed to predict soil temperatures except for the summer maximums in dry soils and the winter minimums in both dry and waterlogged soils.
+CHELSA air temperatures only correlated significantly (p < 0.05) with the winter minimum temperatures (Pearson's r > 0.83), and CHELSA predictions were less accurate (higher RMSE) for waterlogged soils, except for the summer maximums.
 
@@
 Climate change [@RN3586] affects global biodiversity, from drylands [@RN4672] to forests [@RN4673] and oceans [@RN4674].
+An accurate prediction of species responses requires to focus on physiologically-relevant variables related to critical plant growth periods [@RN4765].
 Traditional models of species responses to global warming are based on macroclimatic data from weather stations.
 Recently, several authors have pointed out the need of complementing these models with *in situ* microclimatic measures [@RN4758; @RN4759; @RN4760; @RN4761].
-As has being shown for European forests [@RN4763], ecosystems usually respond to broad climatic changes through local processes.
-An accurate prediction of species responses requires to focus on physiologically-relevant variables related to critical plant growth periods [@RN4765].
+As has been shown for European forests [@RN4763], ecosystems usually respond to broad climatic changes through local processes.
+Increasingly there are downscaling efforts based on atmospheric microclimate networks distributed in complex terrain that focus on topographic effects on air temperature [@RN3279], but hydrologic factors under edaphic control requiere more attention.
 
 Mires are permanent semi-terrestrial peatlands whose soils remain waterlogged but not inundated during most of the year [@RN3161].
-They are azonal habitats whose existence depends on local soil properties rather than macroclimatic zonation [@RN3328].
-Groundwater can produce a buffer effect on soil temperature, keeping soils warmer than air during cold periods, and vice versa [@RN3344; @RN3201].
+The term mire encompasses peatlands that are classified as either ombrotrophic (rain-fed bogs) or minerotrophic (groundwater-fed fens).
+These are azonal habitats whose existence depends on local soil properties rather than macroclimatic zonation [@RN3328].
+It has been known for a relatively long time that groundwater can produce a buffer effect on soil temperature, keeping soils warmer than air during cold periods, and vice versa [@RN4940; @RN3344; @RN3201].
+Moisture reduces exposure to temperature extremes, facilitating climate refugia [@RN4942].
 Root-zone temperature is a major determinant of plant ecophysiology [@RN3024], so the groundwater buffer effect is expected to allow mire plants to live in a wider range of air temperatures than they could otherwise.
-Indeed, mires have a relatively homogeneous flora despite being distributed from Mediterranean to Boreal biomes, and from low valleys to the alpine belt [@RN4678].
-Recently, soil temperature measurements have become available for mires of North America [@RN3204], Western Europe [@RN2356] and Central Europe [@RN4675].
+Indeed, mires have a relatively homogeneous flora across their range despite their broad distribution.
+In Europe, fens are distributed from the Iberian Peninsula to Fennoscandia and from low valleys to the alpine belt [@RN4678].
+In the United States, fens are distributed across the glaciated Midwest and Northeast, as well as portions of the Appalachian Mountains and mountainous West [@RN4933].However, continued aridification is expected to significantly reduce the overall extent of wetlands as it has been shown in the Midwestern United States [@RN4934].
+In the arctic, temperate-continental and suboceanic regions of boreal Russia mires make up an important part of the landscapce, from 20 to 80 % of the surface of different regions stretching from Europe to the Pacific [@RN4937].
+
+Clara M.
+Frederick [-@RN4940] showed the existence of a buffer effect comparing soil temperatures taken at Cedar Bog (Ohio, United States) with temperatures from a neighbouring agricultural station.
+Recently, more soil temperature measurements have become available for mires of North America [@RN3204], Western Europe [@RN2356] and Central Europe [@RN4675].
 Their comparison with air temperatures derived from models has shown that mire soils are indeed warmer in winter and cooler in summer, thus giving support to the existence of the groundwater thermal buffer [@RN2356; @RN4675].
@@
 Therefore, mires are priority habitats for biodiversity conservation, harbouring high numbers of endangered species [@RN3122], and supporting highly-adapted floras in spatially-reduced areas [@RN2960].
-Worryingly, mires retain high levels of methane which can be released due to global warming [@RN4764].
-Habitat distribution models have predicted a loss of mire surface as a consequence of ongoing climate change [@RN2937].
+Worryingly, mires retain high levels of methane and carbon which can be released due to global warming [@RN4764].
+Habitat distribution models have predicted a loss of mire area as a consequence of ongoing climate change [@RN2937].
 
-It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change.
+It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change [@RN4936].
+Both bogs and fens are actively peat forming and are dependant on precipitation, whereas fens rely also on sources of telluric water from mineral ground.
+Moreover they are composed from microreliefs and a specific plant cover zonation which makes them specially sensitive to small changes in wetness.
+Climate change affects the quantity, timing and spacial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935].
+Global warming would also result in warmer groundwater delivered to fens and in drier conditionts, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].
+
 As is the case for all azonal habitats, locally measured temperatures are essential to understand this response.
-Available references recorded temperatures only on waterlogged soils, using model-derived air temperatures for comparison [@RN2356; @RN4675; @RN4679]; or recorded temperatures in both wet and dry spots but for less than a year, lacking representativeness throughout the growth cycle of mire vegetation [@RN3204; @RN3060].
+Available references recorded temperatures only on waterlogged soils, using model-derived air temperatures for comparison [@RN2356; @RN4675; @RN4679]; or recorded temperatures in both wet and dry spots but for less than a year, lacking representation throughout the growth cycle of mire vegetation [@RN3204; @RN3060].
 This article provides the first measurement of the thermal buffer against surrounding non-mire areas, based on soil temperatures recorded during a period of five years.
-These measures are used to test the hypotheses that, when compared to adjacent dry soils, waterlogged mire soil are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales.
+These measures can be used to support previous evidence about the general effects of soil moisture on thermal buffering [@RN3204; @RN2356; @RN4675; @RN4936; @RN4942] and to determine the magnitude of this buffering within microrefugia habitats.
+Specifically, we tested the hypotheses that, when compared to adjacent dry soils, waterlogged mire soil are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales.
 In addition, we compare *in situ* measurements with data derived from the CHELSA climatic models [@RN4766].
@@
 Local geography and climate are dominated by the Cantabrian Mountains (> 1500 m above sea level), which run parallel to the coast and trap the prevailing NW Atlantic winds.
-The resulting humid climate harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246].
-Rain-fed raised bogs and acid valley mires can be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes.
-Glacial lakes undergoing silting develop transition mires and quaking bogs communities in the water-to-land transition.
-Spring fens appear in the mountains above 1000 m; they range from soft-water poor fens on acid bedrocks, to alkaline calcareous fens on limestone.
+In the study area, average annual precipitation ranges from 800 mm at the low elevations to 1800 mm at the high mountains, while average annual temperature ranges from 12.5 ºC to 5 ºC (source = Agencia Estatal de Meteorología, http://www.aemet.es/es/serviciosclimaticos/).
+Under the most extreme emissions scenario, the latest projection expect temperatures to increase by 4 ºC and precipitation to decrease by c.
+10 % by the end of the century, although the precipiation projections are subjected to great uncertainty [@RN4939].
+This humid climate harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246].
+Rain-fed raised bogs (https://eunis.eea.europa.eu/habitats/260) are very rare and appear locally on top of acid valley mires (https://eunis.eea.europa.eu/habitats/526) which be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes.
+Glacial lakes undergoing silting develop transition mires and quaking bogs communities (https://eunis.eea.europa.eu/habitats/1637) in the water-to-land transition.
+Spring fens appear in the mountains above 1000 m; they range from soft-water poor fens (https://eunis.eea.europa.eu/habitats/279) on acid bedrocks, to alkaline calcareous fens (https://eunis.eea.europa.eu/habitats/277) on limestone.
 
 We selected `r length(unique(temperatures$Logger))/2` mire sites representing the regional elevation gradient of mire vegetation and the different mire types (Table 1).
-In each site, we buried two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany) at a depth of 5 cm below the upper layer of the soil: one datalogger in a flat waterlogged spot within the mire; the other one in the close vicinity, but in a flat and dry upland area outside the mire (Fig.
+Although we classify two of these sites as rain-fed bogs based on their vegetation (*Oxycocco-Sphagnetea* Br.-Bl.
+et Tx.
+ex Westhoff
+et al.
+1946), we must note that these bogs are very poorly developped over valley mires and in tight connection to the underlying water table coming from streams, therefore all of our sites have waterlogging feed by streams or springs.
+In each site, we buried two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil.
+In this ecosystem, it is at this depth that true soil begins to develop, under the porous upper layers made up of live mosses.
+We installed one datalogger in a flat waterlogged spot within the mire; the other one in the close vicinity, but in a flat and dry upland area outside the mire (Fig.
 \@ref(fig:fig1)b).
 The vegetation was always either mire or pasture, with no shrubs, trees or any other landscape features shading the measuring points.
-Dataloggers recorded temperature on an hourly basis and stayed on site for approximately five years, after which we retrieved them and downloaded their records.
+Dataloggers recorded temperature once every hour and stayed on site for approximately five years, after which we retrieved them and downloaded their records.
 At the moment of retrieval, the internal clock of all dataloggers had not deviated for more than four hours.
 
+All data processing and analysis was performed in R [@RN2315].
+The original data, R code for the analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/mires.
 To clean the logs we took the following steps: (i) removing records from the first week after installation, to account for the installation process and the settling of the soils; and (ii) keeping only time series with records for both the dry and waterlogged points, because some of the dataloggers had failed at different points in time.
@@
 the difference between the maximum temperature of the warmest month and the minimum temperature of the coldest month.
+These are classical bioclimatic variables that indicate the limiting factors (extreme temperatures) and seasonality (diurnal and annual range) that organisms must tolerate [@RN4943].
 To compare our measurements with model-based predictions of air temperature, we downloaded from CHELSA the same bioclimatic variables for our measuring coordinates.
 
-To test if the differences between the dry and the waterlogged points of each site was significant, we used paired t-tests.
+To test if the differences between dry and waterlogged points were significant, we used paired t-tests (n = 8 paired sites per each of the four bioclimatic variables).
 Tests were one-tailed, according to the following hypotheses: the dry point would have a higher diurnal range, a higher maximum temperature, a lower minimum temperature, and a higher annual range.
-To test whether the CHELSA values predicted our measurements, we fitted linear models.
+To compare our soil measurements with the CHELSA air temperatures we calculated Pearson's correlation between both sets of values, and we also calculated the root-mean-square error (RMSE) of the CHELSA values as a measure of their accuracy (n = 8 sites per combination of bioclimatic variable and dry/waterlogged condition) using the package *Metrics* [@RN4941].
 
@@
 
-T-tests supported the original hypotheses; namely waterlogged measuring points had (a) smaller diurnal fluctuations (t = `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.estimate) %>% round(2)` ºC); (b) lower maximums (t = `r ttests %>% filter(Trait == "Summer max") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Summer max") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Summer max") %>% pull(t.estimate) %>% round(2)` ºC); (c) higher minimums (t = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.estimate) %>% round(2)` ºC), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Annual range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Annual range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Annual range") %>% pull(t.estimate) %>% round(2)` ºC).
+T-tests supported the original hypotheses; namely waterlogged measuring points had (a) smaller diurnal fluctuations (t = `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.estimate) %>% round(2)` ºC, 95 % CI = -∞, `r ttests %>% filter(Trait == "Diurnal range") %>% pull(t.conf.int.2.) %>% round(2)`); (b) lower maximums (t = `r ttests %>% filter(Trait == "Summer max") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Summer max") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Summer max") %>% pull(t.estimate) %>% round(2)` ºC, 95 % CI = -∞, `r ttests %>% filter(Trait == "Summer max") %>% pull(t.conf.int.2.) %>% round(2)`); (c) higher minimums (t = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.estimate) %>% round(2)` ºC, 95 % CI = `r ttests %>% filter(Trait == "Winter min") %>% pull(t.conf.int.1.) %>% round(2)`, ∞), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Annual range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Annual range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Annual range") %>% pull(t.estimate) %>% round(2)` ºC, 95 % CI = -∞, `r ttests %>% filter(Trait == "Annual range") %>% pull(t.conf.int.2.) %>% round(2)`).
 
 The CHELSA climatic model (Table 3 and Fig.
-\@ref(fig:fig4)) predicted well the minimums of the coldest month, especially at dry points (R^2^ = 0.78) but also at waterlogged points (R^2^ = 0.66).
-CHELSA also predicted to some extent the maximums of the warmest month at dry points (R^2^ = 0.42), but failed to predict waterlogged points (R^2^ = 0.16).
-CHELSA did not predict the values for the annual and diurnal range, at both dry and waterlogged points.
+\@ref(fig:fig4)) provided air temperatures that were significantly correlated (p < 0.05) with soil temperatures only in the case of the minimums of the coldest month, both at dry points (Pearson's r = 0.90) and waterlogged points (Pearson's r  = 0.84).
+The correlation between CHELSA and soil values was marginally significant (p = 0.05) in the case of the maximums of the warmest month, but only at dry points (Pearson's r = 0.71).
+The values of CHELSA did not correlate with soil temperatures in the rest of the cases (p > 0.05).
+RMSE indicated that the predictions of CHELSA were more accurete at dry points for the cases of the annual range, the diurnal range and the winter min; the CHELSA values were more accurate at waterlogged points in the case of the summer max.
 
@@
 
-The results presented here prove the existence of a thermal buffer effect in waterlogged mire soils, when compared with adjacent dry soils.
+The results presented here quantify the size of the thermal buffer effect that takes place in waterlogged mire soils [@RN4940], when compared with adjacent dry soils.
 The mire thermal buffer had been compared previously with air temperatures derived from models, with generally similar results [@RN2356; @RN4675].
-The mire buffer had also been compared to dry soils but only during the growing season [@RN3204; @RN3060]; our results confirm those findings and extend them to the full year.
+The mire buffer had also been compared to dry soils at 10 cm depth [@RN3204; @RN3060]; our results confirm those findings at 5 cm depth and extend them to the full year.
+Thus, the pattern is reproducible among years, out to five years.
 Therefore, there exists a thermal buffer effect in mire soils that makes their temperatures less extreme than the surrounding landscape.
+In our study, this effect seems stronger during warmer periods.
 This highlights the importance of using fine-scale microclimatic data to assess vegetation responses to climate change [@RN4683; @RN4758; @RN4763].
@@
 At high elevations of the study region, the soil can remain covered by snow for periods of winter, and this has its own insulating effect on soil temperatures [@RN2392].
-Indeed, snow cover has being described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760].
+Indeed, snow cover has being described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760].Taking into account that the mentioned studies used a 5 cm depth for observation, it may be worth to use a deeper depth in winter to avoid this snow effect.
 
 In concordance with our results, the air temperature measured at 0.5 m from the surface of boreal bogs is lower at the wetter zones, at least during the warmest months [@RN4761].
-Thus, the water buffering effect seems to take place both at southern and northern latitudes and may affect not only the temperature in the soil but also the air temperature within certain highness from the soil.
+Thus, the water buffering effect seems to take place both at southern and northern latitudes and may affect not only the temperature in the soil but also the air temperature within certain height from the soil.
 Nonetheless, Philippov *et al.* did not record winter months, so more monitoring is needed at more northern latitudes, where the winter effect may be more relevant [@RN4675].
@@
 La Malva showed specially high summer temperatures within the dry soils (> 40 ºC).
-This site is in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681].
+It must be noted that this difference of > 15ºC between the dry and waterlogged points, separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939].
+Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.).
 This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344].
-The buffering was also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog.
+The buffering was also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog, i.e.
+a former glacial lake undergoing silting.
+However, more studies focusing on each kind of habitat are needed in order to understand the relation between hydrology conditions and microclimatic effects.
 
 Most models currently employed to predict vegetation and species distribution use macroclimatic parameters, like the ones obtained from CHELSA [@RN4766].
-This study revealed that CHELSA can predict quite well the minimum temperatures at the coldest month in the mire habitats.
-However, it fails to predict temperatures at the warm extremes, especially at the waterlogged points.
+This study revealed that CHELSA values correlate relatively well with soil temperatures in the case of the minimum temperatures at the coldest month, but not with the rest of the bioclimatic variables under study.
+Moreover, CHELSA predictions were less accurate for the waterlogged soils, except for the summer max, which might be the consequence of the noise introduced by sun-heated soil.
+In adittion to the water buffering effect, other traits may explain the differences between our results and CHELSEA values like the height of temperature measurement, 5 cm below the upper layer of the soil in our study versus 2 m above the ground in CHESA; the temporal scale of observation, hourly based for a 5 years period in our study versus 30 arc sec resolution for a 34 years period; and the spatial scale, (AÑADIR).
 This brings out the importance of studying local factors as drivers of microclimatic changes [@RN4762].
@@
 
-In summary, this article shows that waterlogged mire soils have a thermal buffer when compared to adjacent soils, contributing to their behaviour as mild island habitats in a landscape that can be more thermally variable [@RN4675].
+In summary, this article shows that waterlogged mire soils have a thermal buffer when compared to adjacent soils, contributing to their behaviour as mild island habitats in a landscape that can be more thermally variable [@RN4936; @RN4675].
 The effect occurs during cold and warm periods, but it is stronger during the summer, at least in the study area.
+However, the climatic buffering within microrefugia would strongly rely on the water regime, a key factor for their conservation.
+Long-term decreases in precipitation could reduce the groundwater discharge into the mires, with potentially fatal consequences if the mire dries out during warmer summers.
+For the study area, the latests projections for the end of the century [@RN4939] consider a decrease of c.
+10 % in annual precipitation.
 Future recording schemes are needed to obtain local soil temperatures from other latitudes, and from more microtopographies within the same mire.
+We must consider that the microclimatic effect is not limited to mires as it can also be found in other habitats related with thawing snow like alpine and arctic-alpine slopes [@RN3279; @RN4938].
 Our study provides useful microclimate parameters to improve the current models that predict the impact of global warming on specific ecosystems.
@@
 
-Upon publication, the original temperature records along with R analysis scripts will be uploaded to a public repository in GitHub.
+The original data, R code for the analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/mires.
+Upon publication, a version of record of the repository will be deposited in Zenodo.
 
@@
 ```{r table1, echo = FALSE}
-regulartable(table1) %>% width(width = 1) %>% set_caption("Table 1: Mire sites included in this study, indicating the type of fen, the elevation, coordinates, and length of the temperature recording period.
+regulartable(table1) %>% width(width = 1) %>% set_caption("Table 1: Mire sites included in this study, indicating the type of fen, pH, conductivity, elevation, coordinates, and length of the temperature recording period.
+Groundwater pH and conductivity values are the average of 5–10 measures per site taken with a MM40 Portable Multimeter (Crison Instruments S.A., Alella, Spain).
 Coordinates are in decimal degrees WGS84.")
@@
 ```{r table3, echo = FALSE}
-regulartable(table3) %>% width(width = 1) %>% set_caption("Table 3: Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation.")
+regulartable(table3) %>% width(width = 1) %>% set_caption("Table 3: Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation.
+RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy).")
 ```
@@
 (b) Representation of the sampling design at El Riotuertu site.
-The position of the dry and waterlogged temperature logggers is indicated, the blue area represents the extent of the waterlogged mire habitat.", fig.align = "center", fig.width=7, fig.height=6}
-knitr::include_graphics("../Fig1.png")
+The position of the dry and waterlogged temperature logggers is indicated, the blue area represents the extent of the waterlogged mire habitat.", fig.width=7, fig.height=3.5}
+knitr::include_graphics("../results/Fig1.png")
 ```
@@
 The red series was recorded in a neighbouring dry area.
-Dataloggers were buried at 5 cm depth.", fig.align = "center", fig.width=7, fig.height=6}
-plot2
+Dataloggers were buried at 5 cm depth.
+The amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.", fig.width=7, fig.height=10}
+knitr::include_graphics("../results/Fig2.png")
 ```
@@
 ```{r fig3, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Average bioclimatic variables in the dry and waterlogged points.
-The bars represent the mean value, and the brackets the standard error of 8 records.", fig.align = "center", fig.width=7, fig.height=3}
-plot3
+The bars represent the mean value, and the brackets the standard error of 8 records.", fig.width=7, fig.height=3}
+knitr::include_graphics("../results/Fig3.png")
 ```
@@
 ```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Scatter plots of the bioclimatic variables predicted by CHELSA air temperatures vs.
-soil temperatures measured in situ, in dry and waterlogged points.", fig.align = "center", fig.width=7, fig.height=7}
-plot4
+soil temperatures measured in situ, in dry and waterlogged points.", fig.width=7, fig.height=7}
+knitr::include_graphics("../results/Fig4.png")
 ```
+
+```{r fig5, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Complementary temporal analysis of the groundwater buffer effect during peak hot events.
+Variation of soil temperatures at paired sites (dry vs.
+waterlogged points) during exceptionally warm days of the warm season.
+Exceptionally warm days were considered those with a daily maximum (at the dry point) belonging to the 90th percentile of daily maximums for the site.
+The 1:1 line (corresponding to no buffer effect) is shown, points falling to the right of this line show a buffer effect (i.e.
+soil temperatures at the waterlogged point are cooler than the neighnouring dry site).", fig.width=3.5, fig.height=3.5}
+knitr::include_graphics("../results/Fig5.png")
+```

### 2020-08-26 src/cleaning.R

@@ -1,4 +1,4 @@
-library(tidyverse); library(sp); library(lubridate)
+library(tidyverse); library(lubridate)
 
 # Function reads and formats GP5WSHELL csv export files
 
@@ -13,26 +13,6 @@ read_gp5 <- function(x)
            Logger = gsub("\\_.*", "" , x)) %>%
     dplyr::select(Logger, Time, Temperature)
 }
-
-# # Functions convert coordinates to decimal degrees
-# 
-# decdeg30T <- function(x) { x -> o1 
-#   o1 %>% dplyr::select(ED50X, ED50Y) -> o2
-#   SpatialPoints(o2, proj4string = CRS("+proj=utm +zone=30T +datum=WGS84")) -> o3
-#   spTransform(o3, CRS("+proj=longlat +datum=WGS84")) %>% data.frame -> o4
-#   colnames(o4) <- c("Longitude", "Latitude")
-#   o1 %>% dplyr::select(-c(ED50X, ED50Y)) -> o5
-#   cbind(o5, o4)
-# }
-# 
-# decdeg29T <- function(x) { x -> o1 
-#   o1 %>% dplyr::select(ED50X, ED50Y) -> o2
-#   SpatialPoints(o2, proj4string = CRS("+proj=utm +zone=29T +datum=WGS84")) -> o3
-#   spTransform(o3, CRS("+proj=longlat +datum=WGS84")) %>% data.frame -> o4
-#   colnames(o4) <- c("Longitude", "Latitude")
-#   o1 %>% dplyr::select(-c(ED50X, ED50Y)) -> o5
-#   cbind(o5, o4)
-# }
 
 # Read logger header data
 
@@ -49,27 +29,7 @@ header %>%
 header %>% select(-Elevation) %>% 
   merge(average.elevation) -> header # Only one elevation value per site
 
-# Map of the loggers
-
-header %>%
-  ggplot(aes(x = Longitude, y = Latitude)) + 
-  labs(x = "Longitude (º)", y = "Latitude (º)") +
-  geom_polygon(data = map_data("world"), aes(x = long, y = lat, group = group), 
-               color = "black", fill = "grey") + 
-  geom_point(aes(color = "indianred"), alpha = 0.5, size = 3, show.legend = FALSE) +
-  geom_text(aes(x = -3, y = 42, label = "Spain"), size = 4) +
-  geom_text(aes(x = 1, y = 44.8, label = "France"), size = 4) +
[…]

### 2020-08-26 doc/response-reviewers.Rmd
1st revision 🍋

---
title: "Ref. JOC-20-0599 - Response to reviewers"
author: "Eduardo Fernández-Pascual & Eva Correia-Álvarez"
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Editor

> *Both reviewers found this to be an interesting study. There is room to expand on your results/discussion as highlighted by both reviewers to improve upon the rigor and impact of this work. One upshot of the study is that mire regions might be climate refugia. Figure 3 gets at this point for summer Tmax spatially. I wonder if one can also show how soil temperatures may vary at paired sites during exceptionally warm daily temperatures during the warm season (e.g., complementary temporal analysis).*

We have expanded our discussion of the results as indicated. We have included a new figure (Figure 5) showing the complementary variation in paired sites during exceptionally warm days (i.e. days in the 90th percentile of daily maximums for the site). Generally, we have implemented all recommendations of the reviewers, except for the trend analysis suggested by reviewer 2. We do not have access to the data required to do such an analysis, since analysing the regional evolution of climate was not one of the objectives of our project.

# Reviewer: 1

> *This is a short and simple study comparing 5cm soil temperatures in mire habitats with adjacent drier microhabitats. It could be of interest to some readers, although it could use some polish and set in a broader context.  L15 – I think a yes/no type hypothesis about climate variability is not very interesting given we already have a priori expectations in this regard (e.g. L61-65, Ashcroft & Gollan 2013 Agric. For. Meteorol. 176:77-89. I would focus more on a quantitative question like: how different are they? What are you finding that is different from what we already know?*

We have changed the question as suggested to focus on the quantification of the effect, and modified the relevant parts of the manuscript to reflect this change.

> *L25 – What do you mean by failed to predict? E.g. r2 < 0.5? RMSE > 2oC? Spatial pattern wrong? Consistent bias?*

We have changed this sentence to comply with the new approach using correlation and RMSE (see below), and now indicate in the abstract what we mean regarding p-value, r and RMSE.

> *L42 – “has been”*

Changed.

> *L88-94: Like L15 above, we already have strong evidence that soil moisture acts as a buffer, so testing yes/no hypothesis about this make the research more trivial that it could be. I would focus more on how different they are etc.*

We have changed the question as suggested (see above).

> *L142 – significantly? P<0.05?*

We now indicate the p-value threshold whenever we use the word “significant”. In this case, we use the word “considerably” because we refer to a qualitative visual evaluation of the raw data shown in Figure 2, before going into the statistical analyses in the subsequent paragraphs.

> *L155 – This is more like a results section with actual figures. Would be good to see 95% CI for the effect sizes. i.e. instead of just effect size = -2.29oC, you would give a range like -1.5 to -3.0.*

We have included the CI. Please note that since this is the result of a one-tailed t-test, one of the limits of the CIs is -infinite or infinite.

> *Fig. 3 – Make the scales on the y and x axes the same and add a 1:1 line. R2 tells you the correlation between soil and air temps but not the bias. You could have a high r2 and still have a high RMSE indicating poor predictions.*

We have changed the figure as requested. For the RMSE comment, please see below.

> *L160-165 – building on above comment, note both the r2 and RMSE to capture both correlation and bias. These results will be affected by a number of factors: 1) the soil moisture effect you are interested in 2) the height of observation (2m air temperatures v 5cm soil temperatures) 3) the temporal scale of obvservation (the 1-5 years of your study is different to Chelsa) 4) The spatial scale – does Chelsa reflect the actual elevation of your sites.*

We have followed the suggestion and now use RMSE to measure the accuracy of the CHELSA values in relation to our measurements. To indicate the correlation between CHELSA and our measurements, instead of the linear model of the previous version, we now use Pearson’s correlation. We have changed the text accordingly. We also note these considerations regarding the factors affecting the CHELSA/soil measure correspondence in the appropriate place of the discussion.

> *L167 – “Quantify the size of the thermal buffer effect” rather than “prove the existence”. We already know it exists.*

Changed.

> *Discussion: Where results are similar or different from other studies it would be nice to delve deeper into why. Were the observations at the same depth? E.g. We expect diurnal range to decrease with depth so studies at different depths are not necessarily comparable. What do the results tell us more generally about mires elsewhere, or where sites may vary along a moisture gradient rather than just be mire or non-mire?*

We have expanded the discussion as suggested, indicating comparisons about the depth of measuring, and making a general interpretation about mires elsewhere. We also comment the relevance of this effect in all environments with a moisture gradient, such as alpine micro-topographies.

> *198-208 – Don’t read too much into individual sites. When you only have 8 sites each will be unique in one way or another and we expect some results to be different. What we are interested in is trends that are consistent across all 8 sites.*

We agree with the reviewer on this point, we believe it is worthy to highlight some of the extreme cases among our sites for the purpose of discussion, but no firm conclusions can be made with 8 sites. We now state this limitation in the discussion.

# Reviewer: 2

> *Very interesting datasets. More rigor in describing how analyses were performed and digging a little deeper into the results is needed. Peatland terms vary somewhat from N. America to Europe / Eurasia. Clearly defining hydrological terms like "mire" will be useful - is this a broad term for all peatlands or simply groundwater fed ones?  You describe a groundwater buffer effect, but include wetlands in your study whose hydrology may be dominated moreso by precipitation inputs (bogs, raised bogs). While it may not be practical to fully describe source hydrology to your sites from in situ measurements, some more information on hydrology of these wetlands would be useful. Does a gradient in groundwater connection strength exist among your sites?*

We have expanded the description of mire, including the definition of bogs and fens. We also describe better the classification we use (based on EUNIS, we provide links to the EUNIS sheets for our habitat types) and give more details about the hydrology of our sites. Please note that our two bogs are classified as raised bogs because of their vegetation, but they are very poorly developed, with small isolated hummocks. Therefore, they are greatly influenced by soil water coming from streams (they develop in waterlogged valleys) and cannot be considered to be fed only by precipitation.

> *A general comment is, you would expect to see a thermal buffer effect just from the presence of water at all compared to dry sites. Your data indicate that none of the waterlogged sample locations dried out, thus removing the thermal buffer effect during mreasurement (Figure 1). Do you have a hypothesis for what would happen if they dried based on your data? For your low elevation sites, it suggests any locations drying might be really susceptible to warming.*

We have expanded on this comment in the discussion, indicating the relation of the buffer effect with the maintenance of the precipitation regimes, and commenting the precipitation projections for the study region.

> *Is it possible to compare/contrast sites based on sites with weaker vs. stronger groundwater inputs in your study sites? Yes there is an elevation gradient present, but do differences in source hydrology, stronger groundwater inputs explain the really pronounced buffering of some sites vs. others irrespective of elevation?*

We agree with reviewer 1 in that, since we have eight sites, it is not possible to read too much on their individual character or perform any type of statistical test considering their elevation or their source hydrology. This is something that can only be pointed out in the discussion. See also the previous comment, our two bog sites are very poorly developed as bogs and receive water inputs from streams.

> *Your methods section is really short and does not provide extensive detail on how you performed your analysis, what statistical platform you used, elsewhere R is referenced, but not in the methods directly.*

We have expanded the section to indicate the use of R and the relevant packages. When the article is accepted, the R code and the raw data will be made publicly available through GitHub. For the purpose of peer-review, we provide a link to a zip version of the GitHub repository (online submission systems generally do not allow to upload zip files). Please see the methods and the data availability section for more details.

> *Your study might benefit from using a mixed model instead of a simple linear model. I could be convinced a more sophisticated analytical structure like linear mixed modeling (site random effect) is not needed - especially if you expect differences related to hydrology among sites, but those differences if they exist you may want to provide some justification for. If you think your sites have similar hydro-logical characteristics, then a site random effect grouping wet and dry locations by site may make sense to control for noise in the dataset from microclimate factors such as aspect etc. Please at include some justification as to why a mixed modeling approach is not needed or replace the simple linear model with a mixed model.*

Following the comments of the previous reviewer, we have removed the linear model, and now use Pearson’s correlation. The purpose of the text is simply to check the correlation of soil measures with CHELSA values, a mixed model with random factors is not justified.

> *It looks as though some of the temperature variability in your dataset is due to differences in hydrology among sites. If you suspect that is the case as well, then you can pretty easily justify the existing linear model analytical structure, but need more interpretation of hydrological differences in your results and discussion.*

As commented above, we do not believe we can study in more detail the effect of source hydrology because of the limited number of sites, and also because all of our sites receive some groundwater input either from springs or streams.

> *Do any patterns emerge if you code sites by habitat type?  Please include the habitat type labels presented in Table 1 in the figure panels for Figure 1.*

We have included the labels as suggested.

> *Can you clarify if dry sites are in fact adjacent dry uplands or dry raised hummocks (high spots) within the wetlands? A figure with a site-level inset map would also convey this information and would benefit the paper.*

We have clarified that they are adjacent dry uplands, and also included the requested figure.

> *It would be worth describing the >15 degree C differences between wet vs dry sites at low elevation (La Malva). That is very biologically significant! You should put that into perspective for the readers, that level of difference greatly exceeds IPCC projections for climate change in this region... Huge differences within site. Please provide some descriptive statistics on peak differences observed at some of your sites. This is a neat study, there seems to be some more interesting angles you can explore/present in your discussion that will make your paper more interesting to a wider array of readers.*

We have noted this comment about the comparison between the buffer at La Malva and the warming predictions, and also expanded the description of the special case of La Malva. We have also included a new figure (Figure 5) showing the peak differences in the eight sites during exceptionally warm days.

## Additional comments in PDF

> *The authors have collected compelling data in an understudied field. The study is relevant to biodiversity under climate change, and to assumptions of global carbon dynamics. The data collected has a lot of merit and pretty easily stands on its own without much help. However, additional descriptions of methodologies are needed. There are some interesting results here, and a nice temporal record from which to draw conclusions. The authors focus on the waterlogged nature of the soils – but the source of hydrology and how closely coupled groundwater in mires is linked with atmospheric conditions may have some bearing on whether these systems remain buffered from effects of atmospheric warming. More robust discussion on the atmospheric coupling of hydrology in these systems tied to other literature is needed. Do groundwater-fed mires persist in drying climates? The paper may benefit from discussion on changes in climate in the study region. Is this study region getting drier? What effect is that expected to have on the thermal buffer effect described? Some descriptions of the latitudinal and elevational distribution of mires in Europe from literature as well as precipitation range would be useful.  One to two paragraphs with appropriate citations will help. There are a handful of N. American papers that discuss geographic distribution of groundwater-fed mires [fens] as well. Some literature on how extensive these peatlands are in temperate and boreal landscapes would be helpful to convey the relevance of this work*.

We have expanded the sentences describing the distribution of mires in Europe, North America and Russia. For the questions related to hydrology and climate change in the study region, please see the responses to similar comments.

> *Some more description on differences in source hydrology to the sites studied and implications for how that may translate to strength of thermal buffer is warranted. Strictly speaking source hydrology – dominated by groundwater flows vs. hydrology dominated moreso by precipitation should have an effect on the microclimate thermal buffer. Some of the variation among sites that is unexplained by elevation may be a result of stronger vs. weaker connection to groundwater. This is an angle you really don’t address in much detail and is a potentially really important area to explore in discussion. It’s possible that wetlands in similar elevation and latitude could have markedly differing responses to climate change. i.e., stronger buffer effects at certain sites vs. others at the similar elevations – explore and describe these results further.*

Please see the previous response to the issue of hydrology.

> *Including a table with shift in climate trend per decade for precipitation and temperature for the study area would be informative or a citation with similar information.*

We have cited the data with the latest projections made by the national meteorology agency of Spain (AEMET).

> *Some additional historical background on groundwater thermal buffer effect would improve the paper, and further support the notion that these effects have been “known” but little quantified. The authors will probably enjoy reading Clara May Frederick’s, 1974 A Natural History Study of the Vascular Flora of Cedar Bog, Champaign County, Ohio) Ohio Journal of Science who performed early temperature data collection work in such settings as far back as the 1960’s.*

Thank you for this pioneering reference, we have integrated it into our manuscript.

> *A Ph.D. dissertation chapter by Raney (2014) Identifying Potential Refugia from Climate Change in Wetlands: High Resolution Soil Climate Maps For Priority Conservation Areas: Hydrologic Processes Mediate Significant Microclimate Variation uses very similar field data collection  methods to test the groundwater thermal buffer effect in wetlands as the authors, but focused on spatial coverage vs temporal length. That study includes 1-year worth of data. While the Raney study has not undergone formal peer-review and has some limitations, the soil temperature figures and data summaries in that study may be of particular interest to the authors as additional evidence of thermal buffer due to groundwater.*

We have integrated this reference into our manuscript.

> *Line 39 – increasingly there are downscaling efforts based on atmospheric microclimate networks distributed in complex terrain that focus on topographic effects on air temperature. So to a degree microclimate factors are increasingly considered, but not hydrologic factors under edaphic/soil control. It may be a distinction worth pointing out.*

We have included this consideration.

> *Lines 47-59 The term mire may vary somewhat in terms of meaning by region. Please clearly define the primary source of hydrology along with your definition of mire. This will help readers in other regions. If the authors have soil-pH or pore-water pH available for study sites that would be useful but not mandatory to include in Table 1.*

We have included water pH and conductivity in the table. Please see also the response to a similar comment above.

> *Line 56 – insert “across their range” after homogeneous flora.*

Changed.

> *Line 76 – and carbon*

Changed.

> *Line 78 – replace the word “surface” with “area”*

Changed.

> *Figures: Your first figure should be a site map showing regional context, and if possible topographic gradient of your sites.*

We have created the suggested figure.

> *Figure 1 Make all labels bold, and text larger. To reduce space, put Site name labels in the light gray plot space and reduce the resulting blank space between top and bottom panels. This figure will likely take up ½ the size when published, so labels need to be larger/darker to maintain legibility.  Also you might add to your figure caption that the amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.*

Changed as suggested.

> *Any ideas why La Recoleta at higher elevation shows a really strong disparity between wet and dry locations moreso than your other high elevation sites? Do you think aspect or some other factor had an effect? May be worth exploring in your results section a little further.*

A possible explanation is that La Recoleta is a quaking bog (a transition from aquatic to fen vegetation at the margins of a silting glacial lake) and therefore it is more waterlogged than other sites. We have expanded this explanation. However, we agree with reviewer 1 that we should not go too far into interpreting the results of single sites.

> *Line 80 – Groundwater flows are somewhat decoupled from high frequency variation in precipitation, but are still linked to lower-frequency long-term changes in precipitation, groundwater recharge and flow to some degree. Focusing some attention on this phenomenon and possibly your discussion may be useful. Just because there is an observed thermal buffer effect now does not mean it will persist indefinitely, especially if climate becomes more arid – distribution of groundwater flows supporting mires may diminish in frequency, quantity, resulting in periodic drying and even potential loss of this buffer effect. For instance, regions that are wetting and expected to become wetter based on IPCC or other projections may have more stability in the thermal buffer effect vs. drying regions. It would be beneficial to focus some of your discussion on this topic.*

We have expanded the discussion in this direction, noting that the buffer effect is dependent on future changes on precipitation, and describing the latest projection for the study area (a decrease of 10 % by the end of the century).

> *Line 86 change representativeness to representation for simplicity.*

Changed.

> *The materials and methods section needs to be much more robust in terms of explaining the statistical platform used to perform the analyses, the packages used, citations for those platforms and packages. Why the analyses are appropriate. There’s really not much of a description of how linear modeling was performed, assumptions etc. A linear mixed-model approach would make a lot of sense instead of a simple linear model, grouping wet and dry locations by a “Site” factor to control for latent processes.*

Please see a previous response to a similar comment.

> *Need details on sample sizes in your methods.*

We have included the sample sizes.

> *Lines 99-101 Also include quantitative descriptors of climate in the study area (mm precip/year min and max temp/year).*

We have included this information.

> *Line 110 – provide sampling resolution of loggers. A description of why was 5cm depth chosen is needed.*

We have included this information, and the explanation for the sampling depth.

> *Please include a map of the study sites with an inset to show regional/continental position.*

We have included the map.

> *Lines 123 – 133 – some discussion on why these are biologically or environmentally meaningful statistics is needed. Are species distributions controlled by winter minimums? Carbon dynamics by warm temperatures?*

We have included a description of the biological meaning of the bioclimatic variables we chose, with a relevant reference for more information.

> *Statistical analysis: Some simple statistics on precipitation and temperature trends over the last ~hundred years in this study region would be helpful to broaden discussion regarding future uncertainty of microclimates. That would also help to put your study into perspective / macroclimatic context. Are the elevational trends we’re seeing shown for a warming region?  A simple trend analysis would be useful. Mm precipitation shift/decade and °C/decade would be useful metrics to include as well as the significance of those shifts. It is a stronger argument if you can show that there’s a thermal buffer effect detectable in a warming or drying climate. Dendrochronology literature frequently includes analytical methods for such trend analyses. *

We are sorry but we do not have access to the data necessary to make such analysis, since analysing the evolution of climate in the study region was not the goal of this project.

> *Results: More interpretation of the results is needed. Line 155 If you’re only looking at wet vs dry sites, wouldn’t you expect wet sites to have less variability even if the source was not groundwater?*

We interpret the results in the discussion, and comment the generality of the moisture effect.

> *Line 167 – Delete the word prove. The evidence required to prove something is very high. Suggest changing to: The results presented provide strong evidence of...*

Changed.

> *Line 173 – Raney 2014 has similar findings from year-round data for a single year. Warmer in winter, cooler in summer. You should highlight that your findings were extended to a full year and the pattern is reproducible among years out to five years. The fact that your data possess a strong-elevational relationship, i.e., stronger under warming environment is worth more attention.*

We have added these considerations.

> *Line 194 – Replace “highness” with “height”*

Changed.

> *Line 200 – you elude to some topographic microclimate effects on your sites. That is fine as they influence both wet and dry locations equally, but may be worth including aspect measurements for your sites in Table 1. Neat datasets, I look forward to reading the final manuscript.*

Only one of the sites (La Malva) is in a slope and thus has an aspect, the other sites are flat. We have changed the sentence to make this clear.

### 2020-08-29 doc/response-reviewers.Rmd
files submitted to revision

@@
 When the article is accepted, the R code and the raw data will be made publicly available through GitHub.
-For the purpose of peer-review, we provide a link to a zip version of the GitHub repository (online submission systems generally do not allow to upload zip files).
-Please see the methods and the data availability section for more details.
+For the purpose of peer-review, we provide a link to a zip version of the GitHub repository (online submission systems generally do not allow to upload zip files): https://drive.google.com/file/d/1wnrNuQXvXcCruhKXdoFaRcm6tEZPaZkN/view?usp=sharing)
 

### 2020-08-29 doc/response-reviewers.Rmd
Revert "1st revision 🍋"
This reverts commit cf4453643d0bd19f444758d41499e7b905a136d8.

--- 2020-08-26 doc/response-reviewers.Rmd
@@
 title: "Ref. JOC-20-0599 - Response to reviewers"
-author: "Eduardo Fernández-Pascual & Eva Correia-Álvarez"
-date: "`r format(Sys.time(), '%d/%m/%Y')`"
-output: pdf_document
+author: "Eduardo Fernández Pascual & Eva Correia-Álvarez"
+date: "`r format(Sys.time(), '%d %B, %Y')`"
+output: html_document
 ---
@@
 
-# Editor
+# Response to the Editor
 
-> *Both reviewers found this to be an interesting study.
-There is room to expand on your results/discussion as highlighted by both reviewers to improve upon the rigor and impact of this work.
+# Response to Referee 1
+
+# Response to Referee 2
+
+Dear Dr Fernández-Pascual
+
+We recognise that the impact of the COVID-19 pandemic may affect your ability to return your revised manuscript to us within the requested timeframe. If this is the case, please let us know.
+
+Manuscript # JOC-20-0599 entitled "Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils" which you submitted to the International Journal of Climatology, has been reviewed.
+The comments of the referee(s), all of whom are leading international experts in this field, are included with this letter.
+
+The reviewers believe that the paper has potential to be published in the International Journal of Climatology, but major revisions are required.
+Revisions in this category may involve substantial text changes, recalculations or new analyses in addition to more minor clarifications and corrections.
+In your revisions, please address all of the points made by each reviewer, amend the manuscript accordingly, and include a point-by-point response to each reviewer comment with your response.
+
+Both reviewers found this to be an interesting study. There is room to expand on your results/discussion as highlighted by both reviewers to improve upon the rigor and impact of this work.
+
 One upshot of the study is that mire regions might be climate refugia.
 Figure 3 gets at this point for summer Tmax spatially.
-I wonder if one can also show how soil temperatures may vary at paired sites during exceptionally warm daily temperatures during the warm season (e.g., complementary temporal analysis).*
+I wonder if one can also show how soil temperatures may vary at paired sites during exceptionally warm daily temperatures during the warm season (e.g., complementary temporal analysis).
 
-We have expanded our discussion of the results as indicated.
-We have included a new figure (Figure 5) showing the complementary variation in paired sites during exceptionally warm days (i.e.
-days in the 90th percentile of daily maximums for the site).
-Generally, we have implemented all recommendations of the reviewers, except for the trend analysis suggested by reviewer 2.
-We do not have access to the data required to do such an analysis, since analysing the regional evolution of climate was not one of the objectives of our project.
+Because the manuscript needs major revisions, I will send the revised manuscript and your responses back to some or all of the reviewers for further review.
+Please note that submitting a revision of your manuscript does not guarantee eventual acceptance.
+I would like to receive the revised manuscript as soon as is convenient, but by 24-Nov-2020 at the latest.
 
-# Reviewer: 1
+This should be sufficient time given the nature of the revisions, but please communicate with me via email if you need an extension of the deadline.
 
-> *This is a short and simple study comparing 5cm soil temperatures in mire habitats with adjacent drier microhabitats.
+If the reviewers have suggested that the language of the current version of your manuscript does not allow a thorough evaluation, we recommend that you have your paper professionally edited for English language by a service such as Wiley’s at http://wileyeditingservices.com.
+Please note that while this service will greatly improve the readability of your paper, it does not guarantee acceptance of your paper by the journal.
+
+Once again, thank you for submitting your manuscript to the International Journal of Climatology and I look forward to  receiving your revision.
+
+Sincerely,
+
+Dr John Abatzoglou
+Associate Editor, International Journal of Climatology
+jabatzoglou@ucmerced.edu
+
+***************************************************************************
+Instructions for submitting your revised manuscript:
+
+To submit your revised manuscript, log on to http://mc.manuscriptcentral.com/joc and click on ‘Author Centre’.
+Click the link in the ‘Manuscripts with Decisions’ list.
+Click on the ‘create a revision’ link for the appropriate manuscript.
+
+Please note:
+-Once you begin to submit the revision, the 'create a revision' link disappears and the manuscript moves into your Revised Manuscripts in Drafts list.
+The option reappears if you delete the revision before it is actually submitted.
+
+Should a revised manuscript be accepted, the files will be sent electronically to the Production Department.
+It is thus very important that the files uploaded at revision are in their final format and of sufficient quality for publication.
+At revision, you should provide:
+a) The manuscript text in Word format (all tracked changes removed).
+This file should be text only, should not contain figures and must not exceed 1MB in size.
+Tables must not be embedded, but provided in editable format.
+b) Figures as high-resolution TIFF, EPS or PDF files; colour in CMYK format.
+c) Supporting Information files as required. Please note that the maximum file size for each Supporting Information file is 20 MB.
+d) A copy of the text file with changes highlighted.
+e) A letter of response to the reviewers' comments, detailing point-by-point the changes made and giving reasons for those not made.
+
+Revised manuscripts must be in their final form when submitted.
+The proofs received later are for correction of typographical errors only.
+They should not be used for final changes to articles; such changes must be made to the manuscript before it goes to Production
+
+***************************************************************************
+
+Due Date of Revised Submission: 24-Nov-2020
+
+***************************************************************************
+
+Referee(s)' Comments to Author:
+
+Reviewer: 1
+
+Comments to the Author
+This is a short and simple study comparing 5cm soil temperatures in mire habitats with adjacent drier microhabitats.
 It could be of interest to some readers, although it could use some polish and set in a broader context.
+
+Specific comments:
+
 L15 – I think a yes/no type hypothesis about climate variability is not very interesting given we already have a priori expectations in this regard (e.g.
@@
 I would focus more on a quantitative question like: how different are they?
-What are you finding that is different from what we already know?*
+What are you finding that is different from what we already know?
 
-We have changed the question as suggested to focus on the quantification of the effect, and modified the relevant parts of the manuscript to reflect this change.
+L25 – What do you mean by failed to predict? E.g. r2 < 0.5? RMSE > 2oC? Spatial pattern wrong? Consistent bias?
 
-> *L25 – What do you mean by failed to predict? E.g. r2 < 0.5? RMSE > 2oC? Spatial pattern wrong? Consistent bias?*
+L42 – “has been”
 
-We have changed this sentence to comply with the new approach using correlation and RMSE (see below), and now indicate in the abstract what we mean regarding p-value, r and RMSE.
+L88-94: Like L15 above, we already have strong evidence that soil moisture acts as a buffer, so testing yes/no hypothesis about this make the research more trivial that it could be.
+I would focus more on how different they are etc.
 
-> *L42 – “has been”*
+L142 – significantly? P<0.05?
 
-Changed.
+L155 – This is more like a results section with actual figures. Would be good to see 95% CI for the effect sizes. i.e. instead of just effect size = -2.29oC, you would give a range like -1.5 to -3.0.
 
-> *L88-94: Like L15 above, we already have strong evidence that soil moisture acts as a buffer, so testing yes/no hypothesis about this make the research more trivial that it could be.
-I would focus more on how different they are etc.*
-
-We have changed the question as suggested (see above).
-
-> *L142 – significantly? P<0.05?*
-
-We now indicate the p-value threshold whenever we use the word “significant”.
-In this case, we use the word “considerably” because we refer to a qualitative visual evaluation of the raw data shown in Figure 2, before going into the statistical analyses in the subsequent paragraphs.
-
-> *L155 – This is more like a results section with actual figures.
-Would be good to see 95% CI for the effect sizes.
-i.e.
-instead of just effect size = -2.29oC, you would give a range like -1.5 to -3.0.*
-
-We have included the CI. Please note that since this is the result of a one-tailed t-test, one of the limits of the CIs is -infinite or infinite.
-
-> *Fig.
+Fig.
 3 – Make the scales on the y and x axes the same and add a 1:1 line.
 R2 tells you the correlation between soil and air temps but not the bias.
-You could have a high r2 and still have a high RMSE indicating poor predictions.*
+You could have a high r2 and still have a high RMSE indicating poor predictions.
 
-We have changed the figure as requested. For the RMSE comment, please see below.
+L160-165 – building on above comment, note both the r2 and RMSE to capture both correlation and bias.
+These results will be affected by a number of factors: 1) the soil moisture effect you are interested in 2) the height of observation (2m air temperatures v 5cm soil temperatures) 3) the temporal scale of obvservation (the 1-5 years of your study is different to Chelsa) 4) The spatial scale – does Chelsa reflect the actual elevation of your sites.
 
-> *L160-165 – building on above comment, note both the r2 and RMSE to capture both correlation and bias.
-These results will be affected by a number of factors: 1) the soil moisture effect you are interested in 2) the height of observation (2m air temperatures v 5cm soil temperatures) 3) the temporal scale of obvservation (the 1-5 years of your study is different to Chelsa) 4) The spatial scale – does Chelsa reflect the actual elevation of your sites.*
+L167 – “Quantify the size of the thermal buffer effect” rather than “prove the existence”. We already know it exists.
 
-We have followed the suggestion and now use RMSE to measure the accuracy of the CHELSA values in relation to our measurements.
-To indicate the correlation between CHELSA and our measurements, instead of the linear model of the previous version, we now use Pearson’s correlation.
-We have changed the text accordingly.
-We also note these considerations regarding the factors affecting the CHELSA/soil measure correspondence in the appropriate place of the discussion.
-
-> *L167 – “Quantify the size of the thermal buffer effect” rather than “prove the existence”. We already know it exists.*
-
-Changed.
-
-> *Discussion: Where results are similar or different from other studies it would be nice to delve deeper into why.
+Discussion: Where results are similar or different from other studies it would be nice to delve deeper into why.
 Were the observations at the same depth?
@@
 We expect diurnal range to decrease with depth so studies at different depths are not necessarily comparable.
-What do the results tell us more generally about mires elsewhere, or where sites may vary along a moisture gradient rather than just be mire or non-mire?*
+What do the results tell us more generally about mires elsewhere, or where sites may vary along a moisture gradient rather than just be mire or non-mire?
 
-We have expanded the discussion as suggested, indicating comparisons about the depth of measuring, and making a general interpretation about mires elsewhere.
-We also comment the relevance of this effect in all environments with a moisture gradient, such as alpine micro-topographies.
+198-208 – Don’t read too much into individual sites.
+When you only have 8 sites each will be unique in one way or another and we expect some results to be different.
+What we are interested in is trends that are consistent across all 8 sites.
 
-> *198-208 – Don’t read too much into individual sites.
-When you only have 8 sites each will be unique in one way or another and we expect some results to be different.
-What we are interested in is trends that are consistent across all 8 sites.*
+Reviewer: 2
 
-We agree with the reviewer on this point, we believe it is worthy to highlight some of the extreme cases among our sites for the purpose of discussion, but no firm conclusions can be made with 8 sites.
-We now state this limitation in the discussion.
-
-# Reviewer: 2
-
-> *Very interesting datasets.
+Comments to the Author
+Very interesting datasets.
 More rigor in describing how analyses were performed and digging a little deeper into the results is needed.
@@
 Clearly defining hydrological terms like "mire" will be useful - is this a broad term for all peatlands or simply groundwater fed ones?
-You describe a groundwater buffer effect, but include wetlands in your study whose hydrology may be dominated moreso by precipitation inputs (bogs, raised bogs).
+You describe a groundwater buffer effect, but include wetlands in your study whose hydrology may be dominated moreso by precipitation inputs (bogs, raised bogs) .
 While it may not be practical to fully describe source hydrology to your sites from in situ measurements, some more information on hydrology of these wetlands would be useful.
-Does a gradient in groundwater connection strength exist among your sites?*
+Does a gradient in groundwater connection strength exist among your sites?
 
-We have expanded the description of mire, including the definition of bogs and fens.
-We also describe better the classification we use (based on EUNIS, we provide links to the EUNIS sheets for our habitat types) and give more details about the hydrology of our sites.
-Please note that our two bogs are classified as raised bogs because of their vegetation, but they are very poorly developed, with small isolated hummocks.
-Therefore, they are greatly influenced by soil water coming from streams (they develop in waterlogged valleys) and cannot be considered to be fed only by precipitation.
-
-> *A general comment is, you would expect to see a thermal buffer effect just from the presence of water at all compared to dry sites.
+A general comment is, you would expect to see a thermal buffer effect just from the presence of water at all compared to dry sites.
 Your data indicate that none of the waterlogged sample locations dried out, thus removing the thermal buffer effect during mreasurement (Figure 1).
 Do you have a hypothesis for what would happen if they dried based on your data?
-For your low elevation sites, it suggests any locations drying might be really susceptible to warming.*
+For your low elevation sites, it suggests any locations drying might be really susceptible to warming.
 
-We have expanded on this comment in the discussion, indicating the relation of the buffer effect with the maintenance of the precipitation regimes, and commenting the precipitation projections for the study region.
-
-> *Is it possible to compare/contrast sites based on sites with weaker vs.
+Is it possible to compare/contrast sites based on sites with weaker vs.
 stronger groundwater inputs in your study sites?
 Yes there is an elevation gradient present, but do differences in source hydrology, stronger groundwater inputs explain the really pronounced buffering of some sites vs.
-others irrespective of elevation?*
+others irrespective of elevation?
+Your methods section is really short and does not provide extensive detail on how you performed your analysis, what statistical platform you used, elsewhere R is referenced, but not in the methods directly.
 
-We agree with reviewer 1 in that, since we have eight sites, it is not possible to read too much on their individual character or perform any type of statistical test considering their elevation or their source hydrology.
-This is something that can only be pointed out in the discussion.
-See also the previous comment, our two bog sites are very poorly developed as bogs and receive water inputs from streams.
-
-> *Your methods section is really short and does not provide extensive detail on how you performed your analysis, what statistical platform you used, elsewhere R is referenced, but not in the methods directly.*
-
-We have expanded the section to indicate the use of R and the relevant packages.
-When the article is accepted, the R code and the raw data will be made publicly available through GitHub.
-For the purpose of peer-review, we provide a link to a zip version of the GitHub repository (online submission systems generally do not allow to upload zip files).
-Please see the methods and the data availability section for more details.
-
-> *Your study might benefit from using a mixed model instead of a simple linear model.
+Your study might benefit from using a mixed model instead of a simple linear model.
 I could be convinced a more sophisticated analytical structure like linear mixed modeling (site random effect) is not needed - especially if you expect differences related to hydrology among sites, but those differences if they exist you may want to provide some justification for.
 If you think your sites have similar hydro-logical characteristics, then a site random effect grouping wet and dry locations by site may make sense to control for noise in the dataset from microclimate factors such as aspect etc.
-Please at include some justification as to why a mixed modeling approach is not needed or replace the simple linear model with a mixed model.*
+Please at include some justification as to why a mixed modeling approach is not needed or replace the simple linear model with a mixed model.
 
-Following the comments of the previous reviewer, we have removed the linear model, and now use Pearson’s correlation.
-The purpose of the text is simply to check the correlation of soil measures with CHELSA values, a mixed model with random factors is not justified.
+It looks as though some of the temperature variability in your dataset is due to differences in hydrology among sites.
+If you suspect that is the case as well, then you can pretty easily justify the existing linear model analytical structure, but need more interpretation of hydrological differences in your results and discussion.
 
-> *It looks as though some of the temperature variability in your dataset is due to differences in hydrology among sites.
-If you suspect that is the case as well, then you can pretty easily justify the existing linear model analytical structure, but need more interpretation of hydrological differences in your results and discussion.*
+Do any patterns emerge if you code sites by habitat type?  Please include the habitat type labels presented in Table 1 in the figure panels for Figure 1.
 
-As commented above, we do not believe we can study in more detail the effect of source hydrology because of the limited number of sites, and also because all of our sites receive some groundwater input either from springs or streams.
+Can you clarify if dry sites are in fact adjacent dry uplands or dry raised hummocks (high spots) within the wetlands?
+A figure with a site-level inset map would also convey this information and would benefit the paper.
 
-> *Do any patterns emerge if you code sites by habitat type?  Please include the habitat type labels presented in Table 1 in the figure panels for Figure 1.*
-
-We have included the labels as suggested.
-
-> *Can you clarify if dry sites are in fact adjacent dry uplands or dry raised hummocks (high spots) within the wetlands?
-A figure with a site-level inset map would also convey this information and would benefit the paper.*
-
-We have clarified that they are adjacent dry uplands, and also included the requested figure.
-
-> *It would be worth describing the >15 degree C differences between wet vs dry sites at low elevation (La Malva).
+It would be worth describing the >15 degree C differences between wet vs dry sites at low elevation (La Malva).
 That is very biologically significant!
@@
 Please provide some descriptive statistics on peak differences observed at some of your sites.
-This is a neat study, there seems to be some more interesting angles you can explore/present in your discussion that will make your paper more interesting to a wider array of readers.*
 
-We have noted this comment about the comparison between the buffer at La Malva and the warming predictions, and also expanded the description of the special case of La Malva.
-We have also included a new figure (Figure 5) showing the peak differences in the eight sites during exceptionally warm days.
+This is a neat study, there seems to be some more interesting angles you can explore/present in your discussion that will make your paper more interesting to a wider array of readers.
 
-## Additional comments in PDF
+# PDF
 
-> *The authors have collected compelling data in an understudied field.
+The authors have collected compelling data in an understudied field.
 The study is relevant to biodiversity under climate change, and to assumptions of global carbon dynamics.
@@
 American papers that discuss geographic distribution of groundwater-fed mires [fens] as well.
-Some literature on how extensive these peatlands are in temperate and boreal landscapes would be helpful to convey the relevance of this work*.
-
-We have expanded the sentences describing the distribution of mires in Europe, North America and Russia.
-For the questions related to hydrology and climate change in the study region, please see the responses to similar comments.
-
-> *Some more description on differences in source hydrology to the sites studied and implications for how that may translate to strength of thermal buffer is warranted.
+Some literature on how extensive these peatlands are in temperate and boreal landscapes would be helpful to convey the relevance of this work.
+Some more description on differences in source hydrology to the sites studied and implications for how that may translate to strength of thermal buffer is warranted.
 Strictly speaking source hydrology – dominated by groundwater flows vs.
@@
 i.e., stronger buffer effects at certain sites vs.
-others at the similar elevations – explore and describe these results further.*
-
-Please see the previous response to the issue of hydrology.
-
-> *Including a table with shift in climate trend per decade for precipitation and temperature for the study area would be informative or a citation with similar information.*
-
-We have cited the data with the latest projections made by the national meteorology agency of Spain (AEMET).
-
-> *Some additional historical background on groundwater thermal buffer effect would improve the paper, and further support the notion that these effects have been “known” but little quantified.
-The authors will probably enjoy reading Clara May Frederick’s, 1974 A Natural History Study of the Vascular Flora of Cedar Bog, Champaign County, Ohio) Ohio Journal of Science who performed early temperature data collection work in such settings as far back as the 1960’s.*
-
-Thank you for this pioneering reference, we have integrated it into our manuscript.
-
-> *A Ph.D.
+others at the similar elevations – explore and describe these results further.Including a table with shift in climate trend per decade for precipitation and temperature for the study area would be informative or a citation with similar information.
+Some additional historical background on groundwater thermal buffer effect would improve the paper, and further support the notion that these effects have been “known” but little quantified.
+The authors will probably enjoy reading Clara May Frederick’s, 1974 A Natural History Study of the Vascular Flora of Cedar Bog, Champaign County, Ohio) Ohio Journal of Science who performed early temperature data collection work in such settings as far back as the 1960’s.A Ph.D.
 dissertation chapter by Raney (2014) Identifying Potential Refugia from Climate Change in Wetlands: High Resolution Soil Climate Maps For Priority Conservation Areas: Hydrologic Processes Mediate Significant Microclimate Variation uses very similar field data collection  methods to test the groundwater thermal buffer effect in wetlands as the authors, but focused on spatial coverage vs temporal length.
 That study includes 1-year worth of data.
-While the Raney study has not undergone formal peer-review and has some limitations, the soil temperature figures and data summaries in that study may be of particular interest to the authors as additional evidence of thermal buffer due to groundwater.*
-
-We have integrated this reference into our manuscript.
-
-> *Line 39 – increasingly there are downscaling efforts based on atmospheric microclimate networks distributed in complex terrain that focus on topographic effects on air temperature.
+While the Raney study has not undergone
+formal peer-review and has some limitations, the soil temperature figures and data summaries in that study may be of particular interest to the authors as additional evidence of thermal buffer due to groundwater.
+Line 39 – increasingly there are downscaling efforts based on atmospheric microclimate networks distributed in complex terrain that focus on topographic effects on air temperature.
 So to a degree microclimate factors are increasingly considered, but not hydrologic factors under edaphic/soil control.
-It may be a distinction worth pointing out.*
-
-We have included this consideration.
-
-> *Lines 47-59 The term mire may vary somewhat in terms of meaning by region.
+It may be a distinction worth pointing out.Lines 47-59 The term mire may vary somewhat in terms of meaning by region.
 Please clearly define the primary source of hydrology along with your definition of mire.
 This will help readers in other regions.
-If the authors have soil-pH or pore-water pH available for study sites that would be useful but not mandatory to include in Table 1.*
-
-We have included water pH and conductivity in the table. Please see also the response to a similar comment above.
-
-> *Line 56 – insert “across their range” after homogeneous flora.*
-
-Changed.
-
-> *Line 76 – and carbon*
-
-Changed.
-
-> *Line 78 – replace the word “surface” with “area”*
-
-Changed.
-
-> *Figures: Your first figure should be a site map showing regional context, and if possible topographic gradient of your sites.*
-
-We have created the suggested figure.
-
-> *Figure 1 Make all labels bold, and text larger.
+If the authors have soil-pH or pore-water pH available for study sites that would be useful but not mandatory to include in Table 1.Line 56 – insert “across their range” after homogeneous floraLine 76 – and carbon.Line 78 – replace the word “surface” with “area”Figures: Your first figure should be a site map showing regional context, and if possible topographic gradient of your sites.
+Figure 1 Make all labels bold, and text larger.
 To reduce space, put Site name labels in the light gray plot space and reduce the resulting blank space between top and bottom panels.
 This figure will likely take up ½ the size when published, so labels need to be larger/darker to maintain legibility.
-Also you might add to your figure caption that the amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.*
-
-Changed as suggested.
-
-> *Any ideas why La Recoleta at higher elevation shows a really strong disparity between wet and dry locations moreso than your other high elevation sites?
+Also you might add to your figure caption that the amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.
+Any ideas why La Recoleta at higher elevation shows a really strong disparity between wet and dry locations moreso than your other high elevation sites?
 Do you think aspect or some other factor had an effect?
-May be worth exploring in your results section a little further.*
-
-A possible explanation is that La Recoleta is a quaking bog (a transition from aquatic to fen vegetation at the margins of a silting glacial lake) and therefore it is more waterlogged than other sites.
-We have expanded this explanation.
-However, we agree with reviewer 1 that we should not go too far into interpreting the results of single sites.
-
-> *Line 80 – Groundwater flows are somewhat decoupled from high frequency variation in precipitation, but are still linked to lower-frequency long-term changes in precipitation, groundwater recharge and flow to some degree.
+May be worth exploring in your results section a little further.
+Line 80 – Groundwater flows are somewhat decoupled from high frequency variation in precipitation, but are still linked to lower-frequency long-term changes in precipitation, groundwater recharge and flow to some degree.
 Focusing some attention on this phenomenon and possibly your discussion may be useful.
@@
 drying regions.
-It would be beneficial to focus some of your discussion on this topic.*
-
-We have expanded the discussion in this direction, noting that the buffer effect is dependent on future changes on precipitation, and describing the latest projection for the study area (a decrease of 10 % by the end of the century).
-
-> *Line 86 change representativeness to representation for simplicity.*
-
-Changed.
-
-> *The materials and methods section needs to be much more robust in terms of explaining the statistical platform used to perform the analyses, the packages used, citations for those platforms and packages.
+It would be beneficial to focus some of your discussion on this topic.Line 86 change representativeness to representation for simplicity.
+The materials and methods section needs to be much more robust in terms of explaining the statistical platform used to perform the analyses, the packages used, citations for those platforms and packages.
 Why the analyses are appropriate.
 There’s really not much of a description of how linear modeling was performed, assumptions etc.
-A linear mixed-model approach would make a lot of sense instead of a simple linear model, grouping wet and dry locations by a “Site” factor to control for latent processes.*
-
-Please see a previous response to a similar comment.
-
-> *Need details on sample sizes in your methods.*
-
-We have included the sample sizes.
-
-> *Lines 99-101 Also include quantitative descriptors of climate in the study area (mm precip/year min and max temp/year).*
-
-We have included this information.
-
-> *Line 110 – provide sampling resolution of loggers. A description of why was 5cm depth chosen is needed.*
-
-We have included this information, and the explanation for the sampling depth.
-
-> *Please include a map of the study sites with an inset to show regional/continental position.*
-
-We have included the map.
-
-> *Lines 123 – 133 – some discussion on why these are biologically or environmentally meaningful statistics is needed.
+A linear mixed-model approach would make a lot of sense instead of a simple linear model, grouping wet and dry locations by a “Site” factor to control for latent processes.
+Need details on sample sizes in your methods.Lines 99-101 Also include quantitative descriptors of climate in the study area (mm precip/year min and max temp/year).
+Line 110 – provide sampling resolution of loggers.
+A description of why was 5cm depth chosen is needed.
+Please include a map of the study sites with an inset to show regional/continental position.
+Lines 123 – 133 – some discussion on why these are biologically or environmentally meaningful statistics is needed.
 Are species distributions controlled by winter minimums?
-Carbon dynamics by warm temperatures?*
-
-We have included a description of the biological meaning of the bioclimatic variables we chose, with a relevant reference for more information.
-
-> *Statistical analysis: Some simple statistics on precipitation and temperature trends over the last ~hundred years in this study region would be helpful to broaden discussion regarding future uncertainty of microclimates.
+Carbon dynamics by warm temperatures?
+Statistical analysis:Some simple statistics on precipitation and temperature trends over the last ~hundred years in this study region would be helpful to broaden discussion regarding future uncertainty of microclimates.
 That would also help to put your study into perspective / macroclimatic context.
@@
 Dendrochronology literature frequently includes analytical methods for such trend analyses.
-*
-
-We are sorry but we do not have access to the data necessary to make such analysis, since analysing the evolution of climate in the study region was not the goal of this project.
-
-> *Results: More interpretation of the results is needed.
-Line 155 If you’re only looking at wet vs dry sites, wouldn’t you expect wet sites to have less variability even if the source was not groundwater?*
-
-We interpret the results in the discussion, and comment the generality of the moisture effect.
-
-> *Line 167 – Delete the word prove. The evidence required to prove something is very high. Suggest changing to: The results presented provide strong evidence of...*
-
-Changed.
-
-> *Line 173 – Raney 2014 has similar findings from year-round data for a single year.
+Results: More interpretation of the results is needed.
+Line 155 If you’re only looking at wet vs dry sites, wouldn’t you expect wet sites to have less variability even if the source was not groundwater?
+Line 167 – Delete the word prove.
+The evidence required to prove something is very high.
+Suggest changing to: The results presented provide strong evidence of...
+Line 173 – Raney 2014 has similar findings from year-round data for a single year.
 Warmer in winter, cooler in summer.
-You should highlight that your findings were extended to a full year and the pattern is reproducible among years out to five years.
-The fact that your data possess a strong-elevational relationship, i.e., stronger under warming environment is worth more attention.*
-
-We have added these considerations.
-
-> *Line 194 – Replace “highness” with “height”*
-
-Changed.
-
-> *Line 200 – you elude to some topographic microclimate effects on your sites.
+You should highlight that your findings were extended to a full year and the pattern is
+reproducible among years out to five years.
+The fact that your data possess a strong-elevational relationship, i.e., stronger under warming environment is worth more attention.
+Line 194 – Replace “highness” with “height”Line 200 – you elude to some topographic microclimate effects on your sites.
 That is fine as they influence both wet and dry locations equally, but may be worth including aspect measurements for your sites in Table 1.
-Neat datasets, I look forward to reading the final manuscript.*
-
-Only one of the sites (La Malva) is in a slope and thus has an aspect, the other sites are flat. We have changed the sentence to make this clear.
+Neat datasets, I look forward to reading the final manuscript.

### 2020-08-29 README.Rmd
Revert "readme picture"
This reverts commit 64459c81772228a298a7502651cf5307ff5ceb57.

---
title: "README.md"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

![Alkaline fen in the Cantabrian Mountains](data/Sierra Cebolleda.JPG)

# Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils

This repository stores all the information related to the publication *Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils*, including the raw datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the manuscript.

## Contents

This repository is organised following the advice of [Wilson et al. 2017](https://doi.org/10.1371/journal.pcbi.1005510) for recording and storing research projects.

The following materials are available in the folders of this repository:

* `data` Original raw files including soil temperature recording.
* `doc` Files to create the manuscript using `Rmarkdown`.
* `results` Output of the `R` scripts, including clean data and figures.
* `src` Scripts in `R` language used to clean the raw data and perform the analyses of the manuscript.

## Citation

Please cite the dataset as `CITATION`

### 2020-08-29 src/analysis.R
Revert "readme, license, cleaning"
This reverts commit 192e4d49873d6651f1d51f2df53f2ed8160945c7.

@@ -1,10 +1,10 @@
-library(tidyverse); library(lubridate)
+library(tidyverse); library(ggthemes); library(lubridate); library(here)
 
-load("results/logs.RData")
+load(here::here("results", "logs.RData"))
 
 # Calculations
 
-## Bioclimatic variables as defined by US Geological Survey, WorlClim
+## Bioclimatic variables US Geological Survey, WorlClim
 
 ### Bio 2 Annual Mean Diurnal Range
 
@@ -56,10 +56,7 @@ t.testEFP <- function(df)
   t <- t.test(df$Waterlogged, df$Dry, alternative = "less", paired = TRUE)
   data.frame(t$statistic,
              t$p.value,
-             t$estimate,
-             t$conf.int[1],
-             t$conf.int[2]
-  )
+             t$estimate)
 }
 
 t.testEFPgreater <- function(df) 
@@ -67,23 +64,20 @@ t.testEFPgreater <- function(df)
   t <- t.test(df$Waterlogged, df$Dry, alternative = "greater", paired = TRUE)
   data.frame(t$statistic,
              t$p.value,
-             t$estimate,
-             t$conf.int[1],
-             t$conf.int[2]
-  ) # In the case of the minimum temperature is the opposite hypothesis
+             t$estimate) # In the case of the minimum temperature is the opposite hypothesis
 }
 
 rbind(
   merge(header[, 1:8], bioclimatics, by = "Logger") %>% 
     gather(Trait, Value, `Diurnal range`:`Annual range`) %>%
-    dplyr::select(Site, Groundwater, Value, Trait, Value) %>%
+    select(Site, Groundwater, Value, Trait, Value) %>%
     spread(Groundwater, Value) %>%
     filter(! Trait %in% "Winter min") %>%
     group_by(Trait) %>%
     do(t.testEFP(.)),
   merge(header[, 1:8], bioclimatics, by = "Logger") %>% 
     gather(Trait, Value, `Diurnal range`:`Annual range`) %>%
-    dplyr::select(Site, Groundwater, Value, Trait, Value) %>%
+    select(Site, Groundwater, Value, Trait, Value) %>%
     spread(Groundwater, Value) %>%
     filter(Trait %in% "Winter min") %>%
     group_by(Trait) %>%
@@ -104,16 +98,13 @@ temperatures %>%
   recording.period
 
 merge(header, recording.period) %>%
[…]

### 2020-08-29 src/cleaning.R

@@ -1,4 +1,4 @@
-library(tidyverse); library(lubridate)
+library(tidyverse); library(sp); library(lubridate)
 
 # Function reads and formats GP5WSHELL csv export files
 
@@ -13,6 +13,26 @@ read_gp5 <- function(x)
            Logger = gsub("\\_.*", "" , x)) %>%
     dplyr::select(Logger, Time, Temperature)
 }
+
+# # Functions convert coordinates to decimal degrees
+# 
+# decdeg30T <- function(x) { x -> o1 
+#   o1 %>% dplyr::select(ED50X, ED50Y) -> o2
+#   SpatialPoints(o2, proj4string = CRS("+proj=utm +zone=30T +datum=WGS84")) -> o3
+#   spTransform(o3, CRS("+proj=longlat +datum=WGS84")) %>% data.frame -> o4
+#   colnames(o4) <- c("Longitude", "Latitude")
+#   o1 %>% dplyr::select(-c(ED50X, ED50Y)) -> o5
+#   cbind(o5, o4)
+# }
+# 
+# decdeg29T <- function(x) { x -> o1 
+#   o1 %>% dplyr::select(ED50X, ED50Y) -> o2
+#   SpatialPoints(o2, proj4string = CRS("+proj=utm +zone=29T +datum=WGS84")) -> o3
+#   spTransform(o3, CRS("+proj=longlat +datum=WGS84")) %>% data.frame -> o4
+#   colnames(o4) <- c("Longitude", "Latitude")
+#   o1 %>% dplyr::select(-c(ED50X, ED50Y)) -> o5
+#   cbind(o5, o4)
+# }
 
 # Read logger header data
 
@@ -29,7 +49,27 @@ header %>%
 header %>% select(-Elevation) %>% 
   merge(average.elevation) -> header # Only one elevation value per site
 
+# Map of the loggers
+
+header %>%
+  ggplot(aes(x = Longitude, y = Latitude)) + 
+  labs(x = "Longitude (º)", y = "Latitude (º)") +
+  geom_polygon(data = map_data("world"), aes(x = long, y = lat, group = group), 
+               color = "black", fill = "grey") + 
+  geom_point(aes(color = "indianred"), alpha = 0.5, size = 3, show.legend = FALSE) +
+  geom_text(aes(x = -3, y = 42, label = "Spain"), size = 4) +
+  geom_text(aes(x = 1, y = 44.8, label = "France"), size = 4) +
[…]

### 2020-08-29 doc/manuscript/v2_IJOC/response-reviewers.Rmd
Revert "map of the sites"
This reverts commit 5e6ba1be55006768cd3e0dd015b7150cf954eb21.

---
title: "Ref. GEB-2020-0155 - Response to reviewers"
author: "Eduardo Fernández Pascual & Eva Correia-Álvarez"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Please, find below my revision of the manuscript. I have considered and applied all suggestions. Considering the major issues, I have (1) expanded the explanation of the chosen biome and the creation of the species list; (2) clarified the missing full texts issue, providing a more detailed description, a PRISMA flow chart, and BibTeX files with the missing references; and (3) expanded the database to include the controlling factors required by the reviewers, such as substrate and number of containers.

# Response to the Editor

> I recommend that the manuscript be rejected with an opportunity to resubmit. The reviewers and myself agreed that the premise of the study was very interesting and relevant to the journal's audience, and appreciate the extensive work that went into its creation. However, a number of issues were identified, which I think preclude publication in present form, and whose resolution would potentially require extensive work on the part of the authors.

> First, the database's content and scope seems broader than warranted by its contents. The first reviewer and I both agree that the data do not necessarily provide full coverage of the proposed biome, and may contain species from beyond it. Additionally, I note that all of the sPlot records used to create a species list are from the Northern hemisphere, when the temperatre / broadleaf forest biome occurs widely in the southern hemisphere. While some of the identified studies end up sourcing data from southern hemisphere sites, it seems inaccurate and overly ambitious to suggest the dataset cover this biome. I am not sure that database with only northern hemisphere coverage of one biome, and with strong biases towards the species and sites represented in sPlot, would be of sufficient interest to receive support in the journal. I encourage the authors to expand the scope of their study, or alternatively provide a robust justification for the current scope, and a clearer title / narrative concomitant with the data.

The editor is right that the biome is also found in the southern hemisphere, although the extension there is smaller than in the northern hemisphere. In the new version of the manuscript, there is a longer explanation of why the NH biome was chosen, and how the species were selected. Summarising, the temperate broadleaf and mixed forests of the Northern Hemisphere were chosen as the subject of a sPlot project. Only the NH was chosen because it forms a coherent biogeographical unit, since it shared a common history during the Tertiary. The SH forests, on the other hand, are separated (and smaller) units and were not connected to the northern forests. This database was created for that specific sPlot project, and it was decided to do so because we expected that the number of germination references would be the highest, since many classical research groups have worked within the NH temperate broadleaf and mixed forests biome. This is developed in more detail in the manuscript and in the response to the reviewers.

In addition, the title has been changed to *Sylvan Seeds, a database and web app to explore the seed germination ecology of temperate broadleaf and mixed forests of the northern hemisphere*.

In any case, and in my personal view, I believe this is the more complete database of raw germination data for a coherent ecological unit and, spanning three continents of a densely populated and significant biome, it has enough global interest for the journal.

> Reviewer #1 also notes that 1/3 of the considered studies were omitted due to access issues; I suggest the authors try to resolve this point, or robustly clarify the justification for this choice. As the manuscript stands currently this text is not very convincing for a reader.

Please see the response to the reviewers about this point.

> Second, the first reviewer identified several issues related to the methodology used for the meta-analysis and literature search. I am sympathetic to these points and suggest the authors revisit their search criteria and data, or provide a robust response to these criticisms.

Please see the response to the reviewers about this point.

> As such, I do not think the data paper is ready to be accepted in present form, but am very willing to consider a thoroughly revised version.

# Responde to Referee 1

> The manuscript describes a dataset of species germination rates where the species were selected based on the sPlot species lists for temperate biomes. The shiny app is a nice addition that could be of use to the broader public with proper dissemination. It is interesting data to compile that will surely be valuable, but the scope of the biomes investigated and a lack of important methodological detail from the germination experiments limits the utility of the data. These concerns and a few others are detailed below.

> I wonder if defining the biome as “temperate broadleaf and mixed forests” is appropriate. There appears to be a number of species that generally occur in grasslands and biomes/habitats outside of forests (e.g. Dactylis glomerata, Achillea millifolium, Festuca ovina, etc.). Did each paper specify that the populations of seed collection were forested? If not, I recommend using something like a Whittaker biome diagram to describe the biomes as temperate rain, deciduous and dry forests, and grass/shrublands. Or just add additional biomes from the WWF classifications.

The new version of the manuscript contains a more detailed explanation of the reason to select the biome, and of the methodology used to construct the list of species. Please see the response to the editor and the manuscript for more comments on this.

The TBMF biome (in the northern hemisphere) is the area which was used to create the list of frequent species, for which the database was then compiled.  Biomes are defined as *relatively large units of land containing a distinct assemblage of natural communities and species, with boundaries that approximate the original extent of natural communities prior to major land-use change*. Thus, the TBMF biome does not contain only forests, but many other habitats such as heathlands, meadows, alpine grasslands, mires, dunes, etc. However, the plots used to create the list were filtered to be only forested plots (i.e., they were selected when they met the criterion that tree species dominated their plant cover). The list that has been produced includes species that occur with high frequency in forested plots of the TBMF biome in the nothern hemisphere. These are trees, shrubs, and herbaceous understory plants adapted to shade. But also species of forest margins, openings, riparian forests, ruderalized forests, etc.

The species mentioned by the reviewer are indeed species of the TBMF biome, and, even if today they may have their optimum in semi-natural meadows and pastures, they can occur also in forested plots (and in fact they do, as sown by the vegetation relevés we used). *Dactylis glomerata* and *Achillea millifolium* are species of mesic meadows (*Molinio-Arrhenatheretea*), which are secondary communities developped by human intervention over habitats that would be naturally occupied by temperate forests. The traditional interpretation of the origin of these communities is that they originated in forest margin communities that colonized the openings created by humans. Therefore, it is not surprising that these species are recorded with relative frequency in forest plots, as indeed they are. Similarly, it is not surprising that dry grassland (*Festuco-Brometea*) species such as *Festuca ovina* are also recorded in forest plots, especially in contact areas, as these are also communities that expanded over landscapes that would naturally be forests.

I agree that the seed lot coordinates that are provided with the database could be used to construct a Whittaker diagram of the seed lots, or alternatively to asign them to WWF biomes. But I feel that this is more of an analytical usage of the data, which can be done by any user with the information provided in the dataset. Still, I am willing to do such tasks if the editor finds them essential.

> The justification for restricting the search to certain biomes and species is weak (Lines 49-50), especially given sPlot is a global database. If picking these biomes covers most of the available data, why not extend beyond this biome given the added work is relatively minimal? Then the dataset would be truly global. There are a lot of papers on germination outside of this biome that could make for valuable comparisons (e.g., tundra/alpine biomes from Vandvik et al. 2017. Botany 95:73-88; Tropical forests from Yu et al. 2008. Plant Ecology 297:1-16). Why not expand the search to all species within sPlot? If the search is expanded to all species in sPlot, does the number of articles requiring processing become unreasonable? This would make for a better justification for restricting the search.

I have included a more detailed description of the reasons behind chosing the TBMF biome to construct the database: *The building of the dataset was part of a research project conducted within the sPlot initiative (Bruelheide et al., 2019), specifically sPlot’s project #12 (https://www.idiv.de/en/sdiv/working_groups/wg_pool/splot/projects.html). The goal of sPlot #12 was to study taxonomical and functional plant diversity in deciduous forests, which are dominant in temperate latitudes with mild to cold winters and relatively warm and wet summers. In the northern hemisphere, they also form a coherent evolutionary and biogeographic unit, sharing a common Tertiary history until they became separated between North America, Asia and Europe (Milne & Abbott, 2002).*

Completing the database for all the species in sPlot is a worthwhile long-term goal (as mentioned in the *Utility of the database* section of the manuscript), and one that I am interested in contributing to in the future. But it was completely beyond the resources I had to construct this database, which depended on a 2-year personal grant. It took me 9 months of focused work to build a database for 1,393 species. There are 54,519 species in sPlot 2.1 (*Bruelheide, H., Dengler, J., Jiménez‐Alfaro, B., Purschke, O., Hennekens, S. M., Chytrý, M., ... & Aubin, I. (2019). sPlot–A new tool for global vegetation analyses. Journal of Vegetation Science, 30(2), 161-186*). Therefore, a rough estimate is that it would take me 29 years of work to do a database for all sPlot species! Even if the time would probably be shorter (because I do expect the number of references would be lower outside the TBMF biome), the number of references would be expanding each year with new research. So, because one has to start somewhere, I chose to do it in connection to sPlot project 12, which focused on temperate broadleaf and mixed forests of the northern hemisphere; and I chose to do it because this biome has received much attention in seed research.

> The literature search would benefit from more description and/or investigation. What made a full text impossible to access? Not having access to a few full texts is understandable, but 236 would make up more than 1/3 of the 611 that were to be relevant. This data could be very important to the completeness of the dataset. Using PRISMA guidelines for proper meta-analysis data gather is highly recommended (http://www.prisma-statement.org).

Just to be precise, the proportion of not-found information is actually smaller. Of 1489 references for which I searched for the full text, I found 1253 (NOTE: While doing the revisions, I have found and corrected small discrepancies in the numbers of references, caused because a few of the records were removed by a mismatch during taxon name standardization). Of these, 639 (51%) were not relevant. If we assume the same proportion of relevant references in the cases for which I could not find the full text, we can estimate that there were 116 relevant articles which I could not include in the database: 16% of the potentially relevant references (614 found + 116 estimated not found = 730; 116/730 =  0.16). Furthermore, I must note that information is not homogenesouly distributed thorough the references: there are many references for some popular species, and fewer references for rarer species. Therefore the amount of valuable information does not increase linearly with increasing number of references, and it is very likely that a large proportion of that 16% is made up of references considering species which are already in the database. In conclusion, I believe that the amount of missing information is within reasonable limits.

Be that as it may, I have improved the description of data gathering in the text, including a description of the process of finding full texts: *An effort was made to access the full text of all these references, using in order the following channels: (i) “Find Full Text” functionality in EndNote X9, (ii) online access to the journals using the institutional subscriptions of the University of Oviedo; (iiI) request to the authors through ResearchGate; (iv) general search of pdf files in Google using the article details. Unfortunately, 236 references could not be accessed. Most of these belonged to smaller publishers not included in the subscriptions of the University of Oviedo, and to older journal volumes which had not been digitalised.*

I have also included some supplementary materials to describe the data gathering: the boolean search string I used in the Web of Science, and a PRISMA flow chart.

Additionally, I have added the lists of the three types of references (non-relevant, found and not-found) as BibTeX supplementary files. In this way, it will be possible to check what I did, and to complete it in the future by accessing the references that I could not find.

> There appears to be a lot of important information missing in the dataset that would allow a user of the dataset to control for methodological artifacts or biases. For example: what kind of substrate were the seeds germinated in (Agar? Soil? Sand?)? What was the density of seeds sown? How large were the containers holding the seeds? How many replicates were there to determine the number germinated versus germinable? Were any treatments, such as drought or nutrient addition, used in the experiment? Were the seeds sown in a monoculture?

As requested, I have revised the references to retrieve the requested information, and added it to the database: type of substrate; type and dimensions of the containers; number of containers and number of seeds sown per container; treatments of sterilization and nutrients. Density of seeds is not reported in the references, but it can be estimated from the dimension of the containers and the number of seeds sown. All cases included in the database correspond to monoculture experiments, and special treatments (drought, heavy meatals, smoke, etc.) were not included; this is now stated in the text. Please see the manuscript for the description of the new variables in the database.

> Most of the studies would have replicated germination trials (e.g. sown 100 seeds across 5 replicates: 20 seeds per replicate). Therefore, there is likely true error estimates (and mean percent germination) that can be obtained from the papers. These are a highly desirable metrics from the papers and would improve the error estimates in the shiny app as well.

As requested, I have included the number of containers ("replicates") and number of seeds sown per replicate in the database.

> Line 12- add an “and” so that it reads “darkness, and constant or alternating”

Done.

> Line 16- Will the data collection be continuous? Maybe assign a final year here.

Done.

> Line 31- add an of; “all of these”

Done.

> Line 35- use commas instead of –

Done.

> Line 39-40- “which implies a set of derived complications in the analytical treatment of the data”- unclear what this means and how it applies to germination as a number of successes given seeds sown (trials).

I have removed the whole sentence.

> Line 45- Rather than “at this and that temperature”, try “across a temperature gradient”?

Done.

> Line 95- give a brief explanation of GA3

Done.

> Line 104-105- Remove highlight of Fig. 1.

Done.

> Line 108- incorrect degree symbol.

It was the degree symbol rendered in the Calibri font, the new manuscript is in Arial.

> Line 154- remove the “a” to read “can have wide applicability”

Done.

> Line 155-157- Try “The data can extend current trends in global analyses of plant traits at species (Diaz et al., 2016) and community levels (Bruelheide t al., 2018) to seed germination.”

Done.

> Line 160- Remove “to”

Done.

> Line 262- As the authors note, the references will need cleaning.

Indeed.

> Can the species names be searchable in addition to the drop-down menu in the shiny app?

The species name can be deleted, and then one can write in the box. This will trigger an autocomplete, which will allow to search.

> There are special character errors in the “Reference” column of the dataset: e.g., «ali?kan, O., et al. (2012). "Influences of presowing treatments on the germination and emergence of fig seeds (Ficus carica L.)." Acta Scientiarum. Agronomy 34(3): 293-297.

This should be fixed now.

# Responde to Referee 2

> This data paper introduces a new resource available to ecologists: a data base and associated app summarising seed germination data for a large number of plant species in the temperate broadleaf/mixed forest biome. The authors used an extensive literature search to identify sources of germination data, and have gone through a substantial effort in extracting data from a large number of published articles. The resulting ‘Sylvan Seeds’ data base and app has the potential to be a valuable tool for researchers and applied foresters. I had a quick play around with the app – it provides nice species-specific data summaries (although I think the data base itself is going to be a more useful resource).

> I have little to criticise in terms of the approaches used (which seem sound), but believe that the presentation would benefit from a few changes to make it more reader friendly and accessible for people whose main interest might not be in seed ecology and to potential non-academic users of the database/app:

> L1 (Title): I am not sure how many people will immediately know what a ‘shiny app’ is. (I did not...) Could another more general term be used?

I changed it to *web app*.

> L5 (Motivation): This section could perhaps start with a sentence on why seed germination is an important phenomenon in ecology?

Done.

> L9-10: “Data has been gathered using a meta-analytical approach to search the literature.” Meta-analytical approach may suggest that some kind of data analysis was involved. Could be changed to something along the lines of “Data were gathered using a systematic literature search approach” or similar. (Later in the text you could then mention that the exact approach taken was inspired by approaches typically used in meta-analyses?)

Changed as suggested.

> L14: “across the biome and beyond” – unclear meaning

Changed to *Seed lots collected throughout the biome and in neighbouring areas*.

> L16: today -> March 2019?

Changed to the year of the most recent seed lot, 2017.

> L17: frequent -> common?

I believe it is more appropiate to use frequent, as frequency is the common term in plant ecology: *Frequency is the number of times a plant species occurs in a given number of quadrats*. Frequency, thus defined, was the criterion used to include species in the search.

> L20: wide -> wider?

Changed.

> L25: Again, considering the broad readership of Global Ecology and Biogeography, it might be helpful to start with a sentence or two on the importance of studying seed traits/germination/seed ecology (and – given the journal – ideally also something on why a global approach can contribute to the field of seed ecology).

Done.

> L41: scarcity of -> lack of? (Alternatively, if such data bases exist, but there are only few of them, stick to original wording but mention which ones exist.)

Changed to *lack of*.

> L45: at this and that temperature -> at various temperatures

Changed to *accross a temperature gradient* as suggested by referee 1.

> L49-50: home of many of the classical research groups in seed ecology -> research focus of (…) or similar?

Changed to *At the same time, a relatively high number of classic seed biology research groups have worked within the TBMFNH areas, e.g. in the USA (Baskin, 1995), the UK (Grime et al., 1981), Belgium (Van Assche & Vandelook, 2006) and Japan (Washitani, 1988). *.

> L51: see previous comment regarding reference to meta-analysis

Changed to *To gather data for the project, a systematic literature search (Koricheva et al., 2013) was applied to a list of frequent species representing the flora of the TBMFNH.*

> L57: not all readers might be familiar with the term vegetation relevé -> provide brief explanation?

The following explanation has been added: *(i.e. records of plant species co-occurring in a sampling plot)*

> L58: the reference to sPlot’s project #12 might not mean much to people not involved with sPlot. -> Provide link to website or further explanation?

I have included the following description: *The building of the dataset was part of a research project conducted within the sPlot initiative (Bruelheide et al., 2019), specifically sPlot’s project #12 (https://www.idiv.de/en/sdiv/working_groups/wg_pool/splot/projects.html). The goal of sPlot #12 was to study taxonomical and functional plant diversity in deciduous forests, which are dominant in temperate latitudes with mild to cold winters and relatively warm and wet summers. In the northern hemisphere, they also form a coherent evolutionary and biogeographic unit, sharing a common Tertiary history until they became separated between North America, Asia and Europe (Milne & Abbott, 2002).*. Unfortunately, no articles derived from sPlot #12 have been finished so far, so I cannot cite them, which would be the ideal solution.

> L72: ‘To obtain a final list of species (…)’. The primary reason for focusing on the subset of species that were present in a minimum of 5% of the relevés of an ecoregion was presumably to make the work achievable given the large number of species (>7000) in the full species pool?

There was the practical reason mentioned by the reviewer, but also a methodological consideration. I have added the following explanation: *There were two reasons to use this frequency threshold: (i) it made the search achievable with the resources available; and (ii) it filtered out rare species that may not be representative of TBMFNH, and which could have been recorded for a variety of reasons (wrong taxonomical identifications, mistakes in plot coordinates, border areas, unnatural areas such as parks, etc.).*

> L81-82: “(…) for 236 it was impossible to access the full text.” Unclear meaning: Was this because of lack of institutional licenses for accessing journals, or because of something else?

Please see the response to referee 1 about the same question. The causes are now described as follows: *An effort was made to access the full text of all these references, using in order the following channels: (i) “Find Full Text” functionality in EndNote X9, (ii) online access to the journals using the institutional subscriptions of the University of Oviedo; (iiI) request to the authors through ResearchGate; (iv) general search of pdf files in Google using the article details. Unfortunately, 236 references could not be accessed. Most of these belonged to smaller publishers not included in the subscriptions of the University of Oviedo, and to older journal volumes which had not been digitalised.*.

> L95: GA3 -> gibberellic acid (to help readers who are less familiar with the topic, could perhaps also mention that GA3 speeds up seed germination)

Done.

> L96: the term ‘response variable’ suggests that you conducted some kind of statistical analysis. -> Delete term or change wording.

Removed *the response variable* part of the sentence.

### 2020-08-29 src/analysis.R

@@ -157,89 +157,9 @@ bio %>%
          p = round(p, 3),
          R2 = round(R2, 2)) -> table3
 
-#----------------------------------------------------------------------------
-
 # Figures
 
-## Figure 1 - Map
-
-library(raster)
-
-### Get map of Spain
-
-spaingeo <- getData("GADM", country = "spain", level = 1)
-spaindf <- fortify(spaingeo)
-
-spainalt <- getData("alt", country = "spain", level = 1)
-altdf  <- data.frame(rasterToPoints(spainalt))
-colnames(altdf) = c("lon", "lat", "alt")
-altdf %>% filter(alt < 2650) -> altdf
-
-### Draw map of Spain
-
-header %>%
-  group_by(Site, Mire) %>%
-  summarise(long = mean(Longitude), lat = mean(Latitude)) %>%
-  ggplot(aes(long, lat)) +
-  geom_tile(data = altdf, aes(x = lon, y = lat, fill = alt)) +
-  scale_fill_gradient(name = "Elevation (m asl)", low = "grey96", high = "black",
-                      breaks = c(0, 2600)) +
-  geom_blank(data = spaindf, aes(x = long, y = lat)) +
-  geom_map(data = spaindf, map = spaindf,
-           aes(group = group, map_id = id),
-           fill = NA, color = "black", size = .3) +
-  geom_point(size = 2) +
-  ggrepel::geom_label_repel(aes(label = Site), box.padding = .6, size = 2.5) +
-  labs(x = "Longitude (º)", y = "Latitude (º)") +
-  geom_text(aes(x = -6.1, y = 43.75, label = "Cantabrian Sea"), size = 3.5) +
-  geom_text(aes(x = -6.1, y = 43.4, label = "Asturies"), size = 5) +
-  geom_text(aes(x = -5.1, y = 42.9, label = "León"), size = 5) +
-  scale_x_continuous(limits = c(-6.5, -4.7), expand = c(0, 0)) +
-  scale_y_continuous(limits = c(42.8, 43.9), expand = c(0, 0)) +
-  ggthemes::theme_map() +
-  theme(panel.background = element_rect(color = "black", fill = "grey96"),
-        legend.position = "bottom",
-        legend.direction = "horizontal",
-        legend.background = element_blank(),
-        legend.box.background = element_rect(colour = NA),
-        legend.title = element_text(size = 10),
-        legend.text = element_text(size = 8),
[…]

### 2020-08-29 doc/manuscript/v2_IJOC/ms2.Rmd
Revert "Revert "starting IJOC revision""
This reverts commit d4575d285797df49753689d582e79fc79968920a.

@@
 
-This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W) (Fig.
-\@ref(fig:fig1)a).
+This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W).
 Local geography and climate are dominated by the Cantabrian Mountains (> 1500 m above sea level), which run parallel to the coast and trap the prevailing NW Atlantic winds.
@@
 We selected `r length(unique(temperatures$Logger))/2` mire sites representing the regional elevation gradient of mire vegetation and the different mire types (Table 1).
-In each site, we buried two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany) at a depth of 5 cm below the upper layer of the soil: one datalogger in a flat waterlogged spot within the mire; the other one in the close vicinity, but in a flat and dry upland area outside the mire (Fig.
-\@ref(fig:fig1)b).
+In each site, we buried two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany) at a depth of 5 cm below the upper layer of the soil: one datalogger in a flat waterlogged spot within the mire; the other one in the close vicinity, but in a flat and dry area outside the mire.
 The vegetation was always either mire or pasture, with no shrubs, trees or any other landscape features shading the measuring points.
@@
 Time series showed considerably less variation at waterlogged points, for all eight site comparisons  (Fig.
-\@ref(fig:fig2)).
+\@ref(fig:fig1)).
 
 Bioclimatic variables (Table 2 and Fig.
-\@ref(fig:fig3)) supported this notion.
+\@ref(fig:fig2)) supported this notion.
 The mean annual range was wider at dry points, as was the mean diurnal range.
@@
 The opposite was true for the minimums of the coldest months, in which case temperature was generally colder at dry points, although the difference was less pronounced than for the maximums (Fig.
-\@ref(fig:fig3)).
+\@ref(fig:fig2)).
 
@@
 The CHELSA climatic model (Table 3 and Fig.
-\@ref(fig:fig4)) predicted well the minimums of the coldest month, especially at dry points (R^2^ = 0.78) but also at waterlogged points (R^2^ = 0.66).
+\@ref(fig:fig3)) predicted well the minimums of the coldest month, especially at dry points (R^2^ = 0.78) but also at waterlogged points (R^2^ = 0.66).
 CHELSA also predicted to some extent the maximums of the warmest month at dry points (R^2^ = 0.42), but failed to predict waterlogged points (R^2^ = 0.16).
@@
 
-```{r fig1, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "(a) Location of the study sites in northern Spain.
-(b) Representation of the sampling design at El Riotuertu site.
-The position of the dry and waterlogged temperature logggers is indicated, the blue area represents the extent of the waterlogged mire habitat.", fig.align = "center", fig.width=7, fig.height=6}
-knitr::include_graphics("../Fig1.png")
-```
-
-```{r fig2, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites.
+```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites.
 The blue series was recorded within the mire, in a waterlogged area.
@@
 Dataloggers were buried at 5 cm depth.", fig.align = "center", fig.width=7, fig.height=6}
+plot1
+```
+
+```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "Average bioclimatic variables in the dry and waterlogged points.
+The bars represent the mean value, and the brackets the standard error of 8 records.", fig.align = "center", fig.width=7, fig.height=3}
 plot2
@@
 
-```{r fig3, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Average bioclimatic variables in the dry and waterlogged points.
-The bars represent the mean value, and the brackets the standard error of 8 records.", fig.align = "center", fig.width=7, fig.height=3}
+```{r fig3, echo = FALSE, fig.pos = "H", fig.cap = "Scatter plots of the bioclimatic variables predicted by CHELSA air temperatures vs.
+soil temperatures measured in situ, in dry and waterlogged points.", fig.align = "center", fig.width=7, fig.height=7}
 plot3
 ```
-
-```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Scatter plots of the bioclimatic variables predicted by CHELSA air temperatures vs.
-soil temperatures measured in situ, in dry and waterlogged points.", fig.align = "center", fig.width=7, fig.height=7}
-plot4
-```

### 2020-08-29 src/analysis.R
Revert "Revert "map of the sites""
This reverts commit 5ea97e167efffefd974ef25af30df59a3426076c.

@@ -157,9 +157,89 @@ bio %>%
          p = round(p, 3),
          R2 = round(R2, 2)) -> table3
 
+#----------------------------------------------------------------------------
+
 # Figures
 
-## Figure 1 - Datalogger records
+## Figure 1 - Map
+
+library(raster)
+
+### Get map of Spain
+
+spaingeo <- getData("GADM", country = "spain", level = 1)
+spaindf <- fortify(spaingeo)
+
+spainalt <- getData("alt", country = "spain", level = 1)
+altdf  <- data.frame(rasterToPoints(spainalt))
+colnames(altdf) = c("lon", "lat", "alt")
+altdf %>% filter(alt < 2650) -> altdf
+
+### Draw map of Spain
+
+header %>%
+  group_by(Site, Mire) %>%
+  summarise(long = mean(Longitude), lat = mean(Latitude)) %>%
+  ggplot(aes(long, lat)) +
+  geom_tile(data = altdf, aes(x = lon, y = lat, fill = alt)) +
+  scale_fill_gradient(name = "Elevation (m asl)", low = "grey96", high = "black",
+                      breaks = c(0, 2600)) +
+  geom_blank(data = spaindf, aes(x = long, y = lat)) +
+  geom_map(data = spaindf, map = spaindf,
+           aes(group = group, map_id = id),
+           fill = NA, color = "black", size = .3) +
+  geom_point(size = 2) +
+  ggrepel::geom_label_repel(aes(label = Site), box.padding = .6, size = 2.5) +
+  labs(x = "Longitude (º)", y = "Latitude (º)") +
+  geom_text(aes(x = -6.1, y = 43.75, label = "Cantabrian Sea"), size = 3.5) +
+  geom_text(aes(x = -6.1, y = 43.4, label = "Asturies"), size = 5) +
+  geom_text(aes(x = -5.1, y = 42.9, label = "León"), size = 5) +
+  scale_x_continuous(limits = c(-6.5, -4.7), expand = c(0, 0)) +
+  scale_y_continuous(limits = c(42.8, 43.9), expand = c(0, 0)) +
+  ggthemes::theme_map() +
+  theme(panel.background = element_rect(color = "black", fill = "grey96"),
+        legend.position = "bottom",
+        legend.direction = "horizontal",
+        legend.background = element_blank(),
+        legend.box.background = element_rect(colour = NA),
+        legend.title = element_text(size = 10),
[…]

### 2020-08-29 src/analysis.R
Revert "Revert "readme, license, cleaning""
This reverts commit 3bb09c6dae38b6542b23b4a0d5ee89c0eed368e9.

@@ -1,10 +1,10 @@
-library(tidyverse); library(ggthemes); library(lubridate); library(here)
+library(tidyverse); library(lubridate)
 
-load(here::here("results", "logs.RData"))
+load("results/logs.RData")
 
 # Calculations
 
-## Bioclimatic variables US Geological Survey, WorlClim
+## Bioclimatic variables as defined by US Geological Survey, WorlClim
 
 ### Bio 2 Annual Mean Diurnal Range
 
@@ -56,7 +56,10 @@ t.testEFP <- function(df)
   t <- t.test(df$Waterlogged, df$Dry, alternative = "less", paired = TRUE)
   data.frame(t$statistic,
              t$p.value,
-             t$estimate)
+             t$estimate,
+             t$conf.int[1],
+             t$conf.int[2]
+  )
 }
 
 t.testEFPgreater <- function(df) 
@@ -64,20 +67,23 @@ t.testEFPgreater <- function(df)
   t <- t.test(df$Waterlogged, df$Dry, alternative = "greater", paired = TRUE)
   data.frame(t$statistic,
              t$p.value,
-             t$estimate) # In the case of the minimum temperature is the opposite hypothesis
+             t$estimate,
+             t$conf.int[1],
+             t$conf.int[2]
+  ) # In the case of the minimum temperature is the opposite hypothesis
 }
 
 rbind(
   merge(header[, 1:8], bioclimatics, by = "Logger") %>% 
     gather(Trait, Value, `Diurnal range`:`Annual range`) %>%
-    select(Site, Groundwater, Value, Trait, Value) %>%
+    dplyr::select(Site, Groundwater, Value, Trait, Value) %>%
     spread(Groundwater, Value) %>%
     filter(! Trait %in% "Winter min") %>%
     group_by(Trait) %>%
     do(t.testEFP(.)),
   merge(header[, 1:8], bioclimatics, by = "Logger") %>% 
     gather(Trait, Value, `Diurnal range`:`Annual range`) %>%
-    select(Site, Groundwater, Value, Trait, Value) %>%
+    dplyr::select(Site, Groundwater, Value, Trait, Value) %>%
     spread(Groundwater, Value) %>%
     filter(Trait %in% "Winter min") %>%
     group_by(Trait) %>%
@@ -98,13 +104,16 @@ temperatures %>%
   recording.period
 
 merge(header, recording.period) %>%
[…]

### 2020-08-29 doc/manuscript.Rmd
Revert "Revert "1st revision 🍋""
This reverts commit 480c26a5915d0a7fb7c44075fc2bb25e3e0936d0.

@@
 ```{r message = FALSE, echo = FALSE, warning = FALSE}
-library(here); library(tidyverse); library(flextable)
-load(here("results", "logs.RData"))
-load(here("results", "MSoutput.RData"))
+library(tidyverse); library(flextable)
+load("../results/logs.RData")
+load("../results/MSoutput.RData")
 ```
@@
 Ecosystems adapt differently to global warming through microclimate factors.
-Mires are sensitive habitats that strongly rely on local soil properties, which makes them a good model to understand how local climate parameters counteract the effects of climate change.
+Mires are sensitive wetland habitats that strongly rely on local soil properties, which makes them a good model to understand how local climate parameters counteract the effects of climate change.
 We quantified the temperature buffering effect in waterlogged mire soils as compared with adjacent dry soils.
@@
 Climate change [@RN3586] affects global biodiversity, from drylands [@RN4672] to forests [@RN4673] and oceans [@RN4674].
-An accurate prediction of species responses requires to focus on physiologically-relevant variables related to critical plant growth periods [@RN4765].
+An accurate prediction of species responses requires to focus on physiologically relevant climatic variables related to critical plant growth periods [@RN4765].
 Traditional models of species responses to global warming are based on macroclimatic data from weather stations.
@@
 As has been shown for European forests [@RN4763], ecosystems usually respond to broad climatic changes through local processes.
-Increasingly there are downscaling efforts based on atmospheric microclimate networks distributed in complex terrain that focus on topographic effects on air temperature [@RN3279], but hydrologic factors under edaphic control requiere more attention.
+Increasingly there are downscaling efforts based on atmospheric microclimate networks distributed in complex terrain that focus on topographic effects on air temperature [@RN3279], but hydrologic factors under edaphic control require more attention [@RN4942].
 
@@
 These are azonal habitats whose existence depends on local soil properties rather than macroclimatic zonation [@RN3328].
+Mires meet certain criteria that make them especially vulnerable to climate change [@RN4675]: (a) preponderance of species that evolved under a cold climate; (b) low productivity due to nutrient limitation, making them sensitive to increased nutrient cycling caused by warming [@RN4676]; and (c) scattered distribution pattern, which limits species dispersal and migration [@RN4677].
+Therefore, mires are priority habitats for biodiversity conservation, harbouring high numbers of endangered species [@RN3122], and supporting highly adapted floras in spatially reduced areas [@RN2960].
+Worryingly, mires retain high levels of methane and carbon which can be released due to global warming [@RN4764].
+Habitat distribution models have predicted a loss of mire area as a consequence of ongoing climate change [@RN2937].
+
 It has been known for a relatively long time that groundwater can produce a buffer effect on soil temperature, keeping soils warmer than air during cold periods, and vice versa [@RN4940; @RN3344; @RN3201].
-Moisture reduces exposure to temperature extremes, facilitating climate refugia [@RN4942].
-Root-zone temperature is a major determinant of plant ecophysiology [@RN3024], so the groundwater buffer effect is expected to allow mire plants to live in a wider range of air temperatures than they could otherwise.
-Indeed, mires have a relatively homogeneous flora across their range despite their broad distribution.
-In Europe, fens are distributed from the Iberian Peninsula to Fennoscandia and from low valleys to the alpine belt [@RN4678].
-In the United States, fens are distributed across the glaciated Midwest and Northeast, as well as portions of the Appalachian Mountains and mountainous West [@RN4933].However, continued aridification is expected to significantly reduce the overall extent of wetlands as it has been shown in the Midwestern United States [@RN4934].
-In the arctic, temperate-continental and suboceanic regions of boreal Russia mires make up an important part of the landscapce, from 20 to 80 % of the surface of different regions stretching from Europe to the Pacific [@RN4937].
-
 Clara M.
@@
 
-Mires meet certain criteria that make them especially vulnerable to climate change [@RN4675]: (a) preponderance of species that evolved under a cold climate; (b) low productivity due to nutrient limitation, making them sensitive to increased nutrient cycling caused by warming [@RN4676]; and (c) scattered distribution pattern, which limits species dispersal and migration [@RN4677].
-Therefore, mires are priority habitats for biodiversity conservation, harbouring high numbers of endangered species [@RN3122], and supporting highly-adapted floras in spatially-reduced areas [@RN2960].
-Worryingly, mires retain high levels of methane and carbon which can be released due to global warming [@RN4764].
-Habitat distribution models have predicted a loss of mire area as a consequence of ongoing climate change [@RN2937].
+Since root-zone temperature is a major determinant of plant ecophysiology [@RN3024], the groundwater buffer effect is expected to allow mire plants to live in a wider range of air temperatures than they could otherwise.
+Indeed, mires have a relatively homogeneous flora across their range despite their broad distribution.
+In Europe, fens are distributed from the Iberian Peninsula to boreal Fennoscandia and from low valleys to the alpine belt [@RN4678].
+In the arctic, temperate-continental and suboceanic regions of boreal Russia mires make up an important part of the landscape, from 20 to 80 % of the surface of different regions stretching from Europe to the Pacific [@RN4937].
+In the United States, fens are distributed across the glaciated Midwest and Northeast, as well as portions of the Appalachian Mountains and mountainous West [@RN4933].
+However, continued aridification is expected to significantly reduce the overall extent of wetlands as it has been shown in the Midwestern United States [@RN4934].
 
 It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change [@RN4936].
-Both bogs and fens are actively peat forming and are dependant on precipitation, whereas fens rely also on sources of telluric water from mineral ground.
-Moreover they are composed from microreliefs and a specific plant cover zonation which makes them specially sensitive to small changes in wetness.
-Climate change affects the quantity, timing and spacial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935].
-Global warming would also result in warmer groundwater delivered to fens and in drier conditionts, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].
+Both bogs and fens are actively peat forming and are dependent on precipitation, whereas fens rely also on sources of telluric water from mineral ground.
+Moreover, they show a mosaic of microreliefs and a specific plant cover zonation which makes them especially sensitive to small changes in wetness.
+Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935].
+Global warming would also result in warmer groundwater delivered to fens and in drier conditions, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].
 
-As is the case for all azonal habitats, locally measured temperatures are essential to understand this response.
+As is the case for all azonal habitats, locally measured temperatures are essential to understand these processes.
 Available references recorded temperatures only on waterlogged soils, using model-derived air temperatures for comparison [@RN2356; @RN4675; @RN4679]; or recorded temperatures in both wet and dry spots but for less than a year, lacking representation throughout the growth cycle of mire vegetation [@RN3204; @RN3060].
 This article provides the first measurement of the thermal buffer against surrounding non-mire areas, based on soil temperatures recorded during a period of five years.
-These measures can be used to support previous evidence about the general effects of soil moisture on thermal buffering [@RN3204; @RN2356; @RN4675; @RN4936; @RN4942] and to determine the magnitude of this buffering within microrefugia habitats.
+These measures can be used to support previous evidence about the general effects of soil moisture on thermal buffering [@RN3204; @RN2356; @RN4675; @RN4936] and to determine the magnitude of this buffering within microrefugia habitats.
 Specifically, we tested the hypotheses that, when compared to adjacent dry soils, waterlogged mire soil are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales.
-In addition, we compare *in situ* measurements with data derived from the CHELSA climatic models [@RN4766].
+In addition, we compared *in situ* measurements with data derived from the CHELSA climatic models [@RN4766].
 
@@
 In the study area, average annual precipitation ranges from 800 mm at the low elevations to 1800 mm at the high mountains, while average annual temperature ranges from 12.5 ºC to 5 ºC (source = Agencia Estatal de Meteorología, http://www.aemet.es/es/serviciosclimaticos/).
-Under the most extreme emissions scenario, the latest projection expect temperatures to increase by 4 ºC and precipitation to decrease by c.
-10 % by the end of the century, although the precipiation projections are subjected to great uncertainty [@RN4939].
-This humid climate harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246].
-Rain-fed raised bogs (https://eunis.eea.europa.eu/habitats/260) are very rare and appear locally on top of acid valley mires (https://eunis.eea.europa.eu/habitats/526) which be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes.
+Under the most extreme emissions scenario, the latest projections expect temperatures to increase by 4 ºC and precipitation to decrease by c.
+10 % by the end of the century, although the precipitation projections are subjected to great uncertainty [@RN4939].
+The present humid and mild climate of this region harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246].
+Rain-fed raised bogs (https://eunis.eea.europa.eu/habitats/260) are very rare and appear locally on raised hummocks of acid valley mires (https://eunis.eea.europa.eu/habitats/526) which be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes.
 Glacial lakes undergoing silting develop transition mires and quaking bogs communities (https://eunis.eea.europa.eu/habitats/1637) in the water-to-land transition.
@@
 et Tx.
-ex Westhoff
-et al.
-1946), we must note that these bogs are very poorly developped over valley mires and in tight connection to the underlying water table coming from streams, therefore all of our sites have waterlogging feed by streams or springs.
+ex Westhoff et al.
+1946), we must note that these bogs are very poorly developed on hummocks over valley mires and are in tight connection to the underlying water table coming from streams.
+Therefore, soil waterlogging in all our sites is fed by streams or springs.
 In each site, we buried two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil.
@@
 \@ref(fig:fig1)b).
-The vegetation was always either mire or pasture, with no shrubs, trees or any other landscape features shading the measuring points.
+The vegetation at the sampling points was always either mire or pasture, with no shrubs, trees or any other landscape features shading the measuring points.
 Dataloggers recorded temperature once every hour and stayed on site for approximately five years, after which we retrieved them and downloaded their records.
@@
 These are classical bioclimatic variables that indicate the limiting factors (extreme temperatures) and seasonality (diurnal and annual range) that organisms must tolerate [@RN4943].
-To compare our measurements with model-based predictions of air temperature, we downloaded from CHELSA the same bioclimatic variables for our measuring coordinates.
+To compare our measurements with model-based predictions of air temperature, we downloaded from CHELSA the same bioclimatic variables for our measuring coordinates (https://chelsa-climate.org/bioclim/).
 
 To test if the differences between dry and waterlogged points were significant, we used paired t-tests (n = 8 paired sites per each of the four bioclimatic variables).
-Tests were one-tailed, according to the following hypotheses: the dry point would have a higher diurnal range, a higher maximum temperature, a lower minimum temperature, and a higher annual range.
-To compare our soil measurements with the CHELSA air temperatures we calculated Pearson's correlation between both sets of values, and we also calculated the root-mean-square error (RMSE) of the CHELSA values as a measure of their accuracy (n = 8 sites per combination of bioclimatic variable and dry/waterlogged condition) using the package *Metrics* [@RN4941].
+Tests were one-tailed, according to the original hypotheses: the dry point would have a higher diurnal range, a higher maximum temperature, a lower minimum temperature, and a higher annual range.
+To compare our soil measurements with the CHELSA air temperatures we calculated Pearson's correlation between both sets of values, and we also calculated the root-mean-square error (RMSE) of the CHELSA values as a measure of their accuracy (n = 8 sites per combination of bioclimatic variable and dry/waterlogged condition).
+We calculated RMSE using its implementation in the package *Metrics* [@RN4941].
 
@@
 The CHELSA climatic model (Table 3 and Fig.
-\@ref(fig:fig4)) provided air temperatures that were significantly correlated (p < 0.05) with soil temperatures only in the case of the minimums of the coldest month, both at dry points (Pearson's r = 0.90) and waterlogged points (Pearson's r  = 0.84).
+\@ref(fig:fig4)) provided air temperatures that significantly correlated (p < 0.05) with soil temperatures only in the case of the minimums of the coldest month, both at dry points (Pearson's r = 0.90) and waterlogged points (Pearson's r  = 0.84).
 The correlation between CHELSA and soil values was marginally significant (p = 0.05) in the case of the maximums of the warmest month, but only at dry points (Pearson's r = 0.71).
 The values of CHELSA did not correlate with soil temperatures in the rest of the cases (p > 0.05).
-RMSE indicated that the predictions of CHELSA were more accurete at dry points for the cases of the annual range, the diurnal range and the winter min; the CHELSA values were more accurate at waterlogged points in the case of the summer max.
+RMSE indicated that the predictions of CHELSA were more accurate at dry points for the cases of the annual range, the diurnal range and the winter min; the CHELSA values were more accurate at waterlogged points in the case of the summer max.
 
@@
 The mire buffer had also been compared to dry soils at 10 cm depth [@RN3204; @RN3060]; our results confirm those findings at 5 cm depth and extend them to the full year.
-Thus, the pattern is reproducible among years, out to five years.
-Therefore, there exists a thermal buffer effect in mire soils that makes their temperatures less extreme than the surrounding landscape.
-In our study, this effect seems stronger during warmer periods.
-This highlights the importance of using fine-scale microclimatic data to assess vegetation responses to climate change [@RN4683; @RN4758; @RN4763].
+Also, in concordance with our results, the air temperature measured at 0.5 m from the surface of boreal Russian bogs is lower at the wetter zones, at least during the warmest months [@RN4761].
+Thus, the pattern is reproducible among years (out to five years), seems to take place both at southern and northern latitudes, and may affect not only the temperature in the soil but also air temperatures within a certain distance from the soil.
+The thermal buffer effect of mire soils that makes their temperatures less extreme than the surrounding landscape is therefore a general phenomenon.
+The existence of this buffer effect highlights the importance of using fine-scale microclimatic data to assess vegetation responses to climate change [@RN4683; @RN4758; @RN4763].
 
-One important difference with previous studies is the importance of the buffer during winter.
+One important difference with previous studies is the magnitude of the buffer during winter.
 The articles that had used model air temperatures as a control concluded that the buffer effect was stronger at the cold end of the thermal gradient, i.e.
 in winter and at night [@RN2356; @RN4675].
-In the case of this investigation, the situation was reverse: the effect was weaker when considering the minimum temperatures of the cold period.
+In the case of this investigation, the situation is the reverse: the effect is weaker when considering the minimum temperatures of the cold period.
 This indicates the importance of identifying root temperatures when working with plant communities.
 At high elevations of the study region, the soil can remain covered by snow for periods of winter, and this has its own insulating effect on soil temperatures [@RN2392].
-Indeed, snow cover has being described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760].Taking into account that the mentioned studies used a 5 cm depth for observation, it may be worth to use a deeper depth in winter to avoid this snow effect.
+Indeed, snow cover has being described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760].
+In any case, since this study was conducted at the southern limit of mire distribution, more research on the winter buffer is needed at high latitudes, since it might have a deeper biological significance there.
 
-In concordance with our results, the air temperature measured at 0.5 m from the surface of boreal bogs is lower at the wetter zones, at least during the warmest months [@RN4761].
-Thus, the water buffering effect seems to take place both at southern and northern latitudes and may affect not only the temperature in the soil but also the air temperature within certain height from the soil.
-Nonetheless, Philippov *et al.* did not record winter months, so more monitoring is needed at more northern latitudes, where the winter effect may be more relevant [@RN4675].
-
-The buffering effect was much stronger during the summer.
-This was most noticeable in the two lowest elevations, El Molinucu and La Malva.
-La Malva showed specially high summer temperatures within the dry soils (> 40 ºC).
-It must be noted that this difference of > 15ºC between the dry and waterlogged points, separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939].
+In these southern mires, the buffering effect is much stronger during the summer (Fig.
+\@ref(fig:fig5)).
+This is most noticeable in the two lowest elevations, El Molinucu and La Malva.
+La Malva shows especially high summer temperatures within the dry soils (> 40 ºC).
+It must be noted that this difference of > 15ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939].
 Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.).
 This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344].
-The buffering was also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog, i.e.
+The buffering is also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog, i.e.
 a former glacial lake undergoing silting.
-However, more studies focusing on each kind of habitat are needed in order to understand the relation between hydrology conditions and microclimatic effects.
+However, the relatively small number of sites in our study does not allow to explore further the moderating factors that may drive differences in the strength of the buffer.
+More studies focusing on each kind of habitat are needed to understand the relation between hydrology conditions and microclimatic effects.
 
 Most models currently employed to predict vegetation and species distribution use macroclimatic parameters, like the ones obtained from CHELSA [@RN4766].
-This study revealed that CHELSA values correlate relatively well with soil temperatures in the case of the minimum temperatures at the coldest month, but not with the rest of the bioclimatic variables under study.
-Moreover, CHELSA predictions were less accurate for the waterlogged soils, except for the summer max, which might be the consequence of the noise introduced by sun-heated soil.
-In adittion to the water buffering effect, other traits may explain the differences between our results and CHELSEA values like the height of temperature measurement, 5 cm below the upper layer of the soil in our study versus 2 m above the ground in CHESA; the temporal scale of observation, hourly based for a 5 years period in our study versus 30 arc sec resolution for a 34 years period; and the spatial scale, (AÑADIR).
-This brings out the importance of studying local factors as drivers of microclimatic changes [@RN4762].
+This study revealed that CHELSA values correlate relatively well with soil temperatures in the case of the minimum temperatures of the coldest month, but not with the rest of the bioclimatic variables under examination.
+Moreover, CHELSA predictions are less accurate for waterlogged soils, except for the summer max, which might be the consequence of the noise introduced by sun-heated soil.
+In addition to the water buffering effect, other features may explain the differences between our results and CHELSEA values, like the height of temperature measurement (5 cm below the upper layer of the soil in our study versus 2 m above the ground in CHELSA); the temporal scale of observation (hourly based for a 5 years period in our study versus a 34 years period); and the spatial resolution (c.
+1 m in our study versus c.
+1 km).
+Be that as it may, our results bring out the importance of considering local factors as drivers of microclimatic changes [@RN4762].
 These factors contribute to landscape heterogeneity, producing safe sites that can act as micro-refugees and buffer species from regional climatic warming.
@@
 The effect occurs during cold and warm periods, but it is stronger during the summer, at least in the study area.
-However, the climatic buffering within microrefugia would strongly rely on the water regime, a key factor for their conservation.
-Long-term decreases in precipitation could reduce the groundwater discharge into the mires, with potentially fatal consequences if the mire dries out during warmer summers.
-For the study area, the latests projections for the end of the century [@RN4939] consider a decrease of c.
-10 % in annual precipitation.
-Future recording schemes are needed to obtain local soil temperatures from other latitudes, and from more microtopographies within the same mire.
-We must consider that the microclimatic effect is not limited to mires as it can also be found in other habitats related with thawing snow like alpine and arctic-alpine slopes [@RN3279; @RN4938].
-Our study provides useful microclimate parameters to improve the current models that predict the impact of global warming on specific ecosystems.
+However, this climatic buffering within microrefugia relies strongly on the water regime, a key factor for its conservation.
+Long-term decreases in precipitation could reduce the groundwater discharge into the mires, with potentially fatal consequences if the mires dry out during warmer summers.
+Rain-fed bogs will be affected more immediately than spring fens since the latter are somewhat decoupled from short-term rainfall variation.
+However, long-term regional drying would affect all types of mires.
+For the study area, the latest projections envision a decrease of c.
+10 % in annual precipitation at the end of the century, but these projections are still subjected to a high level of uncertainty [@RN4939].
+Future recording schemes are needed to monitor the evolution of the soil thermal buffer in drying regions, to obtain local soil temperatures from other latitudes, and from more microtopographies within the same mire.
+We must also consider that the microclimatic effect is not limited to mires as it can also be found in other habitats where a gradient of moisture exists [@RN4942], such as the mosaic of microtopographies (snow patches, fellfields) that is characteristic of alpine environments [@RN3279; @RN4938].
+This study provides useful microclimate parameters to improve the current models that predict the impact of global warming on moisture-driven ecosystems.
 
@@
 
-The original data, R code for the analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/mires.
+The original data, as well as the R code for the analysis and creation of the manuscript, can be accessed at the GitHub repository https://github.com/efernandezpascual/mires.
 Upon publication, a version of record of the repository will be deposited in Zenodo.
@@
 (b) Representation of the sampling design at El Riotuertu site.
-The position of the dry and waterlogged temperature logggers is indicated, the blue area represents the extent of the waterlogged mire habitat.", fig.width=7, fig.height=3.5}
+The position of the dry and waterlogged temperature loggers is indicated, the blue area represents the extent of the waterlogged mire habitat.", fig.width=7, fig.height=3.5}
 knitr::include_graphics("../results/Fig1.png")
@@
 The 1:1 line (corresponding to no buffer effect) is shown, points falling to the right of this line show a buffer effect (i.e.
-soil temperatures at the waterlogged point are cooler than the neighnouring dry site).", fig.width=3.5, fig.height=3.5}
+soil temperatures at the waterlogged point are cooler than the neighbouring dry site).", fig.width=3.5, fig.height=3.5}
 knitr::include_graphics("../results/Fig5.png")

### 2020-08-29 doc/manuscript.Rmd
Revert "Revert "files submitted to revision""
This reverts commit 7f960cffcfd5329210f832f7d265d502dd3ac8b5.

@@
 The original data, R code for the analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/mires.
-To clean the logs we took the following steps: (i) removing records from the first week after installation, to account for the installation process and the settling of the soils; and (ii) keeping only time series with records for both the dry and waterlogged points, because some of the dataloggers had failed at different points in time.
+(The GitHub repository is private until publication, for peer-review a copy can be downloaded at https://drive.google.com/file/d/1wnrNuQXvXcCruhKXdoFaRcm6tEZPaZkN/view?usp=sharing) To clean the logs we took the following steps: (i) removing records from the first week after installation, to account for the installation process and the settling of the soils; and (ii) keeping only time series with records for both the dry and waterlogged points, because some of the dataloggers had failed at different points in time.
 Afterwards, we calculated four bioclimatic variables for each datalogger: (1) the mean diurnal range; i.e.
@@
 Upon publication, a version of record of the repository will be deposited in Zenodo.
+(The GitHub repository is private until publication, for peer-review a copy can be downloaded at https://drive.google.com/file/d/1wnrNuQXvXcCruhKXdoFaRcm6tEZPaZkN/view?usp=sharing)
 

### 2020-10-11 doc/response-reviewers2.Rmd
minor revision for IJOC

---
title: "Ref. JOC-20-0599.R1 - Response to reviewers"
author: "Eduardo Fernández-Pascual & Eva Correia-Álvarez"
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reviewer: 1

> *My main concern, which I think is possible to smooth out, is linked to the variable length of the monitored time. If there is a comparison made among the sites, this should be based on the same time intervals monitored. In the line 181, I understood that the average for the whole period was calculated. However, the periods were not the same and importantly there are differences among years and an obvious pattern in the data. Looking at Figure 2, there is a clear difference between 2015-2016 and 2019-2020 measurements from any sites (La Recoleta, La Bruxa). It is not clear how much noise this can introduce to the data and results.	I would suggest rerunning the analyses based on data from 2015-2016 only, and for 7 sites which all have the data from this period. If there is no difference, make these results available at least in Supplementary and comment on them in the text. You can also present the data as they are and base your statistical testing only on these data, as being fully comparable.*

To check the reviewer's concern, we have repeated the analysis with only the contemporaneous full-year data series (i.e. Jan 2015-Dec 2017, excluding La Vega Comeya). The output of this analysis did not differ in any significant way from the original analysis using the full datasets. We comment on this in the methods. We have also prepared a supplementary material with the results of the repeated analyses, but in our opinion this supplement adds unnecessary complexity and could be excluded from the final publication, if the editor agrees.

> *Speaking about the testing, please make sure that the data have normal distribution if T-test is your stat (it is sensitive on outliers, especially if n=8). This should be added in the Methods and clearly stated that there was a normal error structure. Otherwise, you better using nonparametric stat as Wilcoxon pair test. The same for Pearson, but the point distribution is shown in Fig. 4 suggests it is okay.*

We have checked whether data meet the normality assumption using the Shapiro-Wilk Normality Test, the assumption was met in all cases. This is now indicated in the text.

> *I am not sure if the statements provided in the lines 208-210 about the wider range and higher maximums were validated based on any stats.*

These statements are validated by the t-tests, described in following paragraphs.

> *I am not sure if the explanation given at 255-259 does not have a flaw in its logic. Maybe I did not get it right, but for me, there is a difference between a gradient and seasonal or diurnal variation. In this part, you are mixing to levels of variation – among sites (differences between sites – a climatic gradient) and within a site, variation over the season or day at the sites. Maybe the problem is only in the wording “the cold end of the thermal gradient”. It confuses me if you call winter and night as the cold end of the gradient. Would it be possible to rephrase this?*

We have substituted the word *gradient* with the word *spectrum*.

> *As there are only 8 values behind each bar in Fig. 3, it might have the sense to present the values themselves. You can then connect individual pairs by a line. Please also add significant above each comparison.*

The change proposed would make Table 2 redundant, and we would prefer to keep it, showing the actual numbers. Since these are the effect sizes in ºC, looking at them gives a quick impression of the strength of the buffer at each site and for each bioclimatic variable.

> *I assume that Patrick A Raney or Patrick A. Raney should be referred from the text as Patrick. It appears at many places in the text, as likely an error of the reference manager.*

Indeed, it was a problem with the reference manager, and should be corrected now.

> *Line 271: missing space at 15 °C (15°C)*

Corrected.

# Reviewer: 2

> *Line 108 – wouldn’t it depend on whether the region is becoming wetter or drier the amount of water being delivered to fens whether they were vulnerable to drying?*

We have changed the sentence to *Global warming would also result in warmer groundwater delivered to fens, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires*.

> *Lines 173-177 –I’d recommend deleting line 173-176 removing the temporary link for peer review to save space in the final manuscript. You can also delete the git-hub reference as you already provide data availability in lines 329.*

We removed these lines.

> *Line 304: micro-refugees  has an extra ‘e’ that should be deleted.*

Corrected.

### 2020-10-11 doc/Supplementary.Rmd
minor revision for IJOC

---
title: "Supporting information for Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils"
output: 
  bookdown::word_document2:
    reference_docx: ijctemplate.docx
bibliography: refs.bib
biblio-style: international-journal-of-climatology.csl
link-citations: yes
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(tidyverse); library(flextable); library(lubridate)
load("../results/logs.RData")

# 2015-2016 only for analysis

temperatures %>%
  filter(Time >= as.Date("2015-01-01")) %>%
  filter(Time <= as.Date("2017-12-31")) %>%
  filter(! Logger %in% c("A511BE", "A511C1")) -> temperatures

# Calculations

## Bioclimatic variables as defined by US Geological Survey, WorlClim

### Bio 2 Annual Mean Diurnal Range

temperatures %>%
  group_by(Logger, Day = floor_date(Time, "day")) %>%
  summarise(Tmax = max(Temperature), 
            Tmin = min(Temperature),
            Tdif = Tmax - Tmin) %>%
  group_by(Logger) %>%
  summarise(`Diurnal range` = mean(Tdif)) -> bio2

### Bio 5 Max Temperature of Warmest Month

temperatures %>%
  group_by(Logger, Day = floor_date(Time, "day")) %>%
  summarise(Tmax = max(Temperature)) %>%
  mutate(Month = month(Day)) %>%
  group_by(Logger, Month) %>%
  summarise(`Summer max` = mean(Tmax)) %>%
  group_by(Logger) %>%
  filter(`Summer max` == max(`Summer max`)) %>%
  select(-Month) -> bio5

### Bio 6 Min Temperature of the Coldest Month

temperatures %>%
  group_by(Logger, Day = floor_date(Time, "day")) %>%
  summarise(Tmin = min(Temperature)) %>%
  mutate(Month = month(Day)) %>%
  group_by(Logger, Month) %>%
  summarise(`Winter min` = mean(Tmin)) %>%
  group_by(Logger) %>%
  filter(`Winter min` == min(`Winter min`)) %>%
  select(-Month) -> bio6

### Merge and calculate Bio 7 Annual Temperature Range

bio2 %>%
  merge(bio5) %>%
  merge(bio6) %>%
  mutate(`Annual range` = 
           `Summer max` -
           `Winter min`) -> bioclimatics

# Tables

## Paired T-tests

t.testEFP <- function(df) 
{
  t <- t.test(df$Waterlogged, df$Dry, alternative = "less", paired = TRUE)
  data.frame(t$statistic,
             t$p.value,
             t$estimate,
             t$conf.int[1],
             t$conf.int[2]
  )
}

t.testEFPgreater <- function(df) 
{
  t <- t.test(df$Waterlogged, df$Dry, alternative = "greater", paired = TRUE)
  data.frame(t$statistic,
             t$p.value,
             t$estimate,
             t$conf.int[1],
             t$conf.int[2]
  ) # In the case of the minimum temperature is the opposite hypothesis
}

rbind(
  merge(header[, 1:8], bioclimatics, by = "Logger") %>% 
    gather(Trait, Value, `Diurnal range`:`Annual range`) %>%
    dplyr::select(Site, Groundwater, Value, Trait, Value) %>%
    spread(Groundwater, Value) %>%
    filter(! Trait %in% "Winter min") %>%
    group_by(Trait) %>%
    do(t.testEFP(.)),
  merge(header[, 1:8], bioclimatics, by = "Logger") %>% 
    gather(Trait, Value, `Diurnal range`:`Annual range`) %>%
    dplyr::select(Site, Groundwater, Value, Trait, Value) %>%
    spread(Groundwater, Value) %>%
    filter(Trait %in% "Winter min") %>%
    group_by(Trait) %>%
    do(t.testEFPgreater(.))) -> ttests

## Table 3 - Chelsa

bioclimatics %>%
  gather(Variable, Value, `Diurnal range`:`Annual range`) %>%
  mutate(Source = "Loggers") -> bioLoggers

header %>%
  select(Logger, `Diurnal range`:`Annual range`) %>%
  gather(Variable, Value, `Diurnal range`:`Annual range`) %>%
  mutate(Source = "CHELSA") -> bioCHELSA

header %>%
  select(Site, Logger, Groundwater, Mire) %>%
  merge(rbind(bioLoggers, bioCHELSA), by = "Logger") %>%
  spread(Source, Value) %>%
  na.omit -> bio

RMSE <- function(df) {
  require(Metrics)
  data.frame(RMSE = rmse(df$Loggers, df$CHELSA))}

CORR <- function(df) {
  c <- cor.test(df$Loggers, df$CHELSA)
  data.frame(t = c$statistic,
             df = c$parameter,
             p = c$p.value,
             r = c$estimate)}

bio %>%
  group_by(Variable, Groundwater) %>%
  do(CORR(.)) -> corrtable # Pearson's correlation

bio %>%
  group_by(Variable, Groundwater) %>%
  do(RMSE(.)) -> rmsetable # Root-mean-square error (RMSE)

merge(corrtable, rmsetable) %>%
  mutate(RMSE = round(RMSE, 2),
         r = round(r, 2),
         t = round(t, 3),
         p = round(p, 3)) %>%
  rename(`RMSE (ºC)` = RMSE,
         `Pearson's r` = r) -> chelsa
```

```{r table1, echo = FALSE}
regulartable(ttests) %>% width(width = 1) %>% set_caption("Table S1: Results of the t-tests comparing the paired waterlogged-dry soil bioclimatic variables. Analysis performed with full-year, contemporaneous data series (2015-2017).")
```

```{r table3, echo = FALSE}
regulartable(chelsa) %>% width(width = 1) %>% set_caption("Table S2: Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation. RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy). Analysis performed with full-year, contemporaneous data series (2015-2017).")
```

### 2020-10-11 doc/manuscript.Rmd
minor revision for IJOC

--- 2020-08-29 doc/manuscript.Rmd
@@
 Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935].
-Global warming would also result in warmer groundwater delivered to fens and in drier conditions, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].
+Global warming would also result in warmer groundwater delivered to fens, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].
 
@@
 These measures can be used to support previous evidence about the general effects of soil moisture on thermal buffering [@RN3204; @RN2356; @RN4675; @RN4936] and to determine the magnitude of this buffering within microrefugia habitats.
-Specifically, we tested the hypotheses that, when compared to adjacent dry soils, waterlogged mire soil are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales.
+Specifically, we tested the hypotheses that, when compared to adjacent dry soils, waterlogged mire soils are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales.
 In addition, we compared *in situ* measurements with data derived from the CHELSA climatic models [@RN4766].
@@
 The present humid and mild climate of this region harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246].
-Rain-fed raised bogs (https://eunis.eea.europa.eu/habitats/260) are very rare and appear locally on raised hummocks of acid valley mires (https://eunis.eea.europa.eu/habitats/526) which be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes.
+Rain-fed raised bogs (https://eunis.eea.europa.eu/habitats/260) are very rare and appear locally on raised hummocks of acid valley mires (https://eunis.eea.europa.eu/habitats/526) which can be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes.
 Glacial lakes undergoing silting develop transition mires and quaking bogs communities (https://eunis.eea.europa.eu/habitats/1637) in the water-to-land transition.
@@
 All data processing and analysis was performed in R [@RN2315].
-The original data, R code for the analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/mires.
 To clean the logs we took the following steps: (i) removing records from the first week after installation, to account for the installation process and the settling of the soils; and (ii) keeping only time series with records for both the dry and waterlogged points, because some of the dataloggers had failed at different points in time.
@@
 We calculated RMSE using its implementation in the package *Metrics* [@RN4941].
+We checked whether our data complied with the normality assumption of the t-test and Pearson's correlation using the Shapiro-Wilk normality test; the assumption was met in all cases (p > 0.05).
+To check whether the analyses were influenced by the fact that recording periods did not match across all eight sites, we repeated the analyses using only full-year, contemporaneous data series (01 Jan 2015 - 31 Dec 2017, excluding La Vega Comeya); the results of these analyses were the same as the ones presented in this article (Supplementary Tables S1 and S2).
 
@@
 One important difference with previous studies is the magnitude of the buffer during winter.
-The articles that had used model air temperatures as a control concluded that the buffer effect was stronger at the cold end of the thermal gradient, i.e.
+The articles that had used model air temperatures as a control concluded that the buffer effect was stronger at the cold end of the thermal spectrum, i.e.
 in winter and at night [@RN2356; @RN4675].
@@
 La Malva shows especially high summer temperatures within the dry soils (> 40 ºC).
-It must be noted that this difference of > 15ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939].
+It must be noted that this difference of > 15 ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939].
 Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.).
@@
 Be that as it may, our results bring out the importance of considering local factors as drivers of microclimatic changes [@RN4762].
-These factors contribute to landscape heterogeneity, producing safe sites that can act as micro-refugees and buffer species from regional climatic warming.
+These factors contribute to landscape heterogeneity, producing safe sites that can act as micro-refuges and buffer species from regional climatic warming.
 
@@
 Upon publication, a version of record of the repository will be deposited in Zenodo.
+
+# Supporting information
+
+*Tables S1-S2* Results of the statistical analyses, repeated using only the contemporaneous full-year data series.
 
@@
 ```{r table3, echo = FALSE}
-regulartable(table3) %>% width(width = 1) %>% set_caption("Table 3: Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation.
+regulartable(table3) %>% width(width = 1) %>% set_caption("Table 3: Comparison of soil bioclimatic variables with CHELSA air temperatures, per bioclimatic variable and groundwater situation.
 RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy).")

### 2020-10-11 src/analysis 2015-2017.R

@@ -2,6 +2,19 @@ library(tidyverse); library(lubridate)
 
 load("results/logs.RData")
 
+# Full records for plotting
+
+merge(temperatures, header) %>%
+  arrange(Elevation, Site) %>%
+  mutate(Facet = paste(Elevation, Mire, sep = " m a.s.l. - ")) -> recordsdf
+
+# 2015-2016 only for analysis
+
+temperatures %>%
+  filter(Time >= as.Date("2015-01-01")) %>%
+  filter(Time <= as.Date("2017-12-31")) %>%
+  filter(! Logger %in% c("A511BE", "A511C1")) -> temperatures
+
 # Calculations
 
 ## Bioclimatic variables as defined by US Geological Survey, WorlClim
@@ -117,13 +130,14 @@ rbind(
     group_by(Trait) %>%
     do(t.testEFPgreater(.))) -> ttests
 
+
 #----------------------------------------------------------------------------
 
 # Tables
 
 ## Table 1 - Study sites
 
-temperatures %>%
+recordsdf %>%
   group_by(Logger) %>%
   summarise(Start = min(Time),
             End = max(Time),
@@ -178,7 +192,8 @@ header %>%
 header %>%
   select(Site, Logger, Groundwater, Mire) %>%
   merge(rbind(bioLoggers, bioCHELSA), by = "Logger") %>%
-  spread(Source, Value) -> bio
+  spread(Source, Value) %>%
+  na.omit -> bio
 
 RMSE <- function(df) {
   require(Metrics)
@@ -275,7 +290,7 @@ ggplotGrob(
     theme(panel.background = element_rect(color = "grey96", fill = "grey96"),
           legend.position = "top", legend.title = element_blank(),
           panel.border = element_rect(colour = "black", fill = NA, size = 1))) -> g2
- 
+
 ### Put together
 
 g1 +
@@ -549,17 +564,17 @@ peakdf %>%
 
 save(ttests, table1, table2, table3, file = "results/MSoutput.RData") # Save numerical results and tables to build manuscript
 
-ggsave(plot1, file = "results/Fig1.tiff", 
-       path = NULL, scale = 1, width = 170, height = 85, units = "mm", dpi = 300) # Save figures to build manuscript
+ggsave(plot1, file = "results/Fig1.png", 
+       path = NULL, scale = 1, width = 170, height = 85, units = "mm", dpi = 600) # Save figures to build manuscript
 
[…]

### 2020-10-11 doc/Supplementary.Rmd
accepted version

@@
 title: "Supporting information for Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils"
-output: 
-  bookdown::word_document2:
-    reference_docx: ijctemplate.docx
+output: github_document
 bibliography: refs.bib
@@
 ```{r message = FALSE, echo = FALSE, warning = FALSE}
-library(tidyverse); library(flextable); library(lubridate)
+library(tidyverse); library(lubridate)
 load("../results/logs.RData")
@@
 ```{r table1, echo = FALSE}
-regulartable(ttests) %>% width(width = 1) %>% set_caption("Table S1: Results of the t-tests comparing the paired waterlogged-dry soil bioclimatic variables.
-Analysis performed with full-year, contemporaneous data series (2015-2017).")
+ttests %>% kableExtra::kable(caption = "**Table S1** *Results of the t-tests comparing the paired waterlogged-dry soil bioclimatic variables.
+Analysis performed with full-year, contemporaneous data series (2015-2017).*", digits = 3, format = "pandoc")
 ```
@@
 ```{r table3, echo = FALSE}
-regulartable(chelsa) %>% width(width = 1) %>% set_caption("Table S2: Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation.
+chelsa %>% kableExtra::kable(caption = "**Table S2** *Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation.
 RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy).
-Analysis performed with full-year, contemporaneous data series (2015-2017).")
+Analysis performed with full-year, contemporaneous data series (2015-2017).*", format = "pandoc")
 ```

### 2020-10-11 doc/manuscript.Rmd
accepted version

@@
 subtitle: "Short title: Testing the thermal buffer of mire groundwater"
-output: 
-  bookdown::word_document2:
-    reference_docx: ijctemplate.docx
+output: github_document
 bibliography: refs.bib
@@
 ```{r message = FALSE, echo = FALSE, warning = FALSE}
-library(tidyverse); library(flextable)
+library(tidyverse)
 load("../results/logs.RData")
@@
 Telephone: +34985104787.
-
-# Financial support
-
-E.F.P. received financial support from the Government of Asturias and the FP7 – Marie Curie - COFUND programme of the European Commission (Grant ‘Clarín’ ACB17-19).
 
@@
 
-This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W) (Fig.
-\@ref(fig:fig1)a).
+This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W) (**Fig.
+1a**).
 Local geography and climate are dominated by the Cantabrian Mountains (> 1500 m above sea level), which run parallel to the coast and trap the prevailing NW Atlantic winds.
@@
 
-We selected `r length(unique(temperatures$Logger))/2` mire sites representing the regional elevation gradient of mire vegetation and the different mire types (Table 1).
+```{r fig1, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 1** **(a)** *Location of the study sites in northern Spain.* **(b)** *Representation of the sampling design at El Riotuertu site.
+The position of the dry and waterlogged temperature loggers is indicated, the blue area represents the extent of the waterlogged mire habitat.*", fig.width=7, fig.height=3.5}
+knitr::include_graphics("../results/Fig1.png")
+```
+
+We selected `r length(unique(temperatures$Logger))/2` mire sites representing the regional elevation gradient of mire vegetation and the different mire types (**Table 1**).
 Although we classify two of these sites as rain-fed bogs based on their vegetation (*Oxycocco-Sphagnetea* Br.-Bl.
@@
 In this ecosystem, it is at this depth that true soil begins to develop, under the porous upper layers made up of live mosses.
-We installed one datalogger in a flat waterlogged spot within the mire; the other one in the close vicinity, but in a flat and dry upland area outside the mire (Fig.
-\@ref(fig:fig1)b).
+We installed one datalogger in a flat waterlogged spot within the mire; the other one in the close vicinity, but in a flat and dry upland area outside the mire (**Fig.
+1b**).
 The vegetation at the sampling points was always either mire or pasture, with no shrubs, trees or any other landscape features shading the measuring points.
 Dataloggers recorded temperature once every hour and stayed on site for approximately five years, after which we retrieved them and downloaded their records.
-At the moment of retrieval, the internal clock of all dataloggers had not deviated for more than four hours.
+At the moment of retrieval, the internal clock of all dataloggers had not deviated for more than four hours
+
+```{r table1, echo = FALSE}
+table1 %>% kableExtra::kable(caption = "**Table 1** *Mire sites included in this study, indicating the type of fen, pH, conductivity, elevation, coordinates, and length of the temperature recording period.
+Groundwater pH and conductivity values are the average of 5–10 measures per site taken with a MM40 Portable Multimeter (Crison Instruments S.A., Alella, Spain).
+Coordinates are in decimal degrees WGS84.*", format = "pandoc")
+```
 
@@
 We checked whether our data complied with the normality assumption of the t-test and Pearson's correlation using the Shapiro-Wilk normality test; the assumption was met in all cases (p > 0.05).
-To check whether the analyses were influenced by the fact that recording periods did not match across all eight sites, we repeated the analyses using only full-year, contemporaneous data series (01 Jan 2015 - 31 Dec 2017, excluding La Vega Comeya); the results of these analyses were the same as the ones presented in this article (Supplementary Tables S1 and S2).
+To check whether the analyses were influenced by the fact that recording periods did not match across all eight sites, we repeated the analyses using only full-year, contemporaneous data series (01 Jan 2015 - 31 Dec 2017, excluding La Vega Comeya); the results of these analyses were the same as the ones presented in this article (Supplementary **Tables S1** and **S2**).
 
@@
 Dataloggers recorded temperatures for five years in five of the sites, four years in two, and two years in one.
-Time series showed considerably less variation at waterlogged points, for all eight site comparisons (Fig.
-\@ref(fig:fig2)).
+Time series showed considerably less variation at waterlogged points, for all eight site comparisons (**Fig.
+2**).
 
-Bioclimatic variables (Table 2 and Fig.
-\@ref(fig:fig3)) supported this notion.
+```{r fig2, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 2** *Hourly soil temperature records at the mire sites.
+The blue series was recorded within the mire, in a waterlogged area.
+The red series was recorded in a neighbouring dry area.
+Dataloggers were buried at 5 cm depth.
+The amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.*", fig.width=7, fig.height=10}
+knitr::include_graphics("../results/Fig2.png")
+```
+
+Bioclimatic variables (**Table 2** and **Fig.
+3**) supported this notion.
 The mean annual range was wider at dry points, as was the mean diurnal range.
 Maximum temperatures of the warmest months were usually higher at dry measuring points.
-The opposite was true for the minimums of the coldest months, in which case temperature was generally colder at dry points, although the difference was less pronounced than for the maximums (Fig.
-\@ref(fig:fig3)).
+The opposite was true for the minimums of the coldest months, in which case temperature was generally colder at dry points, although the difference was less pronounced than for the maximums (**Fig.
+3**).
+
+```{r fig3, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 3** *Average bioclimatic variables in the dry and waterlogged points.
+The bars represent the mean value, and the brackets the standard error of 8 records.*", fig.width=7, fig.height=3}
+knitr::include_graphics("../results/Fig3.png")
+```
 
 Patterns regarding the maximum temperatures and annual range were especially noticeable at both low (El Molinucu, La Malva) and high (La Recoleta) elevations.
-The pattern for the diurnal range, however, was more prominent just at the low sites (El Molinucu, La Malva), whereas the minimums of the coldest months showed no specific pattern (Table 2).
+The pattern for the diurnal range, however, was more prominent just at the low sites (El Molinucu, La Malva), whereas the minimums of the coldest months showed no specific pattern (**Table 2**).
 
@@
 
-The CHELSA climatic model (Table 3 and Fig.
-\@ref(fig:fig4)) provided air temperatures that significantly correlated (p < 0.05) with soil temperatures only in the case of the minimums of the coldest month, both at dry points (Pearson's r = 0.90) and waterlogged points (Pearson's r  = 0.84).
+```{r table2, echo = FALSE}
+table2 %>% kableExtra::kable(caption = "**Table 2** *Groundwater buffer effect per mire and bioclimatic variable.
+The buffer effect was calculated as the difference between the value in the waterlogged and the dry points.*", format = "pandoc")
+```
+
+The CHELSA climatic model (**Table 3** and **Fig.
+4**) provided air temperatures that significantly correlated (p < 0.05) with soil temperatures only in the case of the minimums of the coldest month, both at dry points (Pearson's r = 0.90) and waterlogged points (Pearson's r  = 0.84).
 The correlation between CHELSA and soil values was marginally significant (p = 0.05) in the case of the maximums of the warmest month, but only at dry points (Pearson's r = 0.71).
@@
 RMSE indicated that the predictions of CHELSA were more accurate at dry points for the cases of the annual range, the diurnal range and the winter min; the CHELSA values were more accurate at waterlogged points in the case of the summer max.
+
+```{r table3, echo = FALSE}
+table3 %>% kableExtra::kable(caption = "**Table 3** *Comparison of soil bioclimatic variables with CHELSA air temperatures, per bioclimatic variable and groundwater situation.
+RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy).*", format = "pandoc")
+```
+
+```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 4** *Scatter plots of the bioclimatic variables predicted by CHELSA air temperatures vs.
+soil temperatures measured in situ, in dry and waterlogged points.*", fig.width=7, fig.height=7}
+knitr::include_graphics("../results/Fig4.png")
+```
 
@@
 
-In these southern mires, the buffering effect is much stronger during the summer (Fig.
-\@ref(fig:fig5)).
+In these southern mires, the buffering effect is much stronger during the summer (**Fig.
+5**).
 This is most noticeable in the two lowest elevations, El Molinucu and La Malva.
@@
 More studies focusing on each kind of habitat are needed to understand the relation between hydrology conditions and microclimatic effects.
+
+```{r fig5, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 5** *Complementary temporal analysis of the groundwater buffer effect during peak hot events.
+Variation of soil temperatures at paired sites (dry vs.
+waterlogged points) during exceptionally warm days of the warm season.
+Exceptionally warm days were considered those with a daily maximum (at the dry point) belonging to the 90th percentile of daily maximums for the site.
+The 1:1 line (corresponding to no buffer effect) is shown, points falling to the right of this line show a buffer effect (i.e.
+soil temperatures at the waterlogged point are cooler than the neighbouring dry site).*", fig.width=3.5, fig.height=3.5}
+knitr::include_graphics("../results/Fig5.png")
+```
 
@@
 
+# Acknowledgements
+
+E.F.P. received financial support from the Government of Asturias and the FP7 – Marie Curie - COFUND programme of the European Commission (Grant ‘Clarín’ ACB17-19).
+
 # Data availability
@@
 :::
-
-# Tables
-
-```{r table1, echo = FALSE}
-regulartable(table1) %>% width(width = 1) %>% set_caption("Table 1: Mire sites included in this study, indicating the type of fen, pH, conductivity, elevation, coordinates, and length of the temperature recording period.
-Groundwater pH and conductivity values are the average of 5–10 measures per site taken with a MM40 Portable Multimeter (Crison Instruments S.A., Alella, Spain).
-Coordinates are in decimal degrees WGS84.")
-```
-
-```{r table2, echo = FALSE}
-regulartable(table2) %>% width(width = 1) %>% set_caption("Table 2: Groundwater buffer effect per mire and bioclimatic variable.
-The buffer effect was calculated as the difference between the value in the waterlogged and the dry points.")
-```
-
-```{r table3, echo = FALSE}
-regulartable(table3) %>% width(width = 1) %>% set_caption("Table 3: Comparison of soil bioclimatic variables with CHELSA air temperatures, per bioclimatic variable and groundwater situation.
-RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy).")
-```
-
-# Figures
-
-```{r fig1, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "(a) Location of the study sites in northern Spain.
-(b) Representation of the sampling design at El Riotuertu site.
-The position of the dry and waterlogged temperature loggers is indicated, the blue area represents the extent of the waterlogged mire habitat.", fig.width=7, fig.height=3.5}
-knitr::include_graphics("../results/Fig1.png")
-```
-
-```{r fig2, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites.
-The blue series was recorded within the mire, in a waterlogged area.
-The red series was recorded in a neighbouring dry area.
-Dataloggers were buried at 5 cm depth.
-The amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.", fig.width=7, fig.height=10}
-knitr::include_graphics("../results/Fig2.png")
-```
-
-```{r fig3, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Average bioclimatic variables in the dry and waterlogged points.
-The bars represent the mean value, and the brackets the standard error of 8 records.", fig.width=7, fig.height=3}
-knitr::include_graphics("../results/Fig3.png")
-```
-
-```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Scatter plots of the bioclimatic variables predicted by CHELSA air temperatures vs.
-soil temperatures measured in situ, in dry and waterlogged points.", fig.width=7, fig.height=7}
-knitr::include_graphics("../results/Fig4.png")
-```
-
-```{r fig5, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Complementary temporal analysis of the groundwater buffer effect during peak hot events.
-Variation of soil temperatures at paired sites (dry vs.
-waterlogged points) during exceptionally warm days of the warm season.
-Exceptionally warm days were considered those with a daily maximum (at the dry point) belonging to the 90th percentile of daily maximums for the site.
-The 1:1 line (corresponding to no buffer effect) is shown, points falling to the right of this line show a buffer effect (i.e.
-soil temperatures at the waterlogged point are cooler than the neighbouring dry site).", fig.width=3.5, fig.height=3.5}
-knitr::include_graphics("../results/Fig5.png")
-```

### 2020-10-11 src/analysis 2015-2017.R

@@ -1,580 +1,0 @@
-library(tidyverse); library(lubridate)
-
-load("results/logs.RData")
-
-# Full records for plotting
-
-merge(temperatures, header) %>%
-  arrange(Elevation, Site) %>%
-  mutate(Facet = paste(Elevation, Mire, sep = " m a.s.l. - ")) -> recordsdf
-
-# 2015-2016 only for analysis
-
-temperatures %>%
-  filter(Time >= as.Date("2015-01-01")) %>%
-  filter(Time <= as.Date("2017-12-31")) %>%
-  filter(! Logger %in% c("A511BE", "A511C1")) -> temperatures
-
-# Calculations
-
-## Bioclimatic variables as defined by US Geological Survey, WorlClim
-
-### Bio 2 Annual Mean Diurnal Range
-
-temperatures %>%
-  group_by(Logger, Day = floor_date(Time, "day")) %>%
-  summarise(Tmax = max(Temperature), 
-            Tmin = min(Temperature),
-            Tdif = Tmax - Tmin) %>%
-  group_by(Logger) %>%
-  summarise(`Diurnal range` = mean(Tdif)) -> bio2
-
-### Bio 5 Max Temperature of Warmest Month
-
-temperatures %>%
-  group_by(Logger, Day = floor_date(Time, "day")) %>%
-  summarise(Tmax = max(Temperature)) %>%
-  mutate(Month = month(Day)) %>%
-  group_by(Logger, Month) %>%
-  summarise(`Summer max` = mean(Tmax)) %>%
-  group_by(Logger) %>%
-  filter(`Summer max` == max(`Summer max`)) %>%
-  select(-Month) -> bio5
-
-### Bio 6 Min Temperature of the Coldest Month
-
-temperatures %>%
-  group_by(Logger, Day = floor_date(Time, "day")) %>%
-  summarise(Tmin = min(Temperature)) %>%
-  mutate(Month = month(Day)) %>%
-  group_by(Logger, Month) %>%
-  summarise(`Winter min` = mean(Tmin)) %>%
-  group_by(Logger) %>%
-  filter(`Winter min` == min(`Winter min`)) %>%
-  select(-Month) -> bio6
-
-### Merge and calculate Bio 7 Annual Temperature Range
-
-bio2 %>%
-  merge(bio5) %>%
-  merge(bio6) %>%
-  mutate(`Annual range` = 
-           `Summer max` -
-           `Winter min`) -> bioclimatics
-
[…]

### 2020-10-11 src/analysis.R

@@ -549,17 +549,17 @@ peakdf %>%
 
 save(ttests, table1, table2, table3, file = "results/MSoutput.RData") # Save numerical results and tables to build manuscript
 
-ggsave(plot1, file = "results/Fig1.tiff", 
+ggsave(plot1, file = "results/Fig1.png", 
        path = NULL, scale = 1, width = 170, height = 85, units = "mm", dpi = 300) # Save figures to build manuscript
 
-ggsave(plot2, file = "results/Fig2.tiff", 
+ggsave(plot2, file = "results/Fig2.png", 
        path = NULL, scale = 1, width = 170, height = 250, units = "mm", dpi = 300)
 
-ggsave(plot3, file = "results/Fig3.tiff", 
+ggsave(plot3, file = "results/Fig3.png", 
        path = NULL, scale = 1, width = 170, height = 70, units = "mm", dpi = 300)
 
-ggsave(plot4, file = "results/Fig4.tiff", 
+ggsave(plot4, file = "results/Fig4.png", 
        path = NULL, scale = 1, width = 170, height = 170, units = "mm", dpi = 300)
 
-ggsave(plot5, file = "results/Fig5.tiff", 
+ggsave(plot5, file = "results/Fig5.png", 
        path = NULL, scale = 1, width = 100, height = 100, units = "mm", dpi = 300)

### 2020-10-11 README.Rmd
Revert "Revert "readme picture""
This reverts commit 3dbc162c5e3a7bbd5851d52ce93ff214dbafaa5d.
updated links

@@
 
-![Alkaline fen in the Cantabrian Mountains](data/Sierra Cebolleda.JPG)
+![Alkaline fen in the Cantabrian Mountains](data/Sierra Cebolleda.jpg)
 
@@
 
-This repository stores all the information related to the publication *Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils*, including the raw datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the manuscript.
+This repository stores all the information related to the publication *Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils*, including the raw datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the [manuscript](https://github.com/efernandezpascual/mires/blob/master/doc/manuscript.md).
 
@@
 * `results` Output of the `R` scripts, including clean data and figures.
-* `src` Scripts in `R` language used to clean the raw data and perform the analyses of the manuscript.
+* `src` Scripts in `R` language used to clean the raw data and perform the analyses of the [manuscript](https://github.com/efernandezpascual/mires/blob/master/doc/manuscript.md).
 
@@
 
-Please cite the dataset as `CITATION`
+Please cite any use of this repository as `Fernández-Pascual, E.
+& Correia-Álvarez., E.
+(2021).
+Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils.
+International Journal of Climatology, in press.`

### 2020-10-11 doc/manuscript.Rmd
updated links

@@
 We checked whether our data complied with the normality assumption of the t-test and Pearson's correlation using the Shapiro-Wilk normality test; the assumption was met in all cases (p > 0.05).
-To check whether the analyses were influenced by the fact that recording periods did not match across all eight sites, we repeated the analyses using only full-year, contemporaneous data series (01 Jan 2015 - 31 Dec 2017, excluding La Vega Comeya); the results of these analyses were the same as the ones presented in this article (Supplementary **Tables S1** and **S2**).
+To check whether the analyses were influenced by the fact that recording periods did not match across all eight sites, we repeated the analyses using only full-year, contemporaneous data series (01 Jan 2015 - 31 Dec 2017, excluding La Vega Comeya); the results of these analyses were the same as the ones presented in this article ([Supplementary **Tables S1** and **S2**](https://github.com/efernandezpascual/mires/blob/master/doc/supplementary.md)).
 
@@
 
-*Tables S1-S2* Results of the statistical analyses, repeated using only the contemporaneous full-year data series.
+[**Tables S1-S2**](https://github.com/efernandezpascual/mires/blob/master/doc/supplementary.md) Results of the statistical analyses, repeated using only the contemporaneous full-year data series.
 

### 2020-10-12 doc/manuscript.Rmd
minor address change
zenodo doi

@@
 title: "Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils"
-subtitle: "Short title: Testing the thermal buffer of mire groundwater"
 output: github_document
@@
 
-Eduardo Fernández-Pascual^1^ ^+^, Eva Correia-Álvarez^2^
+Eduardo Fernández-Pascual^1^ ^+^, Eva Correia-Álvarez^1^
 
-^1^ Universidad de Oviedo, ^2^ Independent researcher
+^1^ Universidad de Oviedo
 
@@
 
-# Keywords
-
-bog, CHELSA, climatic model, datalogger, fen, peatland, soil temperature, wetland
+**Keywords:** bog, CHELSA, climatic model, datalogger, fen, peatland, soil temperature, wetland
 
@@
 The original data, as well as the R code for the analysis and creation of the manuscript, can be accessed at the GitHub repository https://github.com/efernandezpascual/mires.
-Upon publication, a version of record of the repository will be deposited in Zenodo.
+A version of record of the repository can be found at https://doi.org/10.5281/zenodo.4081418.
 

### 2020-10-13 doc/manuscript.Rmd
minor ms corrections

@@
 
-Ecosystems adapt differently to global warming through microclimate factors.
-Mires are sensitive wetland habitats that strongly rely on local soil properties, which makes them a good model to understand how local climate parameters counteract the effects of climate change.
+Ecosystems adapt differently to global warming through microclimatic factors.
+Mires are sensitive wetland habitats that strongly rely on local soil properties, which makes them a good model to understand how local climatic parameters counteract the effects of climate change.
 We quantified the temperature buffering effect in waterlogged mire soils as compared with adjacent dry soils.
@@
 Waterlogged soils had less diurnal thermal amplitude (-2.3 ºC), less annual thermal amplitude (-5.1 ºC), cooler summer maximums (-4.3 ºC) and warmer winter minimums (+0.8 ºC).
-CHELSA air temperatures only correlated significantly (p < 0.05) with the winter minimum temperatures (Pearson's r > 0.83), and CHELSA predictions were less accurate (higher RMSE) for waterlogged soils, except for the summer maximums.
+CHELSA air temperatures only correlated significantly (p < 0.05) with winter minimum soil temperatures (Pearson's r > 0.83), and CHELSA predictions were less accurate (higher RMSE) for waterlogged soils, except for the summer maximums.
 
@@
 Since root-zone temperature is a major determinant of plant ecophysiology [@RN3024], the groundwater buffer effect is expected to allow mire plants to live in a wider range of air temperatures than they could otherwise.
-Indeed, mires have a relatively homogeneous flora across their range despite their broad distribution.
+Indeed, mires have a relatively homogeneous flora across their range, despite their broad distribution.
 In Europe, fens are distributed from the Iberian Peninsula to boreal Fennoscandia and from low valleys to the alpine belt [@RN4678].
@@
 It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change [@RN4936].
-Both bogs and fens are actively peat forming and are dependent on precipitation, whereas fens rely also on sources of telluric water from mineral ground.
+Both bogs and fens are actively peat forming and depend on precipitation, but fens rely also on sources of telluric water from mineral ground.
 Moreover, they show a mosaic of microreliefs and a specific plant cover zonation which makes them especially sensitive to small changes in wetness.
@@
 Moreover, CHELSA predictions are less accurate for waterlogged soils, except for the summer max, which might be the consequence of the noise introduced by sun-heated soil.
-In addition to the water buffering effect, other features may explain the differences between our results and CHELSEA values, like the height of temperature measurement (5 cm below the upper layer of the soil in our study versus 2 m above the ground in CHELSA); the temporal scale of observation (hourly based for a 5 years period in our study versus a 34 years period); and the spatial resolution (c.
+In addition to the water buffering effect, other features may explain the differences between our results and CHELSEA values, like the height of temperature measurement (5 cm below the upper layer of the soil in our study versus 2 m above the ground in CHELSA); the temporal scale of observation (hourly based for a 5-year period in our study versus a 34-year period); and the spatial resolution (c.
 1 m in our study versus c.

### 2020-10-13 src/analysis.R

@@ -162,7 +162,8 @@ merge(header[, 1:10], bioclimatics) %>%
   merge(header[, c(3, 10)], by = "Site") %>%
   arrange(Elevation) %>%
   select(Site, `Annual range`:`Winter min`) %>%
-  unique -> table2 
+  unique %>%
+  remove_rownames() -> table2 
 
 ## Table 3 - Chelsa
 

### 2020-10-26 doc/manuscript.Rmd
corrected typos

@@
 Moreover, they show a mosaic of microreliefs and a specific plant cover zonation which makes them especially sensitive to small changes in wetness.
-Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935].
+Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity of organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935].
 Global warming would also result in warmer groundwater delivered to fens, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].
@@
 At high elevations of the study region, the soil can remain covered by snow for periods of winter, and this has its own insulating effect on soil temperatures [@RN2392].
-Indeed, snow cover has being described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760].
+Indeed, snow cover has been described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760].
 In any case, since this study was conducted at the southern limit of mire distribution, more research on the winter buffer is needed at high latitudes, since it might have a deeper biological significance there.
@@
 La Malva shows especially high summer temperatures within the dry soils (> 40 ºC).
-It must be noted that this difference of > 15 ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939].
+It must be noted that this difference of > 15 ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the most extreme emissions scenario [@RN4939].
 Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.).
@@
 
-E.F.P. received financial support from the Government of Asturias and the FP7 – Marie Curie - COFUND programme of the European Commission (Grant ‘Clarín’ ACB17-19).
+Eduardo Fernández-Pascual received financial support from the Government of Asturias and the FP7 – Marie Curie – COFUND programme of the European Commission (Grants “Clarín” ACA14-19 and ACB17-19).
+Logger installation was partly supported by the LIFE+ programme of the European Commission (project LIFE11/NAT/ES/707 - TREMEDAL).
 

### 2020-10-26 src/analysis.R

@@ -288,7 +288,7 @@ g1 +
 img1 <- png::readPNG("data/Riotuertu.png")
 plot1b <- grid::rasterGrob(img1, width = unit(85, "mm"), height = unit(85, "mm"), just = "centre") 
 
-cowplot::plot_grid(plot1a, plot1b, ncol = 2, labels = c("a", "b")) -> plot1
+cowplot::plot_grid(plot1a, plot1b, ncol = 2, labels = c("(a)", "(b)")) -> plot1
 
 ## Figure 2 - Datalogger records
 
@@ -337,7 +337,7 @@ recordsdf %>%
         legend.margin = margin(0, 0, 0, 0),
         legend.box.margin = margin(0, 0, -7.5, 0),
         plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
+  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3))) +
   scale_color_manual(values = c("red3", "turquoise3")) -> plot2
 
 ## Fig 3 - Effect sizes
@@ -379,7 +379,7 @@ bio %>%
   filter(Variable == "Annual range") %>%
   ggplot(aes(CHELSA, Loggers)) + 
   geom_point(size = 4, aes(color = Groundwater), alpha = 0.65) + 
-  geom_smooth(method = "lm", se = F, linetype = "dashed", aes(color = Groundwater), alpha = 0.65, show.legend = FALSE) +
+  geom_smooth(method = "lm", se = F, linetype = "dashed", aes(color = Groundwater), show.legend = FALSE, alpha = 0.65) +
   geom_abline(intercept = 0, slope = 1, linetype = "dashed") + 
   ggthemes::theme_tufte() + 
   scale_x_continuous(labels = scales::number_format(accuracy = 1)) +
@@ -405,7 +405,7 @@ bio %>%
   filter(Variable == "Diurnal range") %>%
   ggplot(aes(CHELSA, Loggers)) + 
   geom_point(size = 4, aes(color = Groundwater), alpha = 0.65) + 
-  geom_smooth(method = "lm", se = F, linetype = "dashed", aes(color = Groundwater), alpha = 0.65, show.legend = FALSE) +
+  geom_smooth(method = "lm", se = F, linetype = "dashed", aes(color = Groundwater), show.legend = FALSE, alpha = 0.65) +
[…]

### 2020-10-26 README.Rmd
zenodo doi
updated readme

@@
 
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4081418.svg)](https://doi.org/10.5281/zenodo.4081418)
+
 ![Alkaline fen in the Cantabrian Mountains](data/Sierra Cebolleda.jpg)
 
-# Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils
+# Mire microclimate: Groundwater buffers temperature in waterlogged versus dry soils
 
-This repository stores all the information related to the publication *Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils*, including the raw datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the [manuscript](https://github.com/efernandezpascual/mires/blob/master/doc/manuscript.md).
+This repository stores all the information related to the publication [Mire microclimate: Groundwater buffers temperature in waterlogged versus dry soils](https://doi.org/10.1002/joc.6893), including the raw datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the manuscript.
 
@@
 
-Please cite any use of this repository as `Fernández-Pascual, E.
-& Correia-Álvarez., E.
-(2021).
-Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils.
-International Journal of Climatology, in press.`
+Please cite any use of this repository as:
+
+> Fernández-Pascual E & Correia-Álvarez E (2021) [Mire microclimate: Groundwater buffers temperature in waterlogged versus dry soils.](https://doi.org/10.1002/joc.6893) *International Journal of Climatology*.

### 2020-10-26 doc/manuscript.Rmd
updated readme

@@
 ```{r fig1, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 1** **(a)** *Location of the study sites in northern Spain.* **(b)** *Representation of the sampling design at El Riotuertu site.
-The position of the dry and waterlogged temperature loggers is indicated, the blue area represents the extent of the waterlogged mire habitat.*", fig.width=7, fig.height=3.5}
+The position of the dry and waterlogged temperature loggers is indicated, the blue area represents the extent of the waterlogged mire habitat.*"}
 knitr::include_graphics("../results/Fig1.png")
@@
 Dataloggers were buried at 5 cm depth.
-The amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.*", fig.width=7, fig.height=10}
+The amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.*"}
 knitr::include_graphics("../results/Fig2.png")
@@
 ```{r fig3, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 3** *Average bioclimatic variables in the dry and waterlogged points.
-The bars represent the mean value, and the brackets the standard error of 8 records.*", fig.width=7, fig.height=3}
+The bars represent the mean value, and the brackets the standard error of 8 records.*"}
 knitr::include_graphics("../results/Fig3.png")
@@
 ```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 4** *Scatter plots of the bioclimatic variables predicted by CHELSA air temperatures vs.
-soil temperatures measured in situ, in dry and waterlogged points.*", fig.width=7, fig.height=7}
+soil temperatures measured in situ, in dry and waterlogged points.*"}
 knitr::include_graphics("../results/Fig4.png")
@@
 The 1:1 line (corresponding to no buffer effect) is shown, points falling to the right of this line show a buffer effect (i.e.
-soil temperatures at the waterlogged point are cooler than the neighbouring dry site).*", fig.width=3.5, fig.height=3.5}
+soil temperatures at the waterlogged point are cooler than the neighbouring dry site).*"}
 knitr::include_graphics("../results/Fig5.png")

### 2020-10-26 src/analysis.R

@@ -558,10 +558,10 @@ ggsave(plot2, file = "results/Fig2.png",
        path = NULL, scale = 1, width = 170, height = 220, units = "mm", dpi = 300)
 
 ggsave(plot3, file = "results/Fig3.png", 
-       path = NULL, scale = 1, width = 170, height = 85, units = "mm", dpi = 300)
+       path = NULL, scale = 1, width = 170, height = 60, units = "mm", dpi = 300)
 
 ggsave(plot4, file = "results/Fig4.png", 
-       path = NULL, scale = 1, width = 170, height = 170, units = "mm", dpi = 300)
+       path = NULL, scale = 1, width = 140, height = 140, units = "mm", dpi = 300)
 
 ggsave(plot5, file = "results/Fig5.png", 
        path = NULL, scale = 1, width = 85, height = 85, units = "mm", dpi = 300)

### 2021-05-16 README.Rmd
journal issue 🌅

@@
 
-> Fernández-Pascual E & Correia-Álvarez E (2021) [Mire microclimate: Groundwater buffers temperature in waterlogged versus dry soils.](https://doi.org/10.1002/joc.6893) *International Journal of Climatology*.
+> Fernández-Pascual E & Correia-Álvarez E (2021) [Mire microclimate: Groundwater buffers temperature in waterlogged versus dry soils.](https://doi.org/10.1002/joc.6893) *International Journal of Climatology* 41, E2949-E2958.
